---
title: "Appendix: Choosing Better NLDR Layouts by Evaluating the Model in the High-dimensional Data Space"
format: 
    jasa-pdf:
        keep-tex: true
author:
  - name: Jayani P. Gamage
    affiliations:
      - name: Monash University
        department: Econometrics & Business Statistics
        address: Clayton
        city: VIC 
        country: Australia
        postal-code: 3800
    orcid: 0000-0002-6265-6481
    email: jayani.piyadigamage@monash.edu
    url: https://jayanilakshika.netlify.app/
  - name: Dianne Cook
    affiliations:
      - name: Monash University
        department: Econometrics & Business Statistics
        address: Clayton
        city: VIC 
        country: Australia
        postal-code: 3800
    orcid: 0000-0002-3813-7155
    email: dicook@monash.edu 
    url: http://www.dicook.org/
  - name: Paul Harrison
    affiliations:
      - name: Monash University
        department: MGBP, BDInstitute
        address: Clayton
        city: VIC 
        country: Australia
        postal-code: 3800
    orcid: 0000-0002-3980-268X
    email: 	paul.harrison@monash.edu
    url: 
  - name: Michael Lydeamore
    affiliations:
      - name: Monash University
        department: Econometrics & Business Statistics
        address: Clayton
        city: VIC 
        country: Australia
        postal-code: 3800
    orcid: 0000-0001-6515-827X
    email: michael.lydeamore@monash.edu
    url: https://www.michaellydeamore.com/
  - name: Thiyanga S. Talagala
    affiliations:
      - name: University of Sri Jayewardenepura
        department: Statistics
        address: Gangodawila
        city: Nugegoda 
        country: Sri Lanka
        postal-code: 10100
    orcid: 0000-0002-0656-9789
    email: ttalagala@sjp.ac.lk 
    url: https://thiyanga.netlify.app/
tbl-cap-location: bottom
bibliography: paper.bib 
header-includes: | 
  \usepackage{amsmath}
  \usepackage{float}
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \usepackage{bm}
  \def\tightlist{}
  \usepackage{setspace}
  \newcommand\pD{$p\text{-}D$}
  \newcommand\kD{$k\text{-}D$}
  \newcommand\dD{$d\text{-}D$}
  \newcommand\gD{$2\text{-}D$}
  \newcommand\tD{$3\text{-}D$}
  \newcommand\fD{$4\text{-}D$}
  \newcommand\sD{$7\text{-}D$}
  \newcommand{\To}{\textbf{to}~}
  \usepackage{abstract}
  \renewcommand{\abstractname}{}    % clear the title
  \renewcommand{\absnamepos}{empty} % originally center
---

```{r include=FALSE}
# Set up chunk for for knitr
knitr::opts_chunk$set(
  fig.width = 5,
  fig.height = 5,
  fig.align = "center",
  out.width = "100%",
  code.line.numbers = FALSE,
  fig.retina = 4,
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  cache = FALSE,
  dev.args = list(pointsize = 11)
)
```

```{r include=FALSE}
#| label: install-libraries
#| warning: false
#| echo: false

options(repos = c(CRAN = "https://cran.rstudio.com")) # Setup mirror

packages_to_check <- c("remotes", "tidyverse", "patchwork", "colorspace", "kableExtra", "conflicted", "ggforce")

for (pkg in packages_to_check) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    message(paste("Installing package:", pkg))
    install.packages(pkg)
  } else {
    installed_version <- packageVersion(pkg)
    available_version <- tryCatch({
      utils::packageDescription(pkg)$Version
    }, error = function(e) NA) # Handle cases where package info isn't readily available

    if (!is.na(available_version) && installed_version < package_version(available_version)) {
      message(paste("A newer version of package", pkg, "is available. Updating..."))
      install.packages(pkg)
    } else {
      message(paste("Package", pkg, "is up-to-date (version", installed_version, ")."))
    }
  }
}

```

```{r}
#| label: load-libraries
#| warning: false
#| echo: false
library(tibble)
library(dplyr)
library(ggplot2)
library(readr)
library(tidyr)
library(patchwork)
library(colorspace)
library(kableExtra)
library(conflicted)
library(ggforce)
library(GGally)
library(scales)

conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::select)
```

```{r}
#| label: plot-theme
theme_set(theme_linedraw() +
   theme(
     #aspect.ratio = 1,
     plot.background = element_rect(fill = 'transparent', colour = NA),
     plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
     panel.background = element_rect(fill = 'transparent', 
                                     colour = NA),
     panel.grid.major = element_blank(), 
     panel.grid.minor = element_blank(), 
     axis.title.x = element_blank(), axis.title.y = element_blank(),
     axis.text.x = element_blank(), axis.ticks.x = element_blank(),
     axis.text.y = element_blank(), axis.ticks.y = element_blank(),
     legend.background = element_rect(fill = 'transparent', 
                                      colour = NA),
     legend.key = element_rect(fill = 'transparent', 
                               colour = NA),
     legend.position = "bottom", 
     legend.title = element_blank(), 
     legend.text = element_text(size=4),
     legend.key.height = unit(0.25, 'cm'),
     legend.key.width = unit(0.25, 'cm'),
     plot.margin = margin(0, 0, 0, 0)
   )
)
```

```{r}
#| label: import-scripts
source("scripts/additional_functions.R")
```

```{r}
#| label: code-setup
set.seed(20240110)
```

```{r}
clr_choice <- "#0077A3"
```

## Methods and hyper-parameters used to generate layouts

@tbl-fig-param contains the list of methods and hyper-parameters used for each of the layouts shown in the paper. 

```{r}
#| label: tbl-fig-param
#| tbl-pos: H
#| tbl-cap: "NLDR methods and hyper-parameters used for each Figure in the main paper."

fig_param_df <- read_csv("misc/fig_nldr_param.csv")

# Create the table
kable(fig_param_df, 
      format = "latex", 
      booktabs = TRUE, escape = FALSE, longtable = TRUE, linesep = "") |>
  kable_styling(position = "center", 
                full_width = FALSE, 
                font_size = 12,
                latex_options = c("repeat_header")) |>
  row_spec(0, bold = TRUE) |>
  column_spec(1:3, width = c("1.5cm", "2cm", "10cm")) |>
  row_spec(8, hline_after = TRUE) |>
  row_spec(12, hline_after = TRUE) |>
  row_spec(18, hline_after = TRUE) |>
  row_spec(26, hline_after = TRUE) |>
  row_spec(28, hline_after = TRUE) |>
  row_spec(34, hline_after = TRUE) |>
  row_spec(35, hline_after = TRUE) |>
  row_spec(36, hline_after = TRUE) |>
  row_spec(39, hline_after = TRUE) |>
  row_spec(40, hline_after = TRUE) |>
  row_spec(42, hline_after = TRUE) |>
  row_spec(44, hline_after = TRUE) |>
  row_spec(46, hline_after = TRUE) |>
  row_spec(54, hline_after = TRUE)
```

## Videos links

Animations of the \pD{} tours that produced specific projections shown in some figures in the main paper are available on YouTube at the links given in @tbl-links.

```{r}
#| label: tbl-links
#| tbl-pos: H
#| tbl-cap: "Videos of the langevitour animations and the linked plots."
# Video links in the paper

links_df <- read_csv("misc/video_links.csv")

# Create the table
kable(links_df, 
      format = "latex", 
      booktabs = TRUE, escape = FALSE, linesep = "") |>
  kable_styling(position = "center", 
                full_width = FALSE, 
                font_size = 12) |>
  row_spec(0, bold = TRUE) |>
  column_spec(1:2, width = c("1.5cm", "14cm"))
```

## Notation

```{r}
#| label: tbl-notation
#| tbl-pos: H
#| tbl-cap: "Summary of notation for describing new methodology."
# Notation used in the paper

notation_df <- read_csv("misc/notation.csv")

# Create the table
kable(notation_df, 
      format = "latex", 
      booktabs = TRUE, escape = FALSE, linesep = "") |>
  kable_styling(position = "center", 
                full_width = FALSE, 
                font_size = 12) |>
  row_spec(0, bold = TRUE) |>
  column_spec(1:2, width = c("3cm", "12cm"))
```

## Scripts

```{r}
#| label: tbl-script-desc
#| tbl-pos: H
#| tbl-cap: "R and Python script files used to generate outputs in the main paper."

fig_param_df <- read_csv("misc/script_summary.csv")

# Create the table
kable(fig_param_df, 
      format = "latex", 
      booktabs = TRUE, escape = FALSE, longtable = TRUE, linesep = "") |>
  kable_styling(position = "center", 
                full_width = FALSE, 
                font_size = 12,
                latex_options = c("repeat_header")) |>
  row_spec(0, bold = TRUE) |>
  column_spec(1:3, width = c("3.5cm", "6.5cm", "3cm")) |>
  row_spec(3, hline_after = TRUE) |>
  row_spec(11, hline_after = TRUE) |>
  row_spec(15, hline_after = TRUE) |>
  row_spec(18, hline_after = TRUE) |>
  row_spec(36, hline_after = TRUE) 
```

## Generating the 2NC7 data

This data is constructed by simulating two clusters, each consisting of $1000$ observations. The C-shaped cluster is generated from $\theta \sim U(\text{-}3\pi/2, 0)$, $X_1 = \sin(\theta)$, $X_2 \sim U(0, 2)$ (adding thickness to the C), $X_3 = \text{sign}(\theta) \times (\cos(\theta) - 1)$, $X_4 = \cos(\theta)$. Observations lie on a \gD{} manifold in \sD{}. The other cluster is from $X_1 \sim U(0, 2)$, $X_2 \sim U(0, 3)$, $X_3 = \text{-}(X_1^3 + X_2)$, and $X_4 \sim U(0, 2)$. It is also curved, but observations lie on a \tD{} manifold in \sD{}. Three more variables, $X_5, X_6, X_7$, that are small amounts of pure noise are added. We would consider $T=(X_1, X_2, X_3, X_4)$ to be the geometric structure (true model) that we hope to capture.

```{r}
proj_obj1_true <- read_rds("data/two_nonlinear/two_nonlinear_true_proj_obj1.rds")

two_curvy_proj_true1 <- plot_proj(
  proj_obj = proj_obj1_true,
  point_param = c(0.5, 0.2, clr_choice), # size, alpha, color
  line_param = c(0.5, 0.3, "#000000"), # linewidth, alpha
  plot_limits = c(-0.75, 0.75),
  axis_text_size = 4,
  is_category = FALSE) +
  interior_annotation(label = "a", cex = 2) +
  theme(aspect.ratio = 1,
        legend.position = "none")

proj_obj2_true <- read_rds("data/two_nonlinear/two_nonlinear_true_proj_obj2.rds")

two_curvy_proj_true2 <- plot_proj(
  proj_obj = proj_obj2_true,
  point_param = c(0.1, 0.5, clr_choice), # size, alpha, color
  line_param = c(0.5, 0.3, "black"), # linewidth, alpha
  plot_limits = c(-0.5, 0.5),
  axis_text_size = 4,
  is_category = FALSE) +
  interior_annotation(label = "b", cex = 2) +
  theme(aspect.ratio = 1)
```

```{r}
#| fig-width: 10
#| fig-height: 5
#| fig-cap: "Two projections of the \\pD{} true model overlaying the data are shown in a, b. Video of the langevitour animations is available at <https://youtu.be/35TrnYJsUUI>."

two_curvy_proj_true1 + two_curvy_proj_true2 +
  plot_layout(ncol = 2)
```

## Computing hexagon grid configurations

Given range of embedding component, $r_2$, number of bins along the x-axis, $b_1$, and buffer proportion, $q$, hexagonal starting point coordinates, $s_1 = \text{-}q$, and $s_2 = \text{-}qr_2$. The purpose is to find width of the hexagon, $a_1$ and number of bins along the y-axis, $b_2$.

Geometric arguments give rise to the following constraints.

$\text{min }a_1 \text{ s.t.}$

$$
s_1 - \frac{a_1}{2} < 0,
$$ {#eq-equation1}

$$
s_1 + (b_1 - 1) \times a_1 \geq 1,
$$ {#eq-equation2}

$$
s_2 - \frac{a_2}{2} < 0,
$$ {#eq-equation4}

$$
s_2 + (b_2 - 1) \times a_2 \geq r_2.
$$ {#eq-equation5}




```{r}
#| label: code-illustration
# Code to draw illustration for notation
## hexagon binning to have regular hexagons

hex_grid_temp <- read_rds("data/two_nonlinear/hex_grid_temp.rds")
all_centroids_df_temp <- read_rds("data/two_nonlinear/all_centroids_df_tempp.rds")
start_pt <- read_rds("data/two_nonlinear/start_pt.rds")
d_rect <- read_rds("data/two_nonlinear/d_rect.rds")
a1 <- read_rds("data/two_nonlinear/a2_data.rds")
a2 <- read_rds("data/two_nonlinear/a1_data.rds")
l <- read_rds("data/two_nonlinear/l_data.rds")
rect_adj <- read_rds("data/two_nonlinear/rect_adj.rds")

hex_param_vis <- ggplot() + 
  geom_polygon(data = hex_grid_temp, 
               aes(x = x, 
                   y = y, 
                   group = h),
               fill = "white", 
               colour = "#bdbdbd") +
  geom_point(data = all_centroids_df_temp, aes(
    x = c_x, 
    y = c_y), 
    colour = "#31a354", size = 0.9) +
  geom_point(data = start_pt, aes(x = c_x, 
                                  y = c_y), 
             colour = "black") + 
  geom_rect(data=d_rect, 
            aes(xmin = x1min - rect_adj$x1,# - rect_adj$s1, 
                xmax = x1max - rect_adj$x1,# - rect_adj$s1, 
                ymin = x2min - rect_adj$x2,# - rect_adj$s2, 
                ymax = x2max - rect_adj$x2),# - rect_adj$s2), 
            fill = "white", 
            colour = "black", 
            alpha = 0, 
            linewidth = 0.7) +
  geom_point(data=d_rect, aes(x=x1min - rect_adj$x1, 
                              y=x2min - rect_adj$x2)) + 
  geom_point(data=d_rect, aes(x=x1max - rect_adj$x1, 
                              y=x2min - rect_adj$x2)) + 
  geom_point(data=d_rect, aes(x=x1min - rect_adj$x1, 
                              y=x2max - rect_adj$x2)) + 
  annotate("text", x=d_rect$x1min - rect_adj$x1, 
           y=d_rect$x2min - rect_adj$x2,
           label = "(0,0)", 
           hjust=-0.1, vjust=-0.3, size = 8) + 
  annotate("text", x=d_rect$x1max - rect_adj$x1, 
           y=d_rect$x2min - rect_adj$x2,
           label = "(0,1)", 
           hjust=1.1, vjust=-0.3, size = 8) + 
  annotate("text", x=d_rect$x1min - rect_adj$x1, 
           y=d_rect$x2max - rect_adj$x2,
           label = "'(0,'*y[2][max]*')'", 
           parse = TRUE,
           hjust=-0.1, vjust=1.2, size = 8) + 
  geom_segment(data=d_rect, aes(
    x = x1min  - rect_adj$x1, # 0 - 0.03, 
    y = -0.31, 
    xend = x1max - rect_adj$x1, #1 - 0.03, 
    yend = -0.31), #-0.35),
    arrow = arrow(length = unit(0.03, "npc"),
                  ends = "both"), 
    colour = "black")+
  annotate("text", x=0.5, y=-0.38, 
           label = "r[1]", parse = TRUE, colour = "black", size = 8) +
  geom_segment(data=d_rect, aes(
    x = -0.25, 
    y = x2min - rect_adj$x2, #0 - 0.05, 
    xend = -0.25, 
    yend = x2max - rect_adj$x2), #r2 - 0.05),
    arrow = arrow(length = unit(0.03, "npc"),
                  ends = "both"), 
    colour = "black")+ 
  annotate("text", x=-0.3, y=0.4, 
           label = "r[2]", parse = TRUE, colour = "black", size = 8) +
  geom_segment(data = a1, aes(
    x = x, #-0.1 + 0.2087578, 
    y = y, #-0.15, 
    xend = xend, #-0.1 + 0.2087578*2, 
    yend = yend), #-0.15),
    arrow = arrow(length = unit(0.03, "npc"),
                  ends = "both"), 
    colour = "black")+ # a1 = 0.2087578
  annotate("text", 
           x=(a1$x+a1$xend)/2, 
           y=a1$y, 
           label = "a[1]", parse = TRUE, 
           colour = "black",
           vjust = 1.2, size = 8) +
  geom_segment(data = a2, aes(
    x = x, #-0.15, 
    y = y, #-0.1*r2 + 0.1807896*2, 
    xend = xend, #-0.15, 
    yend = yend), #-0.1*r2 + 0.1807896*3),
    arrow = arrow(length = unit(0.03, "npc"),
                  ends = "both"), 
    colour = "black") + # a2 = 0.1807896
  annotate("text", x=a2$x, y=(a2$y+a2$yend)/2, 
           label = "a[2]", parse = TRUE,
           colour = "black", hjust=-0.2, size = 8) +
  annotate("text", x=-0.18, y=-0.24, 
           label = "'('*s[1]*','*s[2]*')'", parse = TRUE,
           colour = "black", size = 8) +
  geom_segment(data = l, aes(
    x = x, #-0.15, 
    y = y, #-0.1*r2 + 0.1807896*2, 
    xend = xend, #-0.15, 
    yend = yend), #-0.1*r2 + 0.1807896*3),
    arrow = arrow(length = unit(0.03, "npc"),
                  ends = "both"), 
    colour = "black") + 
  annotate("text", x=l$x + 0.03, y=(l$y+l$yend)/2, 
           label = "l", parse = TRUE,
           colour = "black", hjust=-0.2, size = 8) +
  coord_equal()
```

```{r}
#| label: fig-hex-param
#| fig-cap: "The components of the hexagon grid illustrating notation."
#| out-width: 30%
#| fig-pos: "!ht"
#| eval: false

hex_param_vis
```

Since $a_1$ and $a_2$ are distances,

$$
a_1, a_2 > 0.
$$ Also, $(s_1, s_2) \in (\text{-}0.1, \text{-}0.05)$ as these are multiplicative offsets in the negative direction.

@eq-equation1 can be rearranged as,

$$
a_1 > 2s_1
$$

which given $s_1 < 0$ and $a_1 > 0$ will *always* be true. The same logic follows for @eq-equation4 and substituting $a_2 = \sqrt{3}a_1/{2}$, and $s_2 = \text{-}qr_2$ to @eq-equation4 can be written as,

$$
a_1 > -\frac{4}{\sqrt{3}}qr_2
$$

Also, substituting $a_2 = \sqrt{3}a_1/{2}$, $s_2 = \text{-}qr_2$ and rearranging @eq-equation5 gives:

$$
a_1 \geq \frac{2(r_2 + qr_2)}{\sqrt{3}(b_2 - 1)}.
$$ {#eq-equation6}

Similarly, substituting $s_1 = \text{-}q$ @eq-equation2 becomes,

$$
a_1 \geq \frac{(1 + q)}{(b_1 - 1)}.
$$ {#eq-equation7}

This is a linear optimization problem. Therefore, the optimal solution must occur on a vertex. So, by setting @eq-equation6 equals to @eq-equation7 gives,

$$
\frac{2(r_2 + qr_2)}{\sqrt{3}(b_2 - 1)} = \frac{(1 + q)}{(b_1 - 1)}.
$$ 

After rearranging this,

$$
b_2 = 1 + \frac{2r_2(b_1 - 1)}{\sqrt{3}}
$$

and since $b_2$ should be an integer,

$$
b_2 = \Big\lceil1 +\frac{2r_2(b_1 - 1)}{\sqrt{3}}\Big\rceil.
$$ {#eq-equation8}

Furthermore, with known $b_1$ and $b_2$, by considering @eq-equation2 or @eq-equation5 as the *binding* or *active constraint*, can compute $a_1$.

If @eq-equation2 is active, then,

$$
\frac{(1 + q)}{(b_1 - 1)} < \frac{2(r_2 + qr_2)}{\sqrt{3}(b_2 - 1)}.
$$

Rearranging this gives,

$$
r_2 > \frac{\sqrt{3}(b_2 - 1)}{2(b_1 - 1)}.
$$

Therefore, if this equality is true, then 
$$
a_1 = \frac{(1+q)}{(b_1 - 1)},
$$
otherwise, 
$$
a_1 = \frac{2r_2(1+q)}{\sqrt{3}(b_2 - 1)}.
$$

## Binning the data

Points are assigned to the bin they fall into based on the nearest centroid. If a point is equidistant from multiple centroids, it is assigned to the centroid with the smallest bin ID.

```{r}

## Data set with all centroids
all_centroids_df <- read_rds("data/two_nonlinear/two_nonlinear_all_centroids_two_curvy4.rds")

## Generate all coordinates of hexagons
hex_grid <- read_rds("data/two_nonlinear/two_nonlinear_hex_grid_two_curvy4.rds")

## Data to generate circle
circles <- data.frame(
  x = c(0.608, 0.372),
  y = c(0.306, 0.715),
  r = rep(0.27, 2)
)

emb_df1 <- tibble(emb1 = 0.45, emb2 = 0.47) |>
  add_row(emb1 = 0.54, emb2 = 0.56)

assign_data_vis1 <- ggplot(data = hex_grid, aes(x = x, y = y)) + 
  geom_polygon(fill = "white", color = "#bdbdbd", aes(group = h)) +
  geom_point(data = emb_df1, aes(x = emb1, y = emb2), color = "black", size = 0.7) +
  geom_text(data = all_centroids_df, aes(x = c_x, y = c_y, label = h), size = 5, colour = "#bdbdbd") +
  geom_circle(data = circles, aes(x0 = x, y0 = y, r = r), fill = NA) +
  geom_segment(x = 0.45, y = 0.47, xend = 0.58, yend = 0.3, 
               arrow = arrow(length = unit(0.02, "npc")), color = "#969696") +
  geom_segment(x = 0.54, y = 0.56, xend = 0.4, yend = 0.715, 
               arrow = arrow(length = unit(0.02, "npc")), color = "#969696")
```

```{r}
hex_data <- hex_grid |> 
  dplyr::filter(h == 8)

# Create all pairs of rows
pairs <- expand_grid(row1 = 1:nrow(hex_data), row2 = 1:nrow(hex_data)) |>
  filter(row1 < row2)

# Compute middle points for each pair of rows
middle_points <- pairs |>
  rowwise() |>
  mutate(
    h = hex_data$h[row1],
    x = (hex_data$x[row1] + hex_data$x[row2]) / 2,
    y = (hex_data$y[row1] + hex_data$y[row2]) / 2
  ) |>
  ungroup() |>
  select(h, x, y) |>
  filter(row_number() %in% c(1, 5, 6, 10, 13, 15))

points_df <- bind_rows(hex_data, middle_points)

arrow_df <- tibble(
  x = points_df$x, 
  y = points_df$y,
  xend = c(0.37, points_df$x[2] - 0.1, 0.136, points_df$x[4] - 0.1, 0.61, points_df$x[6] - 0.1, 0.33, points_df$x[8] - 0.1, 0,  points_df$x[10] - 0.1, 0.58, 0.45),
  yend = c(0.82, points_df$y[2] - 0.1, 0.4, points_df$y[4] - 0.1, 0.4, points_df$y[6] - 0.1, 0.82, points_df$y[8] - 0.1, 0.715, points_df$y[10] - 0.1, 0.4, 0.71)
)

assign_data_vis2 <- ggplot(data = hex_grid, aes(x = x, y = y)) + 
  geom_polygon(fill = "white", color = "#bdbdbd", aes(group = h)) +
  geom_point(data = points_df, aes(x = x, y = y), color = "black", size = 0.7) +
  geom_text(data = all_centroids_df, aes(x = c_x, y = c_y, label = h), size = 5, colour = "#bdbdbd") +
  geom_segment(data = arrow_df, aes(x = x, y = y, xend = xend, yend = yend), 
               arrow = arrow(length = unit(0.02, "npc")), color = "#bdbdbd") 
```

```{r}
#| label: fig-assign-data
#| fig-cap: "Binning the data. Points are assigned to the nearest centroid. If a point is equidistant from multiple centroids, assigned to the centroid with the smallest bin ID."
#| fig-width: 12
#| fig-height: 6
#| fig-pos: "!ht"

assign_data_vis1 + assign_data_vis2 +
  plot_layout(ncol = 2)
```

## Area of a hexagon

The area of a hexagon is defined as $A = 3\sqrt{3}l^2/2$, where $l$ is the side length of the hexagon. $l$ can be computed using $a_1$ and $a_2$.


```{r}
# triangle_temp <- tibble(
#   x = c(0, 4, 0),
#   y = c(0, 0, 3)
# )

triangle_line_temp <- tibble(
  x = c(4, 0, 4),
  xend = c(0, 0, 0),
  y = c(0, 0, 0),
  yend = c(0, 3, 3)
)

# triangle_arrow_temp <- tibble(
#   x = c(4 + 0.04, 0 - 0.06, 4 + 0.06),
#   xend = c(0 + 0.04, 0 - 0.06, 0 + 0.06),
#   y = c(0 - 0.04, 0 - 0.06, 0 + 0.06),
#   yend = c(0 - 0.04, 3 - 0.06, 3 + 0.06)
# )

triangle_plot <- ggplot() +
  # geom_point(
  #   data = triangle_temp,
  #   aes(
  #     x = x,
  #     y = y
  #   )
  # ) +
  geom_segment(
    data = triangle_line_temp,
    aes(
      x = x,
      xend = xend,
      y = y,
      yend = yend
    )
  ) + 
  annotate("text", x = -1.5, y = 1.5,
         label = "frac(a[2]-a[1], 2)",
         parse = TRUE,
         color = "black", hjust = -0.2, size = 8) +
annotate("text", x = 1.5, y = -0.5,
         label = "a[1]/2",
         parse = TRUE,
         color = "black", hjust = -0.2, size = 8) +
annotate("text", x = 2.5, y = 1.5,
         label = "l",
         parse = TRUE,
         color = "black", hjust = -0.2, size = 8)
```

```{r}
#| label: fig-tri-param
#| fig-cap: "The components of the right triangle illustrating notation."
#| out-width: 30%
#| fig-pos: "!ht"

triangle_plot
```

By applying the Pythagorean theorem, we obtain,

$$
l^2 = \left(\frac{a_1}{2}\right)^2 + \left(\frac{a_2 - l}{2}\right)^2.
$$
Next, rearranging the terms, we get,

$$
l^2 - \left(\frac{a_2 - l}{2}\right)^2 = \left(\frac{a_1}{2}\right)^2,
$$

$$
\left[l - \left(\frac{a_2 - l}{2}\right)\right]\left[l + \left(\frac{a_2 - l}{2}\right)\right] = \left(\frac{a_1}{2}\right)^2,
$$

$$
3l^2 + 2a_2l - (a_1^2 + a_2^2) = 0.
$$

Finally, by solving the quadratic equation, we compute,

$$
l = \frac{-2a_2 \pm \sqrt{4a_2^2 - 24[-(a_1^2 + a_2^2)]}}{6},
$$

$$
l = \frac{-a_2 \pm \sqrt{a_2^2 - 6[-(a_1^2 + a_2^2)]}}{3},
$$

where $l > 0$.

## Tuning {#sec-tuning}

<!-- The model fitting is based on several parameters, including the hexagon bin parameters and the low-count bin removal process. The hexagon bin parameters define the bottom-left bin position $(s_1, \ s_2)$, the number of bins in the horizontal direction ($b_1$), which also determines the number of bins in the vertical direction ($b_2$), the total number of bins ($b$), and the total number of non-empty bins ($m$). Low count bins are removed using standardized bin counts, defined as $w_h = n_h/n, ~~h=1, \dots m$. -->

<!-- Default values are provided for each of these, but deciding on the best model fit is assisted by examining a range of values. The default number of bins $b=b_1\times b_2$ is computed based on the sample size, by setting $b_1=n^{1/3}$, consistent with the Diaconis-Freedman rule [@freedman1981]. The value of $b_2$ is determined analytically by $b_1, q, r_2$ (@eq-bin2). Values of $b_1$ between $2$ and $b_1 = \sqrt{n/r_2}$ are recommended, where the dependence on $r_2$ reflects the preservation of aspect ratio in the NLDR layout. -->

<!-- @fig-bins-two-curvy shows the hexbin grids for three choices of $b_1$. While the number of bins is the common parameter to modify, bin start positions $(s_1, \ s_2)$ can also be worth experimenting with, because they can also change bin counts.  -->


The model fitting depends on several parameters, including the hexagon bin parameters and the low-count bin removal process. The hexagon bin parameters define the bottom-left bin position $(s_1, s_2)$, the number of bins in the horizontal direction ($b_1$), which determines the number of bins in the vertical direction ($b_2$), the total number of bins ($b$), and the number of non-empty bins ($m$). Low-count bins are removed using standardized bin counts, defined as $w_h = n_h/n$, for $h = 1, \dots, m$.

Default values are provided for these parameters. In particular, the default number of bins is determined from the sample size by setting $b_1 = n^{1/3}$, consistent with the Diaconis–Freedman rule [@freedman1981], with $b_2$ computed to preserve the aspect ratio of the layout.

Rather than selecting a single set of parameters a priori, we recommend a structured procedure to examine the stability of the fitted model across a range of values. A practical workflow is as follows:

1. **Select a grid of binwidths:** Choose a sequence of values for $b_1$ ranging from coarse to fine resolution (e.g., from small values up to $b_1 = \sqrt{n/r_2}$). This controls the level of detail in the fitted wireframe.

2. **Compute HBE across the grid:** For each choice of $b_1$, compute the HBE. Stable regions where HBE changes slowly indicate parameter values where the fitted model is not overly sensitive to binning resolution.

3. **Assess bin occupancy:** Examine summary quantities such as average bin count ($\bar{n}$), average standardized bin count ($\bar{w}$), and the proportion of non-empty bins ($m/b$).
   Reasonable parameter choices typically correspond to:

   - sufficiently large $\bar{n}$ to avoid sparse bins, and
   
   - moderate values of $m/b$, indicating that bins cover the data without excessive fragmentation.

4. **Check robustness to bin removal:** Evaluate the effect of removing low-count bins using a range of thresholds. Parameter values where HBE remains stable under small amounts of bin removal indicate robustness of the fitted structure.

5. **Compare across layouts (if applicable):** When comparing multiple NLDR layouts, examine HBE as a function of average bin count or standardized bin count. This helps account for differences in point density between layouts.

This process does not yield a single "optimal" parameter choice, but instead identifies a range of stable and interpretable settings. The goal is to avoid parameter values that produce unstable fits, overly sparse binning, or misleading structure in the fitted wireframe.

```{r}
sc_ltr_pos <- c(0.08, 0.9)
sc_xlims <- c(-0.3, 1.2)
sc_ylims <- c(-0.17, 1.25)
```

```{r}
#| label: hexbin-regular-two-curvy1
hex_grid_with_counts <- read_rds("data/two_nonlinear/hex_grid_with_counts.rds")

tsne_two_curvy_scaled <- read_rds("data/two_nonlinear/tsne_two_curvy_scaled.rds")

hex_grid_coloured_two_curvy1 <- ggplot() + 
  geom_polygon(
    data = hex_grid_with_counts, 
    aes(x = x, y = y,
      group = h, 
      fill = w_h), colour = "grey70", linewidth=0.2) +
  geom_point(data = tsne_two_curvy_scaled,
             aes(x = emb1, y = emb2),
             colour = clr_choice,
             alpha = 0.3,
             size = 0.5) +
  scale_fill_viridis_c(direction = -1, 
    na.value = "#ffffff", option = "E") +
  interior_annotation("b", sc_ltr_pos, cex = 2) + 
  xlim(sc_xlims) + ylim(sc_ylims)
```

```{r}
#| label: hexbin-regular-two-curvy2

hex_grid_with_counts_two_curvy2 <- read_rds("data/two_nonlinear/two_nonlinear_hex_grid_with_counts_two_curvy2.rds")
tr_from_to_df_two_curvy2 <- read_rds("data/two_nonlinear/two_nonlinear_tr_from_to_df_two_curvy2.rds")

hex_grid_coloured_two_curvy2 <- ggplot() + 
  geom_polygon(
    data = hex_grid_with_counts_two_curvy2, 
    aes(x = x, y = y, 
        group = h, 
        fill = w_h), 
        colour = "grey70", 
        linewidth=0.2) +
  geom_point(data = tsne_two_curvy_scaled,
           aes(x = emb1, y = emb2),
           alpha = 0.3,
           size = 0.5) +
  scale_fill_viridis_c(direction = -1, 
    na.value = "#ffffff", option = "C") +
  xlim(sc_xlims) + ylim(sc_ylims) +
  interior_annotation("a", sc_ltr_pos, cex = 2)

trimesh_two_curvy2 <- ggplot() +
  geom_segment(data = tr_from_to_df_two_curvy2,
               aes(
                 x = x_from,
                 y = y_from,
                 xend = x_to,
                 yend = y_to),
               colour = "#000000") +
  geom_point(data = tsne_two_curvy_scaled,
             aes(
               x = tSNE1,
               y = tSNE2
             ),
             colour = "#636363",
            alpha = 0.5,
            size = 0.5
            ) +
  coord_equal() +
  interior_annotation("a1", sc_ltr_pos) 

```

```{r}
#| label: hexbin-regular-two-curvy3

hex_grid_with_counts_two_curvy3 <- read_rds("data/two_nonlinear/two_nonlinear_hex_grid_with_counts_two_curvy3.rds")
tr_from_to_df_two_curvy3 <- read_rds("data/two_nonlinear/two_nonlinear_tr_from_to_df_two_curvy3.rds")

hex_grid_coloured_two_curvy3 <-  ggplot() + 
  geom_polygon(
    data = hex_grid_with_counts_two_curvy3, 
    aes(x = x, y = y, 
        group = h, 
        fill = w_h), 
        colour = "grey70", linewidth=0.2) +
  geom_point(data = tsne_two_curvy_scaled,
           aes(x = emb1, y = emb2),
           alpha = 0.3,
           size = 0.5) +
  scale_fill_viridis_c(direction = -1, na.value = "#ffffff", option = "D") +
  xlim(sc_xlims) + ylim(sc_ylims) +
  interior_annotation("c", sc_ltr_pos, cex = 2)
```

```{r}
#| echo: false
#| label: fig-bins-two-curvy
#| fig-cap: "Hexbin density plots of tSNE layout of the 2NC7 data, using three different $b_1$ specifications yielding different $b_2, b, m$: (a) **15**, 18, 270, 98, (b) **24**, 29, 696, 209, and (c) **35**, 42, 1470, 386. Color indicates standardized counts, dark indicating high count, and light indicating low count. At the smallest binwidth, the data structure is discontinuous, suggesting that there are too many bins."
#| fig-width: 9
#| fig-height: 3

hex_grid_coloured_two_curvy2 +
  hex_grid_coloured_two_curvy1 +
  hex_grid_coloured_two_curvy3 +
  plot_layout(guides='collect', ncol = 3) &
  theme(legend.position='none', plot.tag = element_text(size = 12))
``` 

<!-- #### Handling of low-density bins-->

<!-- It is worthwhile to consider what desirable aspects of a hexbin result, that maps to summarizing the \pD{} fit well. The binning should capture the underlying data distribution closely, with a minimum number of necessary bins. An ideal binning might be indicated by a more uniform distribution of bin counts or having few relatively empty bins. To help with this assessment average bin count ($\bar{n} = \sum{n_h}/m$), average standardized bin count ($\bar{w} = \sum{w_h}/m$) and proportion of non-empty bins ($m/b$), are also computed. @fig-param-two-curvy shows some choices of plots of these quantities for a single NLDR layout, with three choices of $a_1$ indicated. Some expectations and reasoning for these plots are: -->

<!-- - HBE will increase as $a_1$ increases, so good choices will be just before a big increase. In plot a, HBE changes fairly steadily, so there is no easy choice to make. -->
<!-- - HBE can also be examined against the average standardized bin count or average bin count (plot b). This is similar to the comparison with $a_1$, but to use when comparing different NLDR layouts. Different layouts might produce different densities of points, which will not be captured well by a comparison of HBE vs $a_1$. -->
<!-- - The proportion of non-empty bins is interesting to examine across different binwidths (plot c). A good binning should have just the right number of bins to neatly cover the shape of the data, and no more or fewer. As binwidth gets smaller, $m/b$ should roughly get bigger. -->
<!-- - Bins with a small number of observations might be removed to sharpen the wireframe model. This can have adverse effects, though - failing to extend the wireframe into sparse areas, or resulting in holes in the wireframe. Plot d shows the relationship between HBE (computed for all observations despite some bin removal) and the standardized bin count cutoff used to remove bins. For all three chosen bin widths, a small number of bins can be removed without affecting HBE. -->

<!--two_nonlinear/05_gen_rm_lwd_mse.R-->
```{r}
#| label: rm-lwd-bin-error

error_rm_two_curvy <- read_rds("data/two_nonlinear/error_rm_lwd_diff_bin.rds") |>
  mutate(b1  = as.factor(b1))

rmse_two_curvy_lwd <- ggplot(error_rm_two_curvy, 
                     aes(x = benchmark_rm_lwd, 
                         y = HBE,
                         colour = a1,
                         linetype = a1)) + 
  geom_point(
    size = 1
    ) +
  geom_line(
    linewidth = 0.5
    ) + 
  scale_x_continuous(breaks = sort(unique(error_rm_two_curvy$benchmark_rm_lwd))[c(1, 4, 7, 11, 15)]) +
  scale_color_manual(values=c("#d95f02", "#1f78b4", "#1b9e77"),
                     labels = parse_format()) +
  scale_linetype_manual(values = c("longdash", "solid", "dotted"),
                        labels = parse_format()) +
  xlab(expression(paste("low count bin cutoff based on ", w[h]))) +
  ylab("HBE") +
  #xlim(c(0, 0.14)) +
  ylim(c(0, 0.75)) +
  interior_annotation("d", position = c(0.07, 0.9), cex = 2) +
  theme_minimal() +
  theme(aspect.ratio = 0.75,
        panel.border = element_rect(fill = 'transparent'),
        plot.title = element_text(size = 20, 
                                  hjust = 0.5, 
                                  vjust = -0.5),
        axis.ticks.x = element_line(),
        axis.ticks.y = element_line(),
        legend.position = "none",
        legend.title = element_blank(),
        legend.text = element_text(size = 18),
        axis.title = element_text(size = 18),
        axis.text = element_text(size = 15))
```

```{r}
#| label: error-two-curvy

error_two_curvy_tsne <- read_rds("data/two_nonlinear/error_two_curvy_tsne.rds")

rmse_two_curvy_b <- ggplot(error_two_curvy_tsne, 
                     aes(x = a1, 
                         y = HBE)) +
  geom_vline(xintercept = 0.03,
           colour = "#d95f02", linetype="dashed", 
           linewidth=1) +
  geom_vline(xintercept = 0.05,
             colour = "#1f78b4", 
             linetype="solid",
             linewidth=1) +
  geom_vline(xintercept = 0.08, linetype="dotted",
             colour = "#1b9e77", linewidth=1) +
  geom_line(linewidth = 0.5) + 
  geom_point(size = 1) +
  scale_x_continuous(breaks = seq(0.03, 0.28, 0.03)) + #sort(unique(round(error_two_curvy_tsne$a1, 2)))[c(1, 4, 7, 11, 13, 14, 15, 16)]) +
  labs(x = expression(paste("binwidth (", a[1], ")")), y = "HBE") +
  ylim(c(0, 0.75)) +
  interior_annotation("a", position = c(0.9, 0.1), cex = 2) +
  theme_minimal() +
  theme(aspect.ratio = 0.75,
        panel.border = element_rect(fill = 'transparent'),
        plot.title = element_text(size = 20, hjust = 0.5, vjust = -0.5),
        axis.ticks.x = element_line(),
        axis.ticks.y = element_line(),
        axis.title = element_text(size = 18),
        axis.text = element_text(size = 15))

error_two_curvy_tsne_filtered <- error_two_curvy_tsne |>
  filter(b1 %in% c(15, 24, 35))

rmse_two_curvy_relcount <- ggplot(error_two_curvy_tsne, 
                     aes(x = n_bar, 
                         y = HBE)) +
  geom_vline(xintercept = error_two_curvy_tsne_filtered$n_bar[2],
           colour = "#d95f02", linetype="dashed", 
           linewidth=1) +
  geom_vline(xintercept = error_two_curvy_tsne_filtered$n_bar[3],
             colour = "#1f78b4", 
             linetype="solid",
             linewidth=1) +
  geom_vline(xintercept = error_two_curvy_tsne_filtered$n_bar[1], linetype="dotted",
             colour = "#1b9e77", linewidth=1) +
  geom_line(linewidth = 0.5) + 
  geom_point(size = 1) +
#  scale_x_continuous(breaks = sort(unique(round(error_two_curvy_tsne$a1, 2)))[c(1, 4, 7, 11, 13, 14, 15, 16)]) +
  scale_x_log10() +
  ylim(c(0, 0.75)) +
  ylab("HBE") +
  xlab(expression(paste("average points per bin (", bar(n)[h], ")(log scale))"))) +
  #ylim(c(0, 0.75)) +
  annotate("text", x = 109, y = 0.05, label = "b", size = 7, colour = "grey70") +
  theme_minimal() +
  theme(aspect.ratio = 0.75,
        panel.border = element_rect(fill = 'transparent'),
        plot.title = element_text(size = 20, hjust = 0.5, vjust = -0.5),
        axis.ticks.x = element_line(),
        axis.ticks.y = element_line(),
        axis.title = element_text(size = 18),
        axis.text = element_text(size = 15)) 

a1_m_two_curvy <- ggplot(error_two_curvy_tsne,
                     aes(x = a1,
                         y = prop_bins)) +
  geom_vline(xintercept = 0.03,
           colour = "#d95f02", linetype="dashed", 
           linewidth=1) +
  geom_vline(xintercept = 0.05,
             colour = "#1f78b4", 
             linetype="solid",
             linewidth=1) +
  geom_vline(xintercept = 0.08, linetype="dotted",
             colour = "#1b9e77", linewidth=1) +
  geom_line(linewidth = 0.5) +
  geom_point(size = 1) +
  # scale_y_continuous(breaks = sort(unique(error_two_curvy$b_non_empty))[-3]) +
  scale_x_continuous(breaks = seq(0.03, 0.28, 0.03)) +
  labs(x = expression(paste("binwidth (", a[1], ")")), y = expression("prop. non-empty bins " * bgroup("(", m / b, ")"))) +
  interior_annotation("c", position = c(0.9, 0.1), cex = 2) +
  theme_minimal() +
  theme(aspect.ratio = 0.75,
        panel.border = element_rect(fill = 'transparent'),
        plot.title = element_text(size = 20, hjust = 0.5, vjust = -0.5),
        axis.ticks.x = element_line(),
        axis.ticks.y = element_line(),
        axis.title = element_text(size = 18),
axis.text = element_text(size = 15))
```


```{r}
#| echo: false
#| fig-cap: "Various plots to help tune the model fit: (a) HBE vs $a_1$, (b) HBE vs $\\bar{n}_h$, (c) proportion of non-empty bins ($m/b$) vs $a_1$, (d) HBE vs $w_h$ cutoff used for removing low count bins. Color indicates the three binwidths show in Figure 6: $0.03$ (orange dashed), $0.05$ (blue solid), and $0.08$ (green dotted). The better model fit will have low HBE, but reasonably sized bins that capture the data sufficiently. Proportion of non-empty bins tends to increase with $a_1$ (c). Removing a few low-count bins doesn't substantially change HBE, for all three binwidths (d)."
#| label: fig-param-two-curvy
#| out-width: 100%
#| fig-width: 12
#| fig-height: 10

rmse_two_curvy_b + rmse_two_curvy_relcount + a1_m_two_curvy +
  rmse_two_curvy_lwd + 
  plot_layout(ncol=2)
```

## Curiosities about NLDR results discovered by examining the model in the data space {#sec-curiosities}

With the drawing of the model in the data, several interesting differences between NLDR methods can be observed.

### Some methods appear to order points in the layout

The \gD{} model representations generated from some NLDR methods, especially PaCMAP, are unreasonably flat or like a pancake. A simple example of this can be seen with data simulated to contain five \fD{} Gaussian clusters. Each cluster is essentially a ball in \fD{}, so there is no \gD{} representation, rather the model in each cluster should resemble a crumpled sheet of paper that fills out \fD{}.

@fig-five-gau-projs a1, b1, c1 show the \gD{} layouts for (a) tSNE, (b) UMAP, and (c) PaCMAP, respectively. The default hyper-parameters for each method are used. In each layout we can see an accurate representation where all five clusters are visible, although with varying degrees of separation.

The models are fitted to each these layouts. @fig-five-gau-projs a2, b2, c2 show the fitted models in a projection of the \fD{} space, taken from a tour. These clusters are fully \fD{} in nature, so we would expect the model to be a *crumpled sheet* that stretches in all four dimensions. This is what is mostly observed for tSNE and UMAP. The curious detail is that the model for PaCMAP is closer to a *pancake* in shape in every cluster! This single projection only shows this in three of the five clusters but if we examine a different projection the other clusters exhibit the pancake also. While we don't know what exactly causes this, it is likely due to some ordering of points in the \gD{} PaCMAP layout that induces the flat model. One could imagine that if the method used principal components on all the data, that it might induce some ordering that would produce the flat model. If this were the reason, the pancaking would be the same in all clusters, but it is not: The pancake is visible in some clusters in some projections but in other clusters it is visible in different projections. It might be due to some ordering by nearest neighbors in a cluster. The PaCMAP documentation doesn't provide any helpful clues. That this happens, though, makes the PaCMAP layout inadequate for representing the high-dimensional data. 

<!--Projections-->
<!--five_gau_clusters/02_gen_model_with_tSNE.R-->
```{r}
#| label: five-gau-proj-tsne-model

data_with_clust <- read_rds("data/five_gau_clusters/data_five_gau_with_clusts.rds")
tsne_gau_scaled <- read_rds("data/five_gau_clusters/tsne_gau_scaled.rds")
tr_from_to_df_gau1 <- read_rds("data/five_gau_clusters/tr_from_to_df_gau1.rds")

trimesh_removed_gau_tsne <- ggplot() + 
  geom_point(
    data = tsne_gau_scaled,
    aes(
      x = tSNE1,
      y = tSNE2,
      colour = data_with_clust$cluster
    ),
    alpha = 0.1) + 
  geom_segment(
    data = tr_from_to_df_gau1,
    aes(
      x = x_from,
      y = y_from,
      xend = x_to,
      yend = y_to),
    colour = "#000000",
    linewidth = 0.5) +
  scale_color_manual(values = c('#66c2a5','#fc8d62','#8da0cb','#e78ac3','#a6d854')) +
  interior_annotation("a1", 
                      position = c(0.08, 0.95),
                      cex = 2) +
  theme(
    aspect.ratio = 1
  )

```

```{r}
#| label: five-gau-tsne-projs
## First projection
proj_obj1 <- read_rds("data/five_gau_clusters/tsne_gau_proj_obj1.rds")
proj_obj1[["cluster"]] <- data_with_clust$cluster

five_gau_proj_tsne_model1 <- plot_proj(
  proj_obj = proj_obj1,
  point_param = c(1.5, 0.05),
  line_param = c(0.8, 0.4, "#000000"),
  plot_limits = c(-0.89, 0.89),
  axis_text_size = 5,
  is_category = TRUE
) +
  scale_color_manual(values = c('#66c2a5','#fc8d62','#8da0cb','#e78ac3','#a6d854')) +
  interior_annotation("a2", 
                      position = c(0.08, 0.9),
                      cex = 2) 

## Second projection

proj_obj2 <- read_rds("data/five_gau_clusters/tsne_gau_proj_obj2.rds")

five_gau_proj_tsne_model2 <- plot_proj(
  proj_obj = proj_obj2,
  point_param = c(1.5, 0.05, clr_choice), # size, alpha, color
  line_param = c(0.8, 0.4, "#000000"), # linewidth, alpha
  plot_limits = c(-1.3, 1.3), 
  axis_text_size = 4, 
  is_category = FALSE) +
  interior_annotation(label = "a3", position = c(0.08, 0.9), cex = 2) +
  theme(aspect.ratio = 1,
        legend.position = "none") 
```

<!--five_gau_clusters/03_gen_model_with_UMAP.R-->
```{r}
#| label: five-gau-proj-umap-model

umap_gau_scaled <- read_rds("data/five_gau_clusters/umap_gau_scaled.rds")
tr_from_to_df_gau1_umap <- read_rds("data/five_gau_clusters/tr_from_to_df_gau1_umap.rds")

trimesh_removed_gau_umap <- ggplot() + 
  geom_point(
    data = umap_gau_scaled,
    aes(
      x = UMAP1,
      y = UMAP2,
      colour = data_with_clust$cluster
    ),
    alpha = 0.05) + 
  geom_segment(
    data = tr_from_to_df_gau1_umap,
    aes(
      x = x_from,
      y = y_from,
      xend = x_to,
      yend = y_to),
    colour = "#000000",
    #alpha = 0.4,
    linewidth = 0.5) +
  scale_color_manual(values = c('#66c2a5','#fc8d62','#8da0cb','#e78ac3','#a6d854')) +
  interior_annotation("b1", 
                      position = c(0.95, 0.95),
                      cex = 2) +
  theme(
    aspect.ratio = 1
  )
```

```{r}
#| label: five-gau-umap-projs

## First projection
proj_obj1 <- read_rds("data/five_gau_clusters/umap_gau_proj_obj1.rds")
proj_obj1[["cluster"]] <- data_with_clust$cluster

five_gau_proj_umap_model1 <- plot_proj(
  proj_obj = proj_obj1,
  point_param = c(1.5, 0.05),
  line_param = c(0.8, 0.4, "#000000"),
  plot_limits = c(-0.89, 0.89),
  axis_text_size = 5,
  is_category = TRUE
) +
  scale_color_manual(values = c('#66c2a5','#fc8d62','#8da0cb','#e78ac3','#a6d854')) +
  interior_annotation("b2", 
                      position = c(0.08, 0.9),
                      cex = 2) 

## Second projection
proj_obj2 <- read_rds("data/five_gau_clusters/umap_gau_proj_obj2.rds")

five_gau_proj_umap_model2 <- plot_proj(
  proj_obj = proj_obj2,
  point_param = c(1.5, 0.05, clr_choice), # size, alpha, color
  line_param = c(0.8, 0.4, "#000000"), # linewidth, alpha
  plot_limits = c(-1.3, 1.3), 
  axis_text_size = 4, 
  is_category = FALSE) +
  interior_annotation(label = "b3", position = c(0.08, 0.9), cex = 2) +
  theme(aspect.ratio = 1,
        legend.position = "none")
```

<!--five_gau_clusters/04_gen_model_with_PaCMAP.R-->
```{r}
#| label: five-gau-proj-pacmap-model

pacmap_gau_scaled <- read_rds("data/five_gau_clusters/pacmap_gau_scaled.rds")
tr_from_to_df_gau1_pacmap <- read_rds("data/five_gau_clusters/tr_from_to_df_gau1_pacmap.rds")
  
trimesh_removed_gau_pacmap <- ggplot() + 
  geom_point(
    data = pacmap_gau_scaled,
    aes(
      x = PaCMAP1,
      y = PaCMAP2,
      colour = data_with_clust$cluster),
    alpha = 0.1) +
  geom_segment(
    data = tr_from_to_df_gau1_pacmap,
    aes(
      x = x_from,
      y = y_from,
      xend = x_to,
      yend = y_to),
    colour = "#000000",
    #alpha = 0.4,
    linewidth = 0.5) +
  scale_color_manual(values = c('#66c2a5','#fc8d62','#8da0cb','#e78ac3','#a6d854')) +
  interior_annotation("c1", 
                      position = c(0.08, 0.95),
                      cex = 2) +
  theme(
    aspect.ratio = 1
  )
```

```{r}
#| label: five-gau-pacmap-projs
## First projection
proj_obj1 <- read_rds("data/five_gau_clusters/pacmap_gau_proj_obj1.rds")
proj_obj1[["cluster"]] <- data_with_clust$cluster

five_gau_proj_pacmap_model1 <- plot_proj(
  proj_obj = proj_obj1,
  point_param = c(1.5, 0.05),
  line_param = c(0.8, 0.4, "#000000"),
  plot_limits = c(-0.89, 0.89),
  axis_text_size = 5,
  is_category = TRUE
) +
  scale_color_manual(values = c('#66c2a5','#fc8d62','#8da0cb','#e78ac3','#a6d854')) +
  interior_annotation("c2",
                      position = c(0.08, 0.9),
                      cex = 2)

## Second projection
proj_obj2 <- read_rds("data/five_gau_clusters/pacmap_gau_proj_obj2.rds")

five_gau_proj_pacmap_model2 <- plot_proj(
  proj_obj = proj_obj2,
  point_param = c(1.5, 0.05, clr_choice), # size, alpha, color
  line_param = c(0.8, 0.4, "#000000"), # linewidth, alpha
  plot_limits = c(-1.3, 1.3), 
  axis_text_size = 4, 
  is_category = FALSE) +
  interior_annotation(label = "c3", position = c(0.08, 0.9), cex = 2) +
  theme(aspect.ratio = 1,
        legend.position = "none")

```

```{r}
#| label: fig-five-gau-projs
#| fig-height: 10
#| fig-width: 15
#| fig-pos: "!ht"
#| fig-cap: "NLDR's organize points in the \\gD{} layout in different ways, possibly misleadingly, illustrated using three layouts: (a) tSNE, (b) UMAP, (c) PaCMAP. The data has five Gaussian clusters in \\fD{}. The bottom row of plots shows a \\gD{} projection from a tour on \\fD{} revealing the differences generated by the layouts on the model fits.  We would expect the model fit to be like that in (a2) where it is distinctly is separate for each cluster but like a hairball in each. This would indicate the distinct clusters, each being fully \\fD{}. With (c2), the curiousity is that the model is a \\gD{} pancake shape in \\fD{}, indicating that there is some ordering of points done by PaCMAP, posisbly along some principal component axes. Videos of the langevitour animations are available at <https://youtu.be/I-kxCwVfqiQ>, <https://youtu.be/gD1P01FUPyU>, and <https://youtu.be/MxJ_srOFQNk> respectively."

trimesh_removed_gau_tsne + trimesh_removed_gau_umap + trimesh_removed_gau_pacmap +
five_gau_proj_tsne_model1 + five_gau_proj_umap_model1 + five_gau_proj_pacmap_model1 + 
  plot_layout(guides = "collect", nrow = 2) &
  theme(legend.position='none')
```

### Sparseness creates a contracted \gD{} layout {#sec-effect-dens}

Differences in density can arise by sampling at different rates in different subspaces of \pD{}. For example, the data shown in @fig-one-dens_clust-error all lies on a \gD{} curved sheet in \fD{}, but one end of the sheet is sampled densely and the other very sparsely. It was simulated to illustrate the effect of the density difference on layout generated by an NLDR, illustrated using the tSNE results, but it happens with all methods. 

@fig-one-dens_clust-error (a2, b2) shows a \gD{} layout for tSNE created using the default hyper-parameters. One would expect to see a rectangular shape if the curved sheet is flattened, but the layout is triangular. The other two displays show the residuals as a dot density plot (a1, b1), and a \gD{} projection of the data and the model from \fD{} (a3, b3). Using linked brushing between the plots, we can highlight points in the tSNE layout, and examine where they fall in the original \fD{}. The darker (maroon) points indicate points that have been highlighted by linking. In row a, the points at the top of the triangle are highlighted, and we can see these correspond to higher residuals, and also to all points at the low density end of the curved sheet. In row b, points at the lower left side of the triangle are highlighted which corresponds to smaller residuals and one corner of the sheet at the high density end of the curved sheet. 

The tSNE behaviour is to squeeze the low density area of the data together into the layout. This is common in other NLDR methods also, which means analysts need to be aware that if their data is not sampled relatively uniformly, apparent closeness in the \gD{} may correspond to sparseness in \pD{}.

<!--c_shaped_dens_str/03_gen_model_with_tSNE.R-->
```{r}
#| label: error-model-one-dens-clust

error_df_one_curvy_abs_selected11 <- read_rds("data/one_c_shaped_dens_structure/error_df_one_curvy_abs_selected11.rds")

error_df_one_curvy_abs_deselected11 <- read_rds("data/one_c_shaped_dens_structure/error_df_one_curvy_abs_deselected11.rds")

error_plot_one_curvy_hist_selected <- ggplot(error_df_one_curvy_abs_deselected11, 
      aes(x = sqrt_row_wise_total_error, 
          y = density)) +
  geom_point(alpha=0.7, colour = "#d9d9d9") +
  geom_point(data = error_df_one_curvy_abs_selected11, 
      aes(x = sqrt_row_wise_total_error, 
          y = density),
      alpha=0.7, colour = "#800026", size = 3) +
  xlab(expression(paste("residual (",e[hi],")"))) +
  ylab("") +
  interior_annotation("b1",
                      cex = 2) +
  theme_bw() +
  theme(aspect.ratio = 1,
        axis.title = element_text(size = 18),
        axis.text.x = element_text(size = 15),
        axis.text.y = element_blank())

error_df_one_curvy_abs_selected12 <- read_rds("data/one_c_shaped_dens_structure/error_df_one_curvy_abs_selected12.rds")

error_df_one_curvy_abs_deselected12 <- read_rds("data/one_c_shaped_dens_structure/error_df_one_curvy_abs_deselected12.rds")

error_plot_one_curvy_hist <- ggplot(error_df_one_curvy_abs_deselected12, 
      aes(x = sqrt_row_wise_total_error, 
          y = density)) +
  geom_point(alpha=0.7, colour = "#d9d9d9") +
  geom_point(data = error_df_one_curvy_abs_selected12, 
      aes(x = sqrt_row_wise_total_error, 
          y = density),
      alpha=0.7, colour = "#800026", size = 3) +
  xlab(expression(paste("residual (",e[hi],")"))) +
  ylab("") +
  interior_annotation("a1",
                      cex = 2) +
  theme_bw() +
  theme(aspect.ratio = 1,
        axis.title = element_text(size = 18),
        axis.text = element_text(size = 15),
        axis.text.y = element_blank())
```

```{r}
#| label: link-model-one-dens-clust

plot_tsne_dens <- error_df_one_curvy_abs_deselected12 |>
  ggplot(aes(x = emb1,
             y = emb2)) +
  geom_point(alpha=0.7, colour = "#d9d9d9") +
  geom_point(data = error_df_one_curvy_abs_selected12,
             aes(x = emb1, y = emb2),
             alpha=0.7, colour = "#800026", size = 3) +
  theme(aspect.ratio = 1) +
  interior_annotation("a2",
                      position = c(0.08, 0.9),
                      cex = 2)

plot_tsne_dens_selected <- error_df_one_curvy_abs_deselected11 |>
  ggplot(aes(x = emb1,
             y = emb2)) +
  geom_point(alpha=0.7, colour = "#d9d9d9") +
  geom_point(data = error_df_one_curvy_abs_selected11,
             aes(x = emb1, y = emb2),
             alpha=0.7, colour = "#800026", size = 3) +
  theme(aspect.ratio = 1) +
  interior_annotation("b2",
                      position = c(0.08, 0.9),
                      cex = 2)
```

```{r}
#| label: proj-one-dens-clust

## First projection

projected_df <- read_rds("data/one_c_shaped_dens_structure/projected_df1.rds")
model_df <- read_rds("data/one_c_shaped_dens_structure/model_df1.rds")
axes <- read_rds("data/one_c_shaped_dens_structure/axes1.rds")
circle <- read_rds("data/one_c_shaped_dens_structure/circle1.rds")

five_c_shaped_proj_tsne_model1 <- ggplot() +
      geom_point(
        data = projected_df,
        aes(
          x = proj1,
          y = proj2,
          colour = factor(cluster, levels = c("deselected", "selected")),
          size = factor(cluster, levels = c("deselected", "selected"))),
        alpha = 0.5) +
      geom_segment(
        data = model_df,
        aes(
          x = proj1_from,
          y = proj2_from,
          xend = proj1_to,
          yend = proj2_to),
        colour = "#000000",
        linewidth = 0.8,
        alpha = 0.4) +
      geom_segment(
          data=axes,
          aes(x=x1, y=y1, xend=x2, yend=y2),
          colour="grey70") +
        geom_text(
          data=axes,
          aes(x=x2, y=y2),
          label=rownames(axes),
          colour="grey50",
          size = 4.5) +
        geom_path(
          data=circle,
          aes(x=c1, y=c2), colour="grey70") +
        xlim(c(-0.35, 0.35)) +
        ylim(c(-0.35, 0.35)) +
        interior_annotation(label = "a3", position = c(0.08, 0.9), cex = 2) +
        scale_color_manual(values = c('#d9d9d9', '#800026')) +
        scale_size_manual(values = c('deselected' = 1.5, 'selected' = 3)) +
        theme(aspect.ratio = 1,
              legend.position = "none")

projected_df <- read_rds("data/one_c_shaped_dens_structure/projected_df2.rds")
model_df <- read_rds("data/one_c_shaped_dens_structure/model_df2.rds")
axes <- read_rds("data/one_c_shaped_dens_structure/axes2.rds")
circle <- read_rds("data/one_c_shaped_dens_structure/circle2.rds")

five_c_shaped_proj_tsne_model1_selected <- ggplot() +
      geom_point(
        data = projected_df,
        aes(
          x = proj1,
          y = proj2,
          colour = factor(cluster, levels = c("deselected", "selected")),
          size = factor(cluster, levels = c("deselected", "selected"))),
        alpha = 0.5) +
      geom_segment(
        data = model_df,
        aes(
          x = proj1_from,
          y = proj2_from,
          xend = proj1_to,
          yend = proj2_to),
        colour = "#000000",
        linewidth = 0.8,
        alpha = 0.4) +
      geom_segment(
          data=axes,
          aes(x=x1, y=y1, xend=x2, yend=y2),
          colour="grey70") +
        geom_text(
          data=axes,
          aes(x=x2, y=y2),
          label=rownames(axes),
          colour="grey50",
          size = 4.5) +
        geom_path(
          data=circle,
          aes(x=c1, y=c2), colour="grey70") +
        xlim(c(-0.35, 0.35)) +
        ylim(c(-0.35, 0.35)) +
  interior_annotation(label = "b3", position = c(0.08, 0.9), cex = 2) +
  scale_color_manual(values = c('#d9d9d9', '#800026')) +
  scale_size_manual(values = c('deselected' = 1.5, 'selected' = 3)) +
  theme(aspect.ratio = 1,
        legend.position = "none")
```

```{r}
#| label: fig-one-dens_clust-error
#| fig-width: 15
#| fig-height: 10
#| fig-pos: "!ht"
#| out-width: 100%
#| fig-cap: "Exploring the effect of density on the NLDR layout using a \\gD{} curved sheet in \\fD{} with different density at each end. Three plots are linked: density plot of residuals (a1, b1), NLDR layout (a2, b2), projection of \\fD{} model and data (a3, b3). The brown points indicate the selected set, which are different in each row. In (a2), the top part of the triangular shape is selected which corresponds to higher residuals (a1) and the sparse end of the structure (a3). In (b2) one of other corners is highlighted, which can be seen to correspond to low residuals (b1) and one side of the dense end of the data (b3). While the tSNE layout represents the dense end of the sheet correctly as two corners in the layout, it contracts the sparse end of the sheet into a single corner. Video of the langevitour animation is available at <https://youtu.be/-KsQH0rII2A>."

error_plot_one_curvy_hist + plot_tsne_dens + 
  five_c_shaped_proj_tsne_model1 +
  error_plot_one_curvy_hist_selected + plot_tsne_dens_selected +
  five_c_shaped_proj_tsne_model1_selected +
    plot_layout(guides = "collect", nrow = 2) &
    theme(legend.position='none')
```


## PBMC3k: comparison with results of scDEED recommendations

<!-- ### Comparison with results of scDEED recommendations -->

<!-- XXX What this section needs: -->

<!-- - connect this with previous section, not make it seem like a completely new. connections are: nldr used to show clustering, scDEED provides a different approach to decide which is best. -->
<!-- - why these two? assume one is the published figure, and second is one recommended by scDEED, which our results confirm is better. -->
<!-- - Refer to figure in text -->
<!-- - Fig 16 not necessary -->

<!-- CHANGES WERE MADE BASED ON THIS THINKING -->



<!-- <!-- In the field of single-cell studies, clustering is a common analytical task used to identify groups of cells with similar expression profiles. Non-linear dimensional reduction (NLDR) methods are frequently employed to visualize these clusters and help validate the results. However, it is well known that the 2D embeddings produced by t-SNE and UMAP may not accurately reflect the similarities among cell clusters. ALREADY STATED IN PREVIOUS SECTION -->  

As we were writing this paper @xia2023 appeared proposing a new method called scDEED helping to assess the validity of a \gD{} embedding. scDEED calculates a reliability score for each cell embedding based on the similarity between the cell’s \gD{} embedding neighbors and its neighbors prior to embedding. A low reliability score suggests a dubious embedding. It can help in the deciding on optimal hyper-parameters. Here we illustrate how our method compares with the results from scDEED.

Note that @xia2023 uses a different PBMC dataset than that used by @chen2024, shown by us in the main paper example, which is why this comparison is shown here and not in the main paper. Their data contains $31,021$ cells including cell type labels, and the gene expression levels were in the unit of log-transformed UMI count per $10,000$. They focused on three sequencing methods (inDrops, DropSeq, and SeqWell) and four common cell types Cytotoxic T cell, CD4+T cell, CD14+ Monocyte, and B cell. Pre-processing follows the process in @xia2023 again using the Human Peripheral Blood Mononuclear Cells (PBMC) data. 

For illustration purposes, we only selected cells generated with inDrops ($n=5858$ cells). Also, @xia2023 used first $9$ principal components to generate the UMAP and tSNE with default hyper-parameters. The objective is to what scDEED suggests is the best layout with what HBE would choose. Layout a (@fig-pbmc-mse-umap) is generated from the hyper-parameters suggested by @chen2024, and layout b (@fig-pbmc-mse-umap) is with suggested hyper-parameters by scDEED to be more accurate. <!--Layouts a and b contain $46$ and $83$ dubious cells respectively.--> The HBE vs binwidth ($a_1$) plot (@fig-pbmc-mse-umap) illustrates that our approach would suggest that scDEED is correct here, that layout b is more accurately reflecting the cluster structure in the PBMC data. This is also supported by examining the models in the data space as shown in @fig-model-pbmc-author-proj.

<!-- @fig-pbmc-mse-umap compares the metrics rARNX, rRTA, rSC, rGS, along with HBE computed on $a_1=0.04$. This is a parallel coordinate plot where the y-axis shows a normalized score to ensure the metrics are on the same scale. Each line corresponds to one layout. Most metrics (rSC, rARNX, and HBE) consistently indicate that the optimized layout (b) provides a better representation, while rRTA, and rGS slightly favor the published layout. -->

```{r}
#| label: read-pbmc-umap-layouts
# Read a variety of different NLDR representations of PBMC
# and plot them on same aspect ratio
umap_pbmc <- read_rds("data/pbmc3k/pbmc_scdeed_umap_n_neighbors_30_min_dist_0.3.rds")
umap_pbmc <- as_tibble(umap_pbmc)
names(umap_pbmc) <- c("emb1", "emb2")

nldr1c <- umap_pbmc |>
  ggplot(aes(x = emb1,
             y = emb2)) +
  geom_point(alpha=0.1, size=1, colour='#e41a1c') +
  interior_annotation("a")

tsne_pbmc <- read_rds("data/pbmc3k/pbmc_scdeed_tsne_perplexity_30.rds")
tsne_pbmc <- as_tibble(tsne_pbmc)
names(tsne_pbmc) <- c("emb1", "emb2")

nldr2c <- tsne_pbmc |>
  ggplot(aes(x = emb1,
             y = emb2))+
  geom_point(alpha=0.1, size=1, colour='#377eb8') +
  interior_annotation("b")

```

<!--pbmc3k/11_gen_mse_for_diff_tsne_scD.R-->
<!--pbmc3k/11_gen_mse_for_diff_umap_scD.R-->
```{r}
#| label: combine-data-pbmc-umap-tsne

error_pbmc_umap <- read_rds("data/pbmc3k/error_scdeed_pbmc_umap_30_min_dist_0.3.rds")
error_pbmc_tsne <- read_rds("data/pbmc3k/error_scdeed_pbmc_tsne_perplexity_30.rds")

error_pbmc <- bind_rows(error_pbmc_umap,
                        error_pbmc_tsne)

error_pbmc_avg <- error_pbmc |>
  filter(b1 >= 5) |>
  group_by(method, n_bar) |>
  filter(HBE == min(HBE)) |>
  ungroup()

error_pbmc_scdeed <- error_pbmc |> 
  dplyr::filter(b1 == 30) |>
  dplyr::select(method, HBE)

error_pbmc <- error_pbmc |>
  mutate(a1 = round(a1, 2)) |>
  filter(b1 >= 5) |>
  group_by(method, a1) |>
  filter(HBE == min(HBE)) |>
  ungroup()
```

```{r}
#| label: error-comp-pbmc-umap

error_umap_plot_pbmc <-  plot_hbe(error_pbmc) +
  interior_annotation("i", position = c(0.07, 0.9)) +
  scale_x_continuous(breaks = sort(unique(error_pbmc$a1))[c(1, 5, 9, 12, 15, 18, 20, 21)]) +
  scale_color_manual(values=c('#377eb8', '#e41a1c')) 

error_umap_plot_pbmc_nbar <-  plot_hbe_nbar(error_pbmc_avg) +
  annotate("text", x = 5, y = 4.5, label = "ii", size = 4, colour = "grey70") +
  scale_color_manual(values=c('#377eb8', '#e41a1c')) 
```

```{r}
nldr_metrics <- read_rds("data/pbmc3k/nldr_eval_metrics_scdeed.rds")

nldr_metrics <- inner_join(nldr_metrics, error_pbmc_scdeed, by = c("Method" = "method")) |>
  rename("sc" = "r") |>
  mutate(`rARNX` = 1 - `AUC(R_NX)`) |>
  mutate(`rRTA` = 1 - RTA) |>
  mutate(`rSC` = 1 - sc) |>
  mutate(`rGS` = 1 -  GS)

comp_metrics_pbmc_scdeed <- plot_comp_metrics(nldr_metrics) +
  interior_annotation("iii") +
  interior_annotation("better", position = c(0.09, 0.15)) +
  interior_annotation("worse", position = c(0.09, 0.9)) +
  scale_color_manual(values=c('#377eb8', '#e41a1c'))
```

```{r}
#| label: fig-pbmc-mse-umap
#| fig-pos: "!ht"
#| out-width: 100%
#| fig-cap: "Comparing the published layout (a) with what would be suggested to be optimal by scDEED (b), using HBE for varying (i) binwidth ($a_1$), and (ii) average bin count ($\\bar{n}_h$), on a subset of PBMC3k data. Color represents NLDR layouts. HBE would corroborate that the scDEED optimized layout is better than what was originally published. Plot (ii), which accounts for the density within clusters by using average bin count, shows reduced differences between layouts, indicating that part of the variation in (i) is driven by cluster density rather than true structural differences. Comparison of scaled evaluation metrics (iii) (rRTA, rSC, rGS, rARNX, and HBE using $a_1=0.04$) for two NLDR layouts of the PBMC3k data the originally published layout (a) and the scDEED optimized layout (b). Each line represents a layout, with color matching the corresponding scatterplots. Most metrics (rSC, rARNX, and HBE) consistently indicate that the optimized layout (b) provides a better representation, while rRTA, and rGS slightly favor the published layout."
#| fig-width: 6
#| fig-height: 4

# Second row: horizontal stack
bottom_row <- nldr1c + nldr2c + 
  plot_layout(ncol = 2, widths = c(50, 50))

# Combine top and bottom rows vertically
final_plot <- (error_umap_plot_pbmc | bottom_row)/(error_umap_plot_pbmc_nbar | comp_metrics_pbmc_scdeed) + plot_layout(heights = c(0.7, 1))

final_plot
```

<!--pbmc3k/15_gen_model_with_UMAP_scD.R-->
```{r}
#| label: hexbin-pbmc

umap_pbmc_scaled_with_cluster <- read_rds("data/pbmc3k/umap_pbmc_scaled_with_cluster_scDEED.rds")

tr_from_to_df_pbmc <- read_rds("data/pbmc3k/tr_from_to_df_pbmc_scDEED.rds")

trimesh_removed_pbmc <- ggplot() + 
  geom_point(
    data = umap_pbmc_scaled_with_cluster,
    aes(
      x = emb1,
      y = emb2,
      color = cluster
    ),
    alpha = 0.3#,
    #size = 0.1
  )  +
  geom_segment(data = tr_from_to_df_pbmc, 
               aes(
                 x = x_from, 
                 y = y_from, 
                 xend = x_to, 
                 yend = y_to),
               colour = "#000000",
               linewidth = 1) +
  scale_color_manual(values = c("#8dd3c7", "#fdb462", "#bebada")) +
  interior_annotation("a1", cex = 2) +
  theme(
    aspect.ratio = 1
  )
```


```{r}
#| label: pbmc-umap-model-proj

proj_obj1 <- read_rds("data/pbmc3k/proj_obj1_umap_scDEED.rds")

pbmc_proj_umap_model1 <- plot_proj(
  proj_obj = proj_obj1, 
  point_param = c(1.5, 0.2), # size, alpha, color
  plot_limits = c(-0.04, 0.045), 
  axis_text_size = 4,
  is_category = TRUE) + 
  interior_annotation(label = "a2", cex = 2) +
  scale_color_manual(values = c("#8dd3c7", "#fdb462", "#bebada"))

proj_obj2 <- read_rds("data/pbmc3k/proj_obj2_umap_scDEED.rds")

pbmc_proj_umap_model2 <- plot_proj(
    proj_obj = proj_obj2, 
    point_param = c(1.5, 0.2), # size, alpha, color
    plot_limits = c(-0.03, 0.03), 
    axis_text_size = 4,
    is_category = TRUE) +
  interior_annotation(label = "a3", cex = 2) +
  scale_color_manual(values = c("#8dd3c7", "#fdb462", "#bebada"))

```

<!--pbmc3k/16_gen_model_with_tSNE_scD.R-->
```{r}
#| label: hexbin-pbmc2

tsne_pbmc_scaled_with_cluster <- read_rds("data/pbmc3k/tsne_pbmc_scaled_with_cluster_scDEED.rds")

tr_from_to_df_pbmc <- read_rds("data/pbmc3k/tr_from_to_df_pbmc_tsne_scDEED.rds")

trimesh_removed_pbmc_best <- ggplot() + 
  geom_point(
    data = tsne_pbmc_scaled_with_cluster,
    aes(
      x = emb1,
      y = emb2,
      color = cluster
    ),
    alpha = 0.3#,
    #size = 0.1
  )  +
  geom_segment(data = tr_from_to_df_pbmc, 
               aes(
                 x = x_from, 
                 y = y_from, 
                 xend = x_to, 
                 yend = y_to),
               colour = "#000000",
               linewidth = 1) +
  scale_color_manual(values = c("#8dd3c7", "#fdb462", "#bebada")) +
  interior_annotation("b1", cex = 2) +
  theme(
    aspect.ratio = 1
  )
```

```{r}
#| label: pbmc-tsne-model-proj-best

proj_obj1 <- read_rds("data/pbmc3k/proj_obj1_tsne_scDEED.rds")

pbmc_proj_tsne_model1_best <- plot_proj(
  proj_obj = proj_obj1, 
  point_param = c(1.5, 0.2), # size, alpha, color
  plot_limits = c(-0.04, 0.045), 
  axis_text_size = 4,
  is_category = TRUE) + 
  interior_annotation(label = "b2", cex = 2) +
  scale_color_manual(values = c("#8dd3c7", "#fdb462", "#bebada"))

proj_obj2 <- read_rds("data/pbmc3k/proj_obj2_tsne_scDEED.rds")

pbmc_proj_tsne_model2_best <- plot_proj(
  proj_obj = proj_obj2, 
  point_param = c(1.5, 0.2), # size, alpha, color
  plot_limits = c(-0.03, 0.03), 
  axis_text_size = 4,
  is_category = TRUE) + 
  interior_annotation(label = "b3", cex = 2) +
  scale_color_manual(values = c("#8dd3c7", "#fdb462", "#bebada"))

```

```{r}
#| echo: false
#| fig-cap: "Compare the published \\gD{} layout (@fig-pbmc-mse-umap a) made with UMAP and the \\gD{} layout made with tSNE selected as optimal by scDEED (@fig-pbmc-mse-umap b) and also HBE (@fig-pbmc-mse-umap). The two plots on the right show projections from a tour, with the models overlaid. The published layout a suggested three separated clusters with two of them are close, but this is not present in the data. While there may be three clusters they are not well-separated. The difference in model fit also indicates this: the published layout a does not capture the nonlinear structure of the clusters like the model generated from layout b. This supports the choice that layout b is the better representation of the data, because it shows close clusters. Videos of the langevitour animations are available at <https://youtu.be/ffiB4MGWyn8> and <https://youtu.be/e7XNL18co1c> respectively."
#| label: fig-model-pbmc-author-proj
#| fig-pos: "!ht"
#| fig-width: 10
#| fig-height: 15
#| out-width: 90%

free(trimesh_removed_pbmc) + free(trimesh_removed_pbmc_best) +
  pbmc_proj_umap_model1 + pbmc_proj_tsne_model1_best +
  pbmc_proj_umap_model2 + pbmc_proj_tsne_model2_best +
  plot_layout(nrow=2, byrow=FALSE) &
  theme(legend.position='none')
```

## Compare HBE with existing evaluation metrics

@fig-comp-metric-pbmc and @fig-comp-metric-mnist compare HBE with commonly used evaluation metrics such as rRTA, rARNX, rSC, and rGS across multiple NLDR layouts. These visual comparisons highlight that HBE behaves differently from these existing metrics due to the different settings involved.

```{r}
#| label: read-two-curvy-clust-nldr
# Read a variety of different NLDR representations of two_non_linear_diff_shaped_close_clusters
# and plot them on same aspect ratio
tsne_two_curvy <- read_rds("data/two_nonlinear/two_non_linear_diff_shaped_close_clusters_tsne_perplexity_47.rds")

nldr_two_curvy1 <- tsne_two_curvy |>
  ggplot(aes(x = tSNE1,
             y = tSNE2))+
  geom_point(alpha=0.1, size=1, colour='#a65628') +
  interior_annotation("a") +
  theme(aspect.ratio = 1) 

tsne_two_curvy <- read_rds("data/two_nonlinear/two_non_linear_diff_shaped_close_clusters_tsne_perplexity_62.rds")

nldr_two_curvy6 <- tsne_two_curvy |>
  ggplot(aes(x = tSNE1,
             y = tSNE2))+
  geom_point(alpha=0.1, size=1, colour='#984ea3') +
  interior_annotation("f") +
  theme(aspect.ratio = 1) 

umap_two_curvy <- read_rds("data/two_nonlinear/two_non_linear_diff_shaped_close_clusters_umap_n-neigbors_15_min-dist_0.1.rds")

nldr_two_curvy2 <- umap_two_curvy |>
  ggplot(aes(x = UMAP1,
             y = UMAP2)) +
  geom_point(alpha=0.1, size=1, colour='#636363') +
  interior_annotation("b") +
  theme(aspect.ratio = 1) 

phate_two_curvy <- read_rds("data/two_nonlinear/two_non_linear_diff_shaped_close_clusters_phate_knn_5.rds")

nldr_two_curvy3 <- phate_two_curvy |>
  ggplot(aes(x = PHATE1,
             y = PHATE2))+
  geom_point(alpha=0.1, size=1, colour='#ff7f00') +
  interior_annotation("c") +
  theme(aspect.ratio = 1) 

trimap_two_curvy <- read_rds("data/two_nonlinear/two_non_linear_diff_shaped_close_clusters_trimap_n-inliers_12_n-outliers_4_n-random_3.rds")

nldr_two_curvy4 <- trimap_two_curvy |>
  ggplot(aes(x = TriMAP1,
             y = TriMAP2))+
  geom_point(alpha=0.1, size=1, colour='#4daf4a') +
  interior_annotation("d") +
  theme(aspect.ratio = 1)

pacmap_two_curvy <- read_rds("data/two_nonlinear/two_non_linear_diff_shaped_close_clusters_pacmap_n-neighbors_10_init_random_MN-ratio_0.5_FP-ratio_2.rds")

nldr_two_curvy5 <- pacmap_two_curvy |>
  ggplot(aes(x = PaCMAP1,
             y = PaCMAP2))+
  geom_point(alpha=0.1, size=1, colour='#e41a1c') +
  interior_annotation("e") +
  theme(aspect.ratio = 1)
```

<!--two_nonlinear/04_gen_mse_for_diff_methods.R-->
```{r}
#| label: combine-data-two_curvy

error_two_curvy_umap <- read_rds("data/two_nonlinear/error_two_non_linear_diff_shaped_close_clusters_umap.rds")
error_two_curvy_tsne <- read_rds("data/two_nonlinear/error_two_non_linear_diff_shaped_close_clusters_tsne.rds")
error_two_curvy_phate <- read_rds("data/two_nonlinear/error_two_non_linear_diff_shaped_close_clusters_phate.rds")
error_two_curvy_trimap <- read_rds("data/two_nonlinear/error_two_non_linear_diff_shaped_close_clusters_trimap.rds")
error_two_curvy_pacmap <- read_rds("data/two_nonlinear/error_two_non_linear_diff_shaped_close_clusters_pacmap.rds")
error_two_curvy_tsne2 <- read_rds("data/two_nonlinear/error_two_non_linear_diff_shaped_close_clusters_tsne2.rds")

error_two_curvy_avg <- bind_rows(error_two_curvy_umap, 
                         error_two_curvy_tsne,
                         error_two_curvy_phate,
                         error_two_curvy_trimap,
                         error_two_curvy_pacmap,
                         error_two_curvy_tsne2)

error_two_curvy_avg <- error_two_curvy_avg |>
  filter(b1 >= 5) |>
  group_by(method, n_bar) |>
  filter(HBE == min(HBE)) |>
  ungroup() |>
  mutate(method = factor(method, levels = c("PaCMAP", "PHATE", "TriMAP", "tSNE", "UMAP", "tSNE2")))

error_two_curvy <- bind_rows(error_two_curvy_umap, 
                         error_two_curvy_tsne,
                         error_two_curvy_phate,
                         error_two_curvy_trimap,
                         error_two_curvy_pacmap,
                         error_two_curvy_tsne2)

error_two_curvy_filtered <- error_two_curvy |> 
  dplyr::filter(b1 == 24) |>
  dplyr::select(method, HBE)

error_two_curvy <- error_two_curvy |>
  mutate(a1 = round(a1, 2)) |>
  filter(b1 >= 5) |>
  group_by(method, a1) |>
  filter(HBE == min(HBE)) |>
  ungroup() |>
  mutate(method = factor(method, levels = c("PaCMAP", "PHATE", "TriMAP", "tSNE", "UMAP", "tSNE2")))
```

```{r}
#| label: error-comp-two_curvy

error_plot_two_curvy <-
  plot_hbe(error_two_curvy) +
  interior_annotation("i", position = c(0.07, 0.9)) +
  scale_x_continuous(breaks = seq(0, 0.25, 0.05)) +
    #sort(unique(error_two_curvy$a1))[
    #  seq(1, length(
    #    unique(error_two_curvy$a1)), 
    #    by = 5)]) +
  scale_color_manual(
    values=c("PaCMAP" = '#e41a1c', "PHATE" = '#ff7f00', 
             "TriMAP" = '#4daf4a', "tSNE" = '#a65628', 
             "UMAP" = '#636363', "tSNE2" = '#984ea3')) +
  theme(aspect.ratio = 0.685) 


error_plot_two_curvy_n_bar <-
  plot_hbe_nbar(error_two_curvy_avg) +
  annotate("text", x = 2.5, y = 1.3, label = "ii", size = 4, colour = "grey70") +
  scale_color_manual(
    values=c("PaCMAP" = '#e41a1c', "PHATE" = '#ff7f00', 
             "TriMAP" = '#4daf4a', "tSNE" = '#a65628', 
             "UMAP" = '#636363', "tSNE2" = '#984ea3')) +
  theme(aspect.ratio = 0.685) 
```

<!--two_nonlinear/07_example_evaluation_metrics.R-->
```{r}
nldr_metrics <- read_rds("data/two_nonlinear/nldr_eval_metrics.rds")

nldr_metrics <- inner_join(nldr_metrics, error_two_curvy_filtered, by = c("Method" = "method")) |>
  rename("sc" = "r") |>
  mutate(`rARNX` = 1 - `AUC(R_NX)`) |>
  mutate(`rRTA` = 1 - RTA) |>
  mutate(`rSC` = 1 - sc) |>
  mutate(`rGS` = 1 -  GS)

comp_metrics <- plot_comp_metrics(nldr_metrics) +
  interior_annotation("iii") +
  interior_annotation("better", position = c(0.07, 0.15)) +
  interior_annotation("worse", position = c(0.07, 0.9)) +
  scale_color_manual(
    values=c("PaCMAP" = '#e41a1c', "PHATE" = '#ff7f00', 
             "TriMAP" = '#4daf4a', "tSNE" = '#a65628', 
             "UMAP" = '#636363', "tSNE2" = '#984ea3')) +
  theme(aspect.ratio = 0.22)
```

```{r}
#| fig-cap: "Assessing which of the 6 NLDR layouts (a-f) on the 2NC7 data is the better representation using HBE for varying (i) binwidth ($a_1$), and (ii) average bin count ($\\bar{n}_h$). Color represents the NLDR layout. Layout d is universally poor. Layouts b and e that show two close clusters are universally suboptimal. Layout f with little separation performs well at tiny binwidth (where most points are in their own bin) and poorly as binwidth increases. Layout e has a small separation with oddly shaped clusters. Layout a is the best choice. Plot (ii), which compares HBE values with respect to average bin count, helps account for differences in cluster density; here, the variation among layouts is reduced, showing that some differences observed in (i) arise from density rather than true structure. Comparison of scaled evaluation metrics (rRTA, rSC, rGS, rARNX, and HBE using $a_1=0.05$) for the six NLDR layouts computed on the 2NC7 data using a parallel coordinate plot (iii). Color of the line indicates NLDR layout."
#| label: fig-toy-rmse
#| out-width: 100%
#| fig-height: 6
#| fig-width: 6

# Stack first two error plots vertically (left column)
left_column <- error_plot_two_curvy / error_plot_two_curvy_n_bar +
  plot_layout(heights = c(1, 1))

# Arrange six NLDR plots into two columns (right side)
right_columns <- wrap_plots(
  nldr_two_curvy1, nldr_two_curvy2, nldr_two_curvy3,
  nldr_two_curvy4, nldr_two_curvy5, nldr_two_curvy6,
  ncol = 2
)

# Combine left (2 plots) and right (6 plots) into one aligned block
top_section <- left_column | right_columns +
  plot_layout(widths = c(1.2, 1.2), guides = "collect") 

# Place comp_metrics below and match full width
final_plot_two_curvy <- (top_section) / comp_metrics +
  plot_layout(heights = c(3, 1)) &
  theme(plot.margin = margin(0, 0, 0, 0))

# Show the result
final_plot_two_curvy
```

```{r}
#| label: read-pbmc-nldr
# Read a variety of different NLDR representations of PBMC
# and plot them on same aspect ratio
umap_pbmc <- read_rds("data/pbmc3k/pbmc_umap_30_min_dist_0.3.rds")

nldr1c <- umap_pbmc |>
  ggplot(aes(x = UMAP1,
             y = UMAP2)) +
  geom_point(alpha=0.1, size=1, colour='#a65628') +
  interior_annotation("a", c(0.08, 0.9))

umap_pbmc <- read_rds("data/pbmc3k/pbmc_umap_5_min_dist_0.8.rds")

nldr2c <- umap_pbmc |>
  ggplot(aes(x = emb1,
             y = emb2))+
  geom_point(alpha=0.1, size=1, colour='#999999') +
  interior_annotation("b", c(0.08, 0.9))


umap_pbmc <- read_rds("data/pbmc3k/pbmc_umap_5_min_dist_0.01.rds")

nldr3c <- umap_pbmc |>
  ggplot(aes(x = UMAP1,
             y = UMAP2))+
  geom_point(alpha=0.1, size=1, colour='#e41a1c') +
  interior_annotation("c")

tsne_pbmc <- read_rds("data/pbmc3k/pbmc_tsne_5.rds")

nldr4c <- tsne_pbmc |>
  ggplot(aes(x = tSNE1,
             y = tSNE2))+
  geom_point(alpha=0.1, size=1, colour='#377eb8') +
  interior_annotation("d") 

tsne_pbmc <- read_rds("data/pbmc3k/pbmc_tsne_30.rds")

nldr5c <- tsne_pbmc |>
  ggplot(aes(x = tSNE1,
             y = tSNE2))+
  geom_point(alpha=0.1, size=1, colour='#4daf4a') +
  interior_annotation("e", c(0.08, 0.9)) 

phate_pbmc <- read_rds("data/pbmc3k/pbmc_phate_5.rds")

nldr6c <- phate_pbmc |>
  ggplot(aes(x = PHATE1,
             y = PHATE2))+
  geom_point(alpha=0.1, size=1, colour='#ff7f00') +
  interior_annotation("f") 

trimap_pbmc <- read_rds("data/pbmc3k/pbmc_trimap_12_4_3.rds")

nldr7c <- trimap_pbmc |>
  ggplot(aes(x = TriMAP1,
             y = TriMAP2))+
  geom_point(alpha=0.1, size=1, colour='#984ea3') +
  interior_annotation("g") 

pacmap_pbmc <- read_rds("data/pbmc3k/pbmc_pacmap_30_random_0.9_5.rds")

nldr8c <- pacmap_pbmc |>
  ggplot(aes(x = PaCMAP1,
             y = PaCMAP2))+
  geom_point(alpha=0.1, size=1, colour='#f781bf') +
  interior_annotation("h") 
```


<!--pbmc3k/08_gen_mse_for_diff_methods.R-->
```{r}
#| label: combine-data-pbmc

error_pbmc_umap <- read_rds("data/pbmc3k/error_pbmc_umap_30_min_dist_0.3.rds")
error_pbmc_umap2 <- read_rds("data/pbmc3k/error_pbmc_umap_5_min_dist_0.01.rds")
error_pbmc_umap3 <- read_rds("data/pbmc3k/error_pbmc_umap_5_min_dist_0.8.rds")
error_pbmc_tsne <- read_rds("data/pbmc3k/error_pbmc_tsne_5.rds")
error_pbmc_tsne2 <- read_rds("data/pbmc3k/error_pbmc_tsne_30.rds")
error_pbmc_phate <- read_rds("data/pbmc3k/error_pbmc_phate_5.rds")
error_pbmc_trimap <- read_rds("data/pbmc3k/error_pbmc_trimap_12_4_3.rds")
error_pbmc_pacmap <- read_rds("data/pbmc3k/error_pbmc_pacmap_30_random_0.9_5.rds")

error_pbmc <- bind_rows(error_pbmc_umap, 
                        error_pbmc_umap2,
                        error_pbmc_umap3,
                        error_pbmc_tsne,
                        error_pbmc_tsne2,
                        error_pbmc_phate,
                        error_pbmc_trimap,
                        error_pbmc_pacmap)


error_pbmc_avg <- error_pbmc |>
  filter(b1 >= 5) |>
  group_by(method, n_bar) |>
  filter(HBE == min(HBE)) |>
  ungroup() |>
  mutate(method = factor(method,
                         levels = c("UMAP_5_min_dist_0.8", "UMAP_30_min_dist_0.3", "UMAP_5_min_dist_0.01","tsne_5", "tsne_30", "phate_5", "trimap_12_4_3", "pacmap")))

error_pbmc <- error_pbmc |>
  mutate(a1 = round(a1, 2)) |>
  filter(b1 >= 5) |>
  group_by(method, a1) |>
  filter(HBE == min(HBE)) |>
  ungroup()

error_pbmc <- error_pbmc |>
  mutate(method = factor(method,
                         levels = c("UMAP_5_min_dist_0.8", "UMAP_30_min_dist_0.3", "UMAP_5_min_dist_0.01","tsne_5", "tsne_30", "phate_5", "trimap_12_4_3", "pacmap")))
```

```{r}
#| label: error-comp-pbmc

error_plot_pbmc <- plot_hbe(error_pbmc) +
  interior_annotation("i", position = c(0.07, 0.9)) +
  scale_x_continuous(breaks = seq(0, 0.25, 0.05)) +
                       #sort(unique(error_pbmc$a1))[c(1, 5, 9, 13, 17, 21, 26)]) +
  scale_color_manual(values=c('#999999','#a65628','#e41a1c','#377eb8','#4daf4a','#ff7f00','#984ea3','#f781bf')) +
  ylab("HBE")

error_plot_pbmc_n_bar <- plot_hbe_nbar(error_pbmc_avg) +
  annotate("text", x = 4, y = 6.8, label = "ii", size = 4, colour = "grey70") +
  scale_color_manual(values=c('#999999','#a65628','#e41a1c','#377eb8','#4daf4a','#ff7f00','#984ea3','#f781bf')) +
  ylab("HBE")

```

```{r}
#| fig-cap: "Assessing which of the 8 NLDR layouts on the PBMC3k data  (shown in @fig-NLDR-variety) is the better representation using HBE for varying (i) binwidth ($a_1$), and (ii) average bin count ($\\bar{n}_h$). Color used for the lines and points in the left plot and in the scatterplots represents NLDR layout (a-h). Layout f is universally poor. Layouts a, c, g, and h that show large separations between clusters are universally suboptimal. Layout d with little separation performs well at tiny binwidth (where most points are in their own bin) and poorly as binwidth increases. The choice of the best is between layouts b and e, which have small separations between oddly shaped clusters. Layout e is chosen as the best. Plot (ii), which accounts for the density within clusters by using average bin count, shows reduced differences between layouts, indicating that part of the variation in (i) is driven by cluster density rather than true structural differences."
#| label: fig-pbmc-rmse
#| out-width: 100%

# Stack error plots in one column
error_plots_pbmc <- error_plot_pbmc / error_plot_pbmc_n_bar

design <- "
AACC
AACC
BBCC
BBCC
"
error_plots_pbmc + 
  wrap_plots(nldr1c, nldr2c, nldr3c, nldr4c,
             nldr5c, nldr6c, nldr7c, nldr8c, ncol = 2) +
  plot_layout(design = design)
```

```{r}
nldr_metrics <- read_rds("data/pbmc3k/nldr_eval_metrics.rds")

nldr_metrics <- inner_join(nldr_metrics, error_pbmc, by = c("Method" = "method")) |>
  rename("sc" = "r") |>
  mutate(`rARNX` = 1 - `AUC(R_NX)`) |>
  mutate(`rRTA` = 1 - RTA) |>
  mutate(`rSC` = 1 - sc) |>
  mutate(`rGS` = 1 -  GS) |>
  mutate(Method = factor(Method,
                         levels = c("UMAP_5_min_dist_0.8", "UMAP_30_min_dist_0.3", "UMAP_5_min_dist_0.01","tsne_5", "tsne_30", "phate_5", "trimap_12_4_3", "pacmap")))

comp_metrics_pbmc <- plot_comp_metrics(nldr_metrics) + 
  interior_annotation("better", position = c(0.09, 0.03)) +
  interior_annotation("worse", position = c(0.09, 0.98)) +
  scale_color_manual(
    values=c("UMAP_5_min_dist_0.8" = '#999999', "UMAP_30_min_dist_0.3" = '#a65628', "UMAP_5_min_dist_0.01" = '#e41a1c',"tsne_5" = '#377eb8', "tsne_30" = '#4daf4a', "phate_5" = '#ff7f00', "trimap_12_4_3" = '#984ea3', "pacmap" ='#f781bf')) +
  theme(aspect.ratio = 2)
```

```{r}
#| label: fig-comp-metric-pbmc
#| fig-cap: "Comparison of scaled evaluation metrics (rRTA, rSC, rGS, rARNX, and HBE with $a_1 = 0.06$) for the eight NLDR layouts computed on the PBMC3k data, shown as a parallel coordinate plot. The color of each line corresponds to a NLDR layout. All, except rGS and rARNX agree that layout e is best or very close to best. Layout d is best according to HBE and rARNX but considered to be much less optimal by rRTA, rSC and rGS. Layout f is considered poor by rARNX and HBE. Layout a is cosidered close to the best by rGS and rSC."
#| fig-pos: "!ht"
#| out-width: 100%

design <- "
AABB
AABB
AABB
"
# comp_metrics_pbmc + 
#   wrap_plots(nldr1c, nldr2c, nldr3c, nldr4c,
#              nldr5c, nldr6c, nldr7c, nldr8c, ncol = 2) +
#   plot_layout(design = design)

# Stack first two error plots vertically (left column)
left_column <- error_plot_pbmc / error_plot_pbmc_n_bar +
  plot_layout(heights = c(1, 1))

# Arrange six NLDR plots into two columns (right side)
right_columns <- wrap_plots(nldr1c, nldr2c, nldr3c, nldr4c,
             nldr5c, nldr6c, nldr7c, nldr8c, ncol = 2)

# Combine left (2 plots) and right (6 plots) into one aligned block
top_section <- left_column | right_columns +
  plot_layout(widths = c(1.2, 1.2), guides = "collect") 

# Place comp_metrics below and match full width
final_plot_pbmc <- (top_section) / comp_metrics_pbmc +
  plot_layout(heights = c(3, 1)) &
  theme(plot.margin = margin(0, 0, 0, 0))

# Show the result
final_plot_pbmc
```


```{r}
#| label: read-mnist-nldr
# Read a variety of different NLDR representations of mnist
# and plot them on same aspect ratio
tsne_mnist <- read_rds("data/mnist/mnist_tsne30.rds")

nldr_mnist1 <- tsne_mnist |>
  ggplot(aes(x = emb1,
             y = emb2))+
  geom_point(alpha=0.1, size=1, colour='#984ea3') +
  interior_annotation("a", c(0.08, 0.9)) +
  theme(aspect.ratio = 1)

tsne_mnist2 <- read_rds("data/mnist/mnist_tsne89.rds")

nldr_mnist6 <- tsne_mnist2 |>
  ggplot(aes(x = emb1,
             y = emb2))+
  geom_point(alpha=0.1, size=1, colour='#999999') +
  interior_annotation("f", c(0.08, 0.9)) +
  theme(aspect.ratio = 1)

umap_mnist <- read_rds("data/mnist/mnist_umap.rds")

nldr_mnist2 <- umap_mnist |>
  ggplot(aes(x = emb1,
             y = emb2)) +
  geom_point(alpha=0.1, size=1, colour='#a65628') +
  interior_annotation("b") +
  theme(aspect.ratio = 1)

phate_mnist <- read_rds("data/mnist/mnist_phate.rds")

nldr_mnist3 <- phate_mnist |>
  ggplot(aes(x = emb1,
             y = emb2))+
  geom_point(alpha=0.1, size=1, colour='#377eb8') +
  interior_annotation("c", c(0.94, 0.92)) +
  theme(aspect.ratio = 1)

trimap_mnist <- read_rds("data/mnist/mnist_trimap.rds")

nldr_mnist4 <- trimap_mnist |>
  ggplot(aes(x = emb1,
             y = emb2))+
  geom_point(alpha=0.1, size=1, colour='#ff7f00') +
  interior_annotation("d", position = c(0.86, 0.92)) +
  theme(aspect.ratio = 1)

pacmap_mnist <- read_rds("data/mnist/mnist_pacmap.rds")

nldr_mnist5 <- pacmap_mnist |>
  ggplot(aes(x = emb1,
             y = emb2))+
  geom_point(alpha=0.1, size=1, colour='#e41a1c') +
  interior_annotation("e", c(0.08, 0.9)) +
  theme(aspect.ratio = 1)
```

<!--mnist/03_gen_mse_for_diff_methods.R-->
```{r}
#| label: combine-data-mnist

error_mnist_umap <- read_rds("data/mnist/error_mnist_umap.rds")
error_mnist_tsne <- read_rds("data/mnist/error_mnist_tsne.rds")
error_mnist_tsne2 <- read_rds("data/mnist/error_mnist_tsne2.rds")
error_mnist_phate <- read_rds("data/mnist/error_mnist_phate.rds")
error_mnist_trimap <- read_rds("data/mnist/error_mnist_trimap.rds")
error_mnist_pacmap <- read_rds("data/mnist/error_mnist_pacmap.rds")

error_mnist <- bind_rows(error_mnist_umap, 
                        error_mnist_tsne,
                        error_mnist_tsne2,
                        error_mnist_phate,
                        error_mnist_trimap,
                        error_mnist_pacmap)

error_mnist_avg <- error_mnist |>
  filter(b1 >= 5) |>
  group_by(method, n_bar) |>
  filter(HBE == min(HBE)) |>
  ungroup() |>
  mutate(method = factor(method,
                         levels = c("UMAP", "tSNE", "tSNE2", "PHATE", "TriMAP", "PaCMAP")))

error_mnist <- error_mnist |>
  mutate(a1 = round(a1, 2)) |>
  filter(b1 >= 5) |>
  group_by(method, a1) |>
  filter(HBE == min(HBE)) |>
  ungroup() |>
  mutate(method = factor(method,
                         levels = c("UMAP", "tSNE", "tSNE2", "PHATE", "TriMAP", "PaCMAP")))
```

```{r}
#| label: error-comp-mnist

error_plot_mnist <- plot_hbe(error_mnist) +
  interior_annotation("i", position = c(0.07, 0.9)) +
  scale_x_continuous(breaks = seq(0, 0.25, 0.05)) +
                       #sort(
    #unique(error_mnist$a1))[append(seq(1, length(unique(error_mnist$a1)), 
 #                                      by = 5), 24)]) +
  scale_color_manual(
    values=c('#a65628', '#984ea3', '#999999', 
             '#377eb8', '#ff7f00', '#e41a1c')) +
  theme(aspect.ratio=0.685) +
  ylab("HBE")

error_plot_mnist_n_bar <- plot_hbe_nbar(error_mnist_avg) +
  annotate("text", x = 4, y = 2.1, label = "ii", size = 4, colour = "grey70") +
  scale_color_manual(
    values=c('#a65628', '#984ea3', '#999999', 
             '#377eb8', '#ff7f00', '#e41a1c')) +
  theme(aspect.ratio=0.685) +
  ylab("HBE")

```

```{r}
nldr_metrics <- read_rds("data/mnist/nldr_eval_metrics.rds")

nldr_metrics <- inner_join(nldr_metrics, error_mnist, by = c("Method" = "method")) |>
  rename("sc" = "r") |>
  mutate(`rARNX` = 1 - `AUC(R_NX)`) |>
  mutate(`rRTA` = 1 - RTA) |>
  mutate(`rSC` = 1 - sc) |>
  mutate(`rGS` = 1 -  GS) |>
  mutate(Method = factor(Method,
                         levels = c("UMAP", "tSNE", "tSNE2", "PHATE", "TriMAP", "PaCMAP")))

comp_metrics_mnist <- plot_comp_metrics(nldr_metrics) +
  interior_annotation("better", position = c(0.09, 0.03)) +
  interior_annotation("worse", position = c(0.09, 0.98)) +
  scale_color_manual(
    values=c("UMAP" = '#a65628', "tSNE" = '#984ea3', "tSNE2" = '#999999', 
             "PHATE" = '#377eb8', "TriMAP" = '#ff7f00', "PaCMAP" = '#e41a1c')) +
  theme(aspect.ratio = 1.5)
```

```{r}
#| label: fig-comp-metric-mnist
#| fig-cap: "Comparison of scaled evaluation metrics (rARNX, rRTA, rSC, rGS, and HBE using $a_1=0.04$) for six NLDR layouts computed on the MNIST digit 1 data using a parallel coordinate plot. Each line represents a layout (a–f), with colors corresponding to the scatterplots shown on the right. The metrics display different ranking patterns, indicating that no single measure fully captures embedding quality. Layout a is identified as the best according to HBE and rRTA, but is considered much less optimal by rARNX, rSC, and rGS. Layout e is considered the worst, or close to the poorest, by all metrics. Layouts a and f show similar patterns of agreement across metrics, except for rRTA, where layout a performs the best and layout f the worst. Layout c is the worst in rARNX, rSC, and HBE."
#| fig-pos: "!ht"
#| out-width: 100%
#| fig-height: 6
#| fig-width: 6

# comp_metrics_mnist + wrap_plots(nldr_mnist1, nldr_mnist2, nldr_mnist3, 
#                                     nldr_mnist4, nldr_mnist5, nldr_mnist6, ncol = 2)

# Stack first two error plots vertically (left column)
left_column <- error_plot_mnist / error_plot_mnist_n_bar +
  plot_layout(heights = c(1, 1))

# Arrange six NLDR plots into two columns (right side)
right_columns <- wrap_plots(nldr_mnist1, nldr_mnist2, nldr_mnist3, 
                                    nldr_mnist4, nldr_mnist5, nldr_mnist6, ncol = 2)

# Combine left (2 plots) and right (6 plots) into one aligned block
top_section <- left_column | right_columns +
  plot_layout(widths = c(1.2, 1.2), guides = "collect") 

# Place comp_metrics below and match full width
final_plot_mnist <- (top_section) / comp_metrics_mnist +
  plot_layout(heights = c(3, 1)) &
  theme(plot.margin = margin(0, 0, 0, 0))

# Show the result
final_plot_mnist
```

\clearpage
