---
title: "Choosing Better NLDR Layouts by Evaluating the Model in the High-dimensional Data Space"
format: 
    jasa-pdf:
        keep-tex: true
author:
  - name: Jayani P. Gamage
    affiliations:
      - name: Monash University
        department: Econometrics & Business Statistics
        address: Clayton
        city: VIC
        country: Australia
        postal-code: 3800
    orcid: 0000-0002-6265-6481
    email: jayani.piyadigamage@monash.edu
    url: https://jayanilakshika.netlify.app/
  - name: Dianne Cook
    affiliations:
      - name: Monash University
        department: Econometrics & Business Statistics
        address: Clayton
        city: VIC
        country: Australia
        postal-code: 3800
    orcid: 0000-0002-3813-7155
    email: dicook@monash.edu
    url: http://www.dicook.org/
  - name: Paul Harrison
    affiliations:
      - name: Monash University
        department: MGBP, BDInstitute
        address: Clayton
        city: VIC
        country: Australia
        postal-code: 3800
    orcid: 0000-0002-3980-268X
    email: 	paul.harrison@monash.edu
    url:
  - name: Michael Lydeamore
    affiliations:
      - name: Monash University
        department: Econometrics & Business Statistics
        address: Clayton
        city: VIC
        country: Australia
        postal-code: 3800
    orcid: 0000-0001-6515-827X
    email: michael.lydeamore@monash.edu
    url: https://www.michaellydeamore.com/
  - name: Thiyanga S. Talagala
    affiliations:
      - name: University of Sri Jayewardenepura
        department: Statistics
        address: Gangodawila
        city: Nugegoda
        country: Sri Lanka
        postal-code: 10100
    orcid: 0000-0002-0656-9789
    email: ttalagala@sjp.ac.lk
    url: https://thiyanga.netlify.app/
tbl-cap-location: bottom
abstract: |
  Nonlinear dimension reduction (NLDR) techniques such as t-SNE and UMAP provide low-dimensional representations of high-dimensional data, capturing complex structures in a concise visual summary. While NLDR has an important role in data analysis, these methods can badly miss the mark by exaggerating existing structures or suggesting spurious patterns. NLDR applies nonlinear transformations to create layouts, and the choice of method and hyperparameters can produce wildly different representations. It is difficult to decide which are reasonable and which are misleading, and we show they can be poor. To help, we have developed a procedure that treats the low-dimensional NLDR layout as a model that can be lifted back into the original high-dimensional space. This enables us to assess goodness of fit by computing the error and visualize the fitted model overlaid on the data using grand tours. These elements allow us to quantitatively compare multiple layouts, determine which representations are acceptable, and examine whether an embedding captures the data structure throughout the entire space, only in certain subspaces, or fits poorly overall. This approach has also proven useful for understanding how different methodological choices affect the resulting embeddings. 
  
keywords: [high-dimensional data, dimension reduction, data visualization, statistical graphics, tour, unsupervised learning]
keywords-formatted: [high-dimensional data, dimension reduction, data visualization, statistical graphics, unsupervised learning]

bibliography: paper.bib  
header-includes: | 
  \usepackage{amsmath}
  \usepackage{float}
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \usepackage{bm}
  \def\tightlist{}
  \usepackage{setspace}
  \newcommand\pD{$p\text{-}D$}
  \newcommand\kD{$k\text{-}D$}
  \newcommand\dD{$d\text{-}D$}
  \newcommand\gD{$2\text{-}D$}
  \newcommand\tD{$3\text{-}D$}
  \newcommand\fD{$4\text{-}D$}
  \newcommand\sD{$7\text{-}D$}
---

<!-- Looking at nonlinear Dimension Reductions as Models in the Data Space 

Quantitative Visual Tools to Honestly Select Nonlinear Dimension Reductions

Stop Lying to Me: Visual Evidence for Honest Dimension Reduction

When Looks Deceive: Quantifying the Honesty of Nonlinear Layouts

Not Just Pretty Pictures: Quantitative Visual Evaluation of Nonlinear DR Methods

Beauty Isn’t Truth: Choosing Nonlinear Embeddings that Stay True to the Data

Truth in the Tangle: Choosing Nonlinear Reductions with Quantitative Visual Evidence

Beyond Subjectivity: New Visual Tools to Choose a Nonlinear Dimension Reduction

Choosing better NLDR layouts by evaluating the model in the high dimensional data space
-->

```{r include=FALSE}
# Set up chunk for knitr
knitr::opts_chunk$set(
  fig.width = 5,
  fig.height = 5,
  fig.align = "center",
  out.width = "100%",
  code.line.numbers = FALSE,
  fig.retina = 4,
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  cache = FALSE,
  dev.args = list(pointsize = 11)
)
```

<!--This paper is build with quollr 0.3.15-->

```{r include=FALSE}
#| label: install-libraries
#| warning: false
#| echo: false

options(repos = c(CRAN = "https://cran.rstudio.com")) # Setup mirror

packages_to_check <- c("remotes", "tidyverse", "patchwork", "colorspace", "kableExtra", "conflicted")

for (pkg in packages_to_check) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    message(paste("Installing package:", pkg))
    install.packages(pkg)
  } else {
    installed_version <- packageVersion(pkg)
    available_version <- tryCatch({
      utils::packageDescription(pkg)$Version
    }, error = function(e) NA) # Handle cases where package info isn't readily available

    if (!is.na(available_version) && installed_version < package_version(available_version)) {
      message(paste("A newer version of package", pkg, "is available. Updating..."))
      install.packages(pkg)
    } else {
      message(paste("Package", pkg, "is up-to-date (version", installed_version, ")."))
    }
  }
}

```

```{r}
#| label: load-libraries
#| warning: false
#| echo: false
library(tibble)
library(dplyr)
library(ggplot2)
library(readr)
library(tidyr)
library(patchwork)
library(colorspace)
library(kableExtra)
library(conflicted)
library(ggbeeswarm)
library(scales)
library(GGally)
library(quollr)

conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::select)
```

```{r}
#| label: plot-theme
theme_set(theme_linedraw() +
   theme(
     #aspect.ratio = 1,
     plot.background = element_rect(fill = 'transparent', colour = NA),
     plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
     panel.background = element_rect(fill = 'transparent', 
                                     colour = NA),
     panel.grid.major = element_blank(), 
     panel.grid.minor = element_blank(), 
     axis.title.x = element_blank(), axis.title.y = element_blank(),
     axis.text.x = element_blank(), axis.ticks.x = element_blank(),
     axis.text.y = element_blank(), axis.ticks.y = element_blank(),
     legend.background = element_rect(fill = 'transparent', 
                                      colour = NA),
     legend.key = element_rect(fill = 'transparent', 
                               colour = NA),
     legend.position = "bottom", 
     legend.title = element_blank(), 
     legend.text = element_text(size=4),
     legend.key.height = unit(0.25, 'cm'),
     legend.key.width = unit(0.25, 'cm'),
     plot.margin = margin(0, 0, 0, 0)
   )
)
```

```{r}
#| label: import-scripts
source("scripts/additional_functions.R")
```

```{r}
#| label: code-setup
set.seed(20240110)
```

<!-- 
Check-list before submission
* Is it all American spelling
* Spelling checked generally
* Code all runs given fresh workspace
* Code has a readme, explaining how the paper results are reproduced
* Re-write abstract
-->

\spacingset{2.0} <!--% command in JASA style, comment to go back to double spacing-->

## Introduction

Nonlinear dimension reduction (NLDR) is popular for making a convenient low-dimensional (\kD{}) representation of high-dimensional (\pD{}) data ($k < p$). Recently developed methods include t-distributed stochastic neighbor embedding (tSNE) [@laurens2008], uniform manifold approximation and projection (UMAP) [@leland2018], potential of heat-diffusion for affinity-based trajectory embedding (PHATE) algorithm [@moon2019], large-scale dimensionality reduction using triplets (TriMAP) [@amid2022], and pairwise controlled manifold approximation (PaCMAP) [@yingfan2021]. 

However, the representation generated can vary dramatically from method to method, choice of hyper-parameter, or even random seed, as illustrated by @fig-NLDR-variety. <!--The specific method and hyper-parameters used to produce each layout (see Supplementary Materials) are not essential for the discussion. The dilemma for the analyst is which representation to use.--> The choice may result in different procedures being used in the downstream analysis or different inferential conclusions. Various academics have expressed concerns about current practices and procedures for choosing (e.g., @irizarry2024, @chari2023). The research described here provides new numerical and visual tools to help choose a suitable layout. 

```{r}
#| label: read-pbmc-nldr
# Read a variety of different NLDR representations of PBMC
# and plot them on same aspect ratio
clr_choice <- "#0077A3"
umap_pbmc <- read_rds("data/pbmc3k/pbmc_umap_30_min_dist_0.3.rds")

nldr1 <- umap_pbmc |>
  ggplot(aes(x = UMAP1,
             y = UMAP2)) +
  geom_point(alpha=0.1, size=1, colour=clr_choice) +
  interior_annotation("a", c(0.08, 0.9), cex = 1.3)

nldr1c <- umap_pbmc |>
  ggplot(aes(x = UMAP1,
             y = UMAP2)) +
  geom_point(alpha=0.1, size=1, colour='#a65628') +
  interior_annotation("a", c(0.08, 0.9))

# umap_pbmc <- read_rds("data/pbmc3k/pbmc_umap_12_min_dist_0.99.rds")
umap_pbmc <- read_rds("data/pbmc3k/pbmc_umap_5_min_dist_0.8.rds")
nldr2 <- umap_pbmc |>
  ggplot(aes(x = emb1,
             y = emb2))+
  geom_point(alpha=0.1, size=1, colour=clr_choice) +
  interior_annotation("b", c(0.08, 0.9), cex = 1.3)

nldr2c <- umap_pbmc |>
  ggplot(aes(x = emb1,
             y = emb2))+
  geom_point(alpha=0.1, size=1, colour='#999999') +
  interior_annotation("b", c(0.08, 0.9))


umap_pbmc <- read_rds("data/pbmc3k/pbmc_umap_5_min_dist_0.01.rds")

nldr3 <- umap_pbmc |>
  ggplot(aes(x = UMAP1,
             y = UMAP2))+
  geom_point(alpha=0.1, size=1, colour=clr_choice) +
  interior_annotation("c", cex = 1.3)

nldr3c <- umap_pbmc |>
  ggplot(aes(x = UMAP1,
             y = UMAP2))+
  geom_point(alpha=0.1, size=1, colour='#e41a1c') +
  interior_annotation("c")

tsne_pbmc <- read_rds("data/pbmc3k/pbmc_tsne_5.rds")

nldr4 <- tsne_pbmc |>
  ggplot(aes(x = tSNE1,
             y = tSNE2))+
  geom_point(alpha=0.1, size=1, colour=clr_choice) +
  interior_annotation("d", cex = 1.3)

nldr4c <- tsne_pbmc |>
  ggplot(aes(x = tSNE1,
             y = tSNE2))+
  geom_point(alpha=0.1, size=1, colour='#377eb8') +
  interior_annotation("d") 

tsne_pbmc <- read_rds("data/pbmc3k/pbmc_tsne_30.rds")

nldr5 <- tsne_pbmc |>
  ggplot(aes(x = tSNE1,
             y = tSNE2))+
  geom_point(alpha=0.1, size=1, colour=clr_choice) +
  interior_annotation("e", c(0.08, 0.9), cex = 1.3)

nldr5c <- tsne_pbmc |>
  ggplot(aes(x = tSNE1,
             y = tSNE2))+
  geom_point(alpha=0.1, size=1, colour='#4daf4a') +
  interior_annotation("e", c(0.08, 0.9)) 

phate_pbmc <- read_rds("data/pbmc3k/pbmc_phate_5.rds")
nldr6 <- phate_pbmc |>
  ggplot(aes(x = PHATE1,
             y = PHATE2))+
  geom_point(alpha=0.1, size=1, colour=clr_choice) +
  interior_annotation("f", cex = 1.3)

nldr6c <- phate_pbmc |>
  ggplot(aes(x = PHATE1,
             y = PHATE2))+
  geom_point(alpha=0.1, size=1, colour='#ff7f00') +
  interior_annotation("f") 

trimap_pbmc <- read_rds("data/pbmc3k/pbmc_trimap_12_4_3.rds")
nldr7 <- trimap_pbmc |>
  ggplot(aes(x = TriMAP1,
             y = TriMAP2))+
  geom_point(alpha=0.1, size=1, colour=clr_choice) +
  interior_annotation("g", cex = 1.3)

nldr7c <- trimap_pbmc |>
  ggplot(aes(x = TriMAP1,
             y = TriMAP2))+
  geom_point(alpha=0.1, size=1, colour='#984ea3') +
  interior_annotation("g") 

pacmap_pbmc <- read_rds("data/pbmc3k/pbmc_pacmap_30_random_0.9_5.rds")
nldr8 <- pacmap_pbmc |>
  ggplot(aes(x = PaCMAP1,
             y = PaCMAP2))+
  geom_point(alpha=0.1, size=1, colour=clr_choice) +
  interior_annotation("h", cex = 1.3)

nldr8c <- pacmap_pbmc |>
  ggplot(aes(x = PaCMAP1,
             y = PaCMAP2))+
  geom_point(alpha=0.1, size=1, colour='#f781bf') +
  interior_annotation("h") 
```


```{r}
#| label: fig-NLDR-variety
#| echo: false
#| fig-cap: "Eight different NLDR representations of the same data, produced by tSNE, UMAP, PHATE, TriMAP and PaCMAP with a variety of hyper-parameter choices. The variety in layouts makes it difficult to choose which best represents the data distribution."
#| fig-width: 8
#| fig-height: 4
#| out-width: 80%
# (a) UMAP (n_neighbors = 30, min_dist = 0.3), (b) UMAP (n_neighbors = 5, min_dist = 0.01), (c) UMAP (n_neighbors = 15, min_dist = 0.99), (f) tSNE (perplexity = 5), (g) tSNE (perplexity = 30), (l) TriMAP (n_inliers = 12, n_outliers = 4, n_random = 3), (q) PaCMAP (n_neighbors = 30, init = random, MN_ratio = 0.9, FP_ratio = 5)
nldr1 + nldr2 + nldr3 + nldr4 +
  nldr5 + nldr6 + nldr7 + nldr8 +
  plot_layout(ncol = 4)
```

The paper is organized as follows. @sec-background gives more details behind NLDR and high-dimensional data visualization. <!--@sec-method--> Sections [-@sec-NLDRmodel], [-@sec-algorithm], [-@sec-highd], [-@sec-summary] describes the new methodology, illustrated using a simulated data example. @sec-prediction extends the approach for prediction of new observations. @sec-interactive illustrates interactive graphics to diagnose an NLDR fit. A workflow for choosing from a selection of layouts is provided in @sec-bestfit. Two applications <!--illustrating the use of the new methodology for bioinformatics and image classification--> are in @sec-applications. @sec-discussion discusses limitations and future directions. In addition, there is an R package `quollr` which implements the methods.

## Background literature on dimension reduction {#sec-background}

<!-- - Connection between NLDR and MDS-->
Making useful low-dimensional (\kD{}) layouts of high-dimensional (\pD{}) data has historically been achieved using multidimensional scaling (MDS) [@kruskal1964] ].  In MDS, the \kD{} layout is constructed by minimizing a stress function that compares distances between points in \pD{} with potential distances between points in \kD{}. Various formulations of the stress function result in non-metric scaling [@saeed2018] and isomap [@silva2002]. Principal components analysis (PCA) produces the same layout as classical MDS, if interpoint distances are Euclidean. Recent overviews of MDS, PCA and  statistically conducting linear and nonlinear dimension reduction can be found in @borg2005, @johnstone2009, @jolliffe2011, @Koch_2013. 

The relatively recent NLDR methods, tSNE, UMAP, PHATE, TriMAP, and PaCMAP, can be considered new approaches tackling the same problem as MDS. The methods tSNE, UMAP, TriMAP, and PaCMAP are similar in that they determine the \kD{} representation by minimizing the divergence between two inter-point distance distributions. PHATE is different, and is based on a diffusion process spreading to capture geometric shapes that include both global and local structure. (See @coifman2005 for more details on diffusion processes.)  

The array of layouts in @fig-NLDR-variety illustrate what can result by choosing different methods, tuning with different hyper-parameters, and even the random seed that initiates the computation. Each layout provides a different lens on the interpoint distances in the data. The key structures in the high-dimensional data that we might hypothesize from these \gD{} layouts are: (1) highly *separated clusters* (a, b, e, g, h) with the number of clusters ranging from 3-6; (2) *stringy branches* (f), and (3) *barely separated clusters* (c, d). They are at odds with each other, and it begs us to decide which are actually present in the \pD{} data. 

The alternative approach to visualizing the high-dimensional data is to use linear projections, such as PCA. Tours, defined by @As85, broaden the scope by providing movies of linear projections showing views of the data from all directions. @lee2021 has a review of tour methods. Many tour algorithms are available in the R package `tourr` [@wickham2011], and versions enabling better interactivity in `langevitour` [@harisson2024] and `detourr` [@hart2022]. Linear projections are a safe way to view high-dimensional data because they do not warp the space, so they are more faithful representations of the structure. However, linear projections can be cluttered, and global patterns can obscure local structure. The simple activity of projecting data from \pD{} suffers from piling [@laa2022], where data concentrates in the center of projections. NLDR is designed to escape these issues, to exaggerate structure so that it can be observed. But as a result, NLDR can hallucinate wildly, to suggest patterns that are not actually present in the data. 

@Lee_Laa_Cook_2022 provided interactive graphics to link between the NLDR view and tours, to help understand how the NLDR transformed the interpoint distances into a \pD{} layout. Here, we further develop this approach to directly use the tour to examine how the NLDR is warping the space. It follows what @wickham2015 describes as **model-in-the-data-space**. The fitted model should be overlaid on the data to examine the fit relative to the spread of the observations. While this is straightforward, and commonly done when data is \gD{}, it is also possible in \pD{}, for many models, when a tour is used. 

@wickham2015 provides several examples of models overlaid on the data in \pD{}. In hierarchical clustering, a representation of the dendrogram using points and lines can be constructed by augmenting the data with points marking the merging of clusters. Showing the movie of linear projections shows how the algorithm sequentially fitted the cluster model to the data. For linear discriminant analysis or model-based clustering, the model can be indicated by $(p-1)\text{-}D$ ellipses. It is possible to see whether the elliptical shapes appropriately match the variance of the relevant clusters, and to compare and contrast different fits. For PCA, one can display the model (a \kD{} plane of the reduced dimension) using wireframes of transformed cubes. Using a wireframe is the approach we take here to represent the NLDR model in \pD{}.

<!--## Methodology {#sec-method}-->

## Considering the NLDR layout as a model {#sec-NLDRmodel}

At first glance, thinking of NLDR as a modeling technique might seem strange. It is a simplified representation or abstraction of a system, process, or phenomenon in the real world. The \pD{}   observations are the realization of the phenomenon, and the \kD{} NLDR layout is the simplified representation. Typically, $k=2$ is used. From a statistical perspective, we can consider the distances between points in the \gD{}   layout to be variance that the model explains, and the (relative) difference with their distances in \pD{}   as the error, or unexplained variance. We can also imagine that the positioning of points in \gD{}    represent the fitted values, which will have some prescribed position in \pD{}   that can be compared with their observed values. This is the conceptual framework underlying the more formal versions of factor analysis [@joreskog1969] and MDS. (Note that, for this thinking, the full \pD{}   data needs to be available, not just the interpoint distances.)

We define the NLDR as a function $\mathbfit{Y}_{n\times 2} = g(\mathbfit{X_{n\times p}})$, $g\text{:}~ \mathbb{R}^{p} \rightarrow  \in \mathbb{R}^{2}$, with hyper-parameters $\mathbfit{\theta}$, provided not estimated. These parameters, $\mathbfit{\theta}$, depend on the choice of $g$, and can be considered part of model fitting in the traditional sense. Common choices for $g$ include functions used in tSNE, UMAP, PHATE, TriMAP, PaCMAP, or MDS, although in theory any function that does this mapping is suitable. XXX add something on fitting

With our goal being to make a representation of this \gD{} layout that can be lifted into high-dimensional space, the layout needs to be augmented to include neighbor information. A simple approach would be to triangulate the points and add edges. A more stable approach is to first bin the data, reducing it from $n$ to $m\leq n$ observations, and connect the bin centroids. We recommend using a hexagon grid because it better reflects the data distribution and has fewer artifacts than a rectangular grid [@carr1987]. This process serves to reduce some noisiness in the resulting surface shown in \pD{}. The steps in this process are shown in @fig-NLDR-two-curvy, and documented below.

<!--two_nonlinear/06_gen_model_with_tSNE.R-->
```{r}
#| label: fit_model-two-curvy

tsne_two_curvy_scaled <- read_rds("data/two_nonlinear/tsne_two_curvy_scaled.rds")
hex_grid_with_counts <- read_rds("data/two_nonlinear/hex_grid_with_counts.rds")
hex_grid_nonempty <- read_rds("data/two_nonlinear/hex_grid_nonempty.rds")
tr_from_to_df_two_curvy <- read_rds("data/two_nonlinear/tr_from_to_df_two_curvy.rds")
df_bin_centroids_two_curvy <- read_rds("data/two_nonlinear/df_bin_centroids_two_curvy.rds")

sc_ltr_pos <- c(0.08, 0.9)
sc_xlims <- c(-0.3, 1.2)
sc_ylims <- c(-0.17, 1.25)
```

<!--Full hexagon grid with UMAP data-->

```{r}
#| label: hexbin-two-curvy

nldr_two_curvy_original <- ggplot(tsne_two_curvy_scaled, 
                                  aes(x = emb1, y = emb2)) +
  geom_point(alpha = 0.5, colour = clr_choice, size = 0.5) 

# First plot with xlim and ylim
nldr_two_curvy <- nldr_two_curvy_original +
  interior_annotation("a", sc_ltr_pos, cex = 2) +
  xlim(sc_xlims) +
  ylim(sc_ylims) +
  annotate("text", x=c(0.1, 0.9), y=c(0, 0.1), 
           label = c("cluster 2", "cluster 1"), 
           colour=clr_choice, size=5)

hex_grid_two_curvy <- ggplot(
  data = hex_grid_with_counts, 
  aes(x = x, y = y)) +
  geom_polygon(colour = "grey70", 
               aes(group = h), 
               fill = "#ffffff") +
  geom_point(data = tsne_two_curvy_scaled, 
             aes(x = emb1, y = emb2), 
             alpha = 0.5, size = 0.5, colour = clr_choice) +
  xlim(sc_xlims) + ylim(sc_ylims) +
  interior_annotation("b", sc_ltr_pos, cex = 2)

```

<!--Non-empty bins with bin centroids-->

```{r}
#| label: empty-bin-two-curvy

hex_grid_nonempty_two_curvy <- ggplot(
  data = hex_grid_nonempty, 
  aes(x = x, y = y)) +
  geom_polygon(colour = "grey70", 
               aes(group = h), 
               fill = "#ffffff") +
  geom_point(data = df_bin_centroids_two_curvy, 
             aes(x = c_x, y = c_y), 
             colour = "#000000",
             size = 1) +
  xlim(sc_xlims) + ylim(sc_ylims) +
  interior_annotation("c", sc_ltr_pos, cex = 2) 
```

<!--2D model-->
```{r}
#| label: triangulate-two-curvy

trimesh_two_curvy <- ggplot() +
  geom_segment(data = tr_from_to_df_two_curvy,
               aes(
                 x = x_from,
                 y = y_from,
                 xend = x_to,
                 yend = y_to),
               colour = "#000000") +
  geom_point(data = tsne_two_curvy_scaled,
             aes(
               x = emb1,
               y = emb2
             ),
             colour = "#636363",
             alpha = 0.5,
             size = 0.5
  ) +
  coord_equal() +
  interior_annotation("a1", sc_ltr_pos)

trimesh_removed_two_curvy <- ggplot() +
  geom_segment(data = tr_from_to_df_two_curvy,
               aes(
                 x = x_from,
                 y = y_from,
                 xend = x_to,
                 yend = y_to),
               colour = "#000000") +
  coord_equal() +
  xlim(sc_xlims) + ylim(sc_ylims) +
  interior_annotation("d", sc_ltr_pos, cex = 2)

trimesh_two_curvy_removed1_tsne_with_data <- ggplot() +
  geom_point(data = tsne_two_curvy_scaled,
             aes(
               x = emb1,
               y = emb2
             ),
             colour = clr_choice,
             alpha = 0.5,
             size = 0.5) +
  geom_segment(data = tr_from_to_df_two_curvy,
               aes(
                 x = x_from,
                 y = y_from,
                 xend = x_to,
                 yend = y_to),
               colour = "#000000") +
  interior_annotation("a", cex = 1.2) +
  theme(aspect.ratio = 1)
```


```{r}
#| label: compute-error-tsne

## Compute error
error_df_two_curvy_abs <- read_rds("data/two_nonlinear/error_df_two_curvy_abs.rds")

error_plot_two_curvy <- error_df_two_curvy_abs |>
  ggplot(aes(x = emb1,
             y = emb2,
             colour = error_cat)) +
  geom_point(alpha=0.5) +
  interior_annotation(label = "b", cex = 1.5) +
  scale_colour_manual(values = c("#FFFFC8", "#FFFAC0", "#FEF2B3", "#FCE9A3",
                                 "#FADE8F", "#F9D378", "#F8C65D", "#F6B938",
                                 "#F5AA00", "#F49B00", "#F28A00", "#F07800",
                                 "#ED6400", "#EA4D00", "#DE3900", "#CD2A00",
                                 "#BB1B00", "#A70C00", "#93001C", "#7D0025")) +
  xlab("tSNE1") +
  ylab("tSNE2") +
  annotate("text", x=c(0.2, 0.8), y=c(0, 0.15), 
           label = c("cluster 2", "cluster 1"), 
           colour=clr_choice, size=3) +
  theme_bw() +
  theme(
    aspect.ratio = 1,
    legend.position = "none",
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank()
  )

error_plot_two_curvy_quasi <- ggplot(
  error_df_two_curvy_abs, 
  aes(x = sqrt_row_wise_total_error, 
      y = 0, 
      colour = error_cat)) +
  geom_quasirandom(alpha = 0.5) +
  interior_annotation(label = "a", cex = 1.5) +
  scale_colour_manual(values = c("#FFFFC8", "#FFFAC0", "#FEF2B3", "#FCE9A3",
                                 "#FADE8F", "#F9D378", "#F8C65D", "#F6B938",
                                 "#F5AA00", "#F49B00", "#F28A00", "#F07800",
                                 "#ED6400", "#EA4D00", "#DE3900", "#CD2A00",
                                 "#BB1B00", "#A70C00", "#93001C", "#7D0025")) +
  xlab("residuals") +
  ylab("") +
  theme_bw() +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        panel.grid.major.y = element_blank(),
        panel.grid.minor.y = element_blank(),
        legend.position = "none",
        plot.margin = margin(0, 0, 0, 0))
```

```{r}
#| label: best-fit-two-curvy-proj-tsne

## First projection
proj_obj1 <- read_rds("data/two_nonlinear/two_nonlinear_proj_obj1.rds")

two_curvy_proj_tsne_first_model1 <- plot_proj(
    proj_obj = proj_obj1, 
    point_param = c(0.5, 0.2, clr_choice), # size, alpha, color
    line_param = c(0.5, 0.5, "#000000"), # linewidth, alpha
    plot_limits = c(-0.75, 0.75), 
    axis_text_size = 2, 
    is_category = FALSE) +
  interior_annotation(label = "b", cex = 1.2) +
  annotate("text", x=c(0, -0.2), y=c(0.55, 0), 
        label = c("cluster 1", "cluster 2"), 
        colour=clr_choice, size=3) +
  theme(aspect.ratio = 1) 


projected_df <- proj_obj1$projected_df
axes <- proj_obj1$axes
circle <- proj_obj1$circle
cluster <- proj_obj1$cluster

projected_df <- projected_df |>
      dplyr::mutate(cluster = proj_obj1$cluster)

two_curvy_proj_tsne_first_model1_coloured <- projected_df |>
  ggplot(
    aes(
      x = proj1,
      y = proj2,
      colour = cluster)) +
  geom_point(
    size = 0.5,
    alpha = 0.2) +
  geom_segment(
    data=axes,
    aes(x=x1, y=y1, xend=x2, yend=y2),
    colour="grey70") +
  geom_text(
    data=axes,
    aes(x=x2, y=y2),
    label=rownames(axes),
    colour="grey50",
    size = 2) +
  geom_path(
    data=circle,
    aes(x=c1, y=c2), colour="grey70") +
  scale_colour_manual(values = c("#FFFFC8", "#FFFAC0", "#FEF2B3", "#FCE9A3",
                                 "#FADE8F", "#F9D378", "#F8C65D", "#F6B938",
                                 "#F5AA00", "#F49B00", "#F28A00", "#F07800",
                                 "#ED6400", "#EA4D00", "#DE3900", "#CD2A00",
                                 "#BB1B00", "#A70C00", "#93001C", "#7D0025")) +
  xlim(c(-0.75, 0.75)) +
  ylim(c(-0.75, 0.75)) +
  interior_annotation(label = "c", cex = 1.5) +
  annotate("text", x=c(0, -0.2), y=c(0.55, 0), 
        label = c("cluster 1", "cluster 2"), 
        colour=clr_choice, size=3) +
  theme(aspect.ratio = 1,
        legend.position = "none") 

## Second projection

proj_obj2 <- read_rds("data/two_nonlinear/two_nonlinear_proj_obj2.rds")

two_curvy_proj_tsne_all_model1 <- plot_proj(
  proj_obj = proj_obj2, 
  point_param = c(0.1, 0.5, clr_choice), # size, alpha, color
  line_param = c(0.5, 0.5, "black"), # linewidth, alpha
  plot_limits = c(-0.5, 0.5), 
  axis_text_size = 2, 
  is_category = FALSE) +
  interior_annotation(label = "c", cex = 1.2) +
  annotate("text", x=c(-0.25, 0.2), y=c(0.25, 0.25), 
        label = c("cluster 1", "cluster 2"), 
        colour=clr_choice, size=3) +
  theme(aspect.ratio = 1)

```

To illustrate the method and how to use it to choose from a selection of layouts, we simulated \sD{} data, which we call the "2NC7" data, that has two separated nonlinear clusters, one forming a \gD{} curved shape (cluster 1), and the other a \tD{} curved shape (cluster 2), each consisting of $1000$ observations. The first four variables, $X_1, X_2, X_3, X_4$ hold this cluster structure, and the remaining three $X_5, X_6, X_7$ are low variance noise. Because, no error was simulated in the first four variables, we would consider $T=(X_1, X_2, X_3, X_4)$ to be **the geometric structure (true model)** that we hope to capture. In practice, all variables might have some small error, and structure to detect might be in any or all variables. The 2NC7 data is simple but complex enough illustrate the new methodology. We would consider the two separated clusters as *global structure*, and the nonlinear surfaces to be relatively *local structure*. @fig-NLDR-two-curvy uses an **reasonable \gD{} NLDR layout** of 2NC7, produced by tSNE, which reveals the two clusters with moderate separation (global), and flattens the curvilinear forms while (hopefully) preserving the proximity of points (local). Note that, because cluster 2 is \tD{}, it cannot be fully captured in a \gD{} layout, and the flat rectangle is as good as can be expected. 

```{r}
#| label: fig-NLDR-two-curvy
#| echo: false
#| fig-cap: "Key steps for constructing the model on the tSNE layout ($k=2$): (a) data, (b) hexagon bins, (c) bin centroids, and (d) triangulated centroids. The layout used is a reasonable reprersntation of the 2NC7 data, which contains two curvilinear clusters that have been flattened into a triangular and rectangular shape."
#| fig-width: 12
#| fig-height: 4
#| out-width: 100%
 
nldr_two_curvy + hex_grid_two_curvy + 
  hex_grid_nonempty_two_curvy + 
  trimesh_removed_two_curvy +
  plot_layout(ncol = 4)
```

## Algorithm to represent the model in \gD{} {#sec-algorithm}

### Scale the layout data

Because the algorithm is based on distances between points, scaling the layout data to range between [0, 1] in the first coordinate, is recommended. By default, the aspect ratio produced by the NLDR $(r_1, r_2, ..., r_k)$ is preserved in the other coordinates. When $k=2$, as in hexagon binning, the default range for the second coordinate is $[0, r_2/r_1]$.

### Hexagon grid configuration

Although there are several implementations available (e.g., @dan2023), the full set of data objects needed for this project. So we have re-implemented the method so that all items needed for creating the \pD{} representation are available (illustrated in @fig-hex-param). 

The \gD{} hexagon grid is uniquely defined by the hexagon ids, $h (= 1, \dots, b)$, and their \gD{} centroids, $C_{h}^{(2)} = (c_{h1}, c_{h2})$. The algorithm inputs $b_1$, which determines $b_2$ using the NLDR ratio, producing $b = b_1 \times b_2$, the total number of bins in the grid. 

A buffer is used to ensure that the grid covers the range of data values, and provide flexibility for where the the grid positining. It is well-known that where a grid starts can change binning results, even though most documentation focuses on binwidth! The buffer ($q$) is set as a proportion of the range. By default,  $q=0.1$, which is 10% of the range. <!--The buffer should extend a full hexagon width ($a_1$) and height ($a_2$) beyond the data, in all directions.--> The lower left position where the grid starts is set to be $(s_1 = -q, s_2 = -qr_2)$, specifying the centroid of the lower left hexagon, $C_{1}^{(2)} = (c_{11}, c_{12})$. 

Considering the upper bound of the first NLDR component, the horizontal binwidth must satisfy $a_1 > (1+2q)/(b_1 -1)$. Similarly, the vertical binwidth must satisfy 

$$
a_2 \geq \frac{r_2 + q(1 + r_2)}{(b_2 - 1)}.
$${#eq-bin1a}

Because we are using regular hexagons $a_2 = \sqrt{3}a_1/2$, and @eq-bin1a can be re-written as

$$
a_1 \geq \frac{2(r_2 + q(1 + r_2))}{\sqrt{3}(b_2 - 1)}.
$$

This is a linear optimization problem, resulting in calculating $b_2$ as

$$
b_2 = \Big\lceil1 +\frac{2(r_2 + q(1 + r_2)(b_1 - 1)}{\sqrt{3}(1 + 2q)}\Big\rceil.
$$

where $\lceil . \rceil$ indicate "smallest integer greater than or equal to".

```{r}
#| label: code-illustration

# Code to draw illustration for notation
## hexagon binning to have regular hexagons

hex_grid_temp <- read_rds("data/two_nonlinear/hex_grid_temp.rds")
all_centroids_df_temp <- read_rds("data/two_nonlinear/all_centroids_df_tempp.rds")
start_pt <- read_rds("data/two_nonlinear/start_pt.rds")
d_rect <- read_rds("data/two_nonlinear/d_rect.rds")
a1 <- read_rds("data/two_nonlinear/a2_data.rds")
a2 <- read_rds("data/two_nonlinear/a1_data.rds")
l <- read_rds("data/two_nonlinear/l_data.rds")
rect_adj <- read_rds("data/two_nonlinear/rect_adj.rds")

hex_param_vis <- ggplot() + 
  geom_polygon(data = hex_grid_temp, 
               aes(x = x, 
                   y = y, 
                   group = h),
               fill = "white", 
               colour = "#bdbdbd") +
  geom_point(data = all_centroids_df_temp, aes(
    x = c_x, 
    y = c_y), 
    colour = "#31a354", size = 0.9) +
  geom_point(data = start_pt, aes(x = c_x, 
                                  y = c_y), 
             colour = "black") + 
  geom_rect(data=d_rect, 
            aes(xmin = x1min - rect_adj$x1,# - rect_adj$s1, 
                xmax = x1max - rect_adj$x1,# - rect_adj$s1, 
                ymin = x2min - rect_adj$x2,# - rect_adj$s2, 
                ymax = x2max - rect_adj$x2),# - rect_adj$s2), 
            fill = "grey70", 
            colour = "black", 
            alpha = 0.5, 
            linewidth = 0.7) +
  geom_point(data=d_rect, aes(x=x1min - rect_adj$x1, 
                              y=x2min - rect_adj$x2)) + 
  geom_point(data=d_rect, aes(x=x1max - rect_adj$x1, 
                              y=x2min - rect_adj$x2)) + 
  geom_point(data=d_rect, aes(x=x1min - rect_adj$x1, 
                              y=x2max - rect_adj$x2)) + 
  annotate("text", x=d_rect$x1min - rect_adj$x1, 
           y=d_rect$x2min - rect_adj$x2,
           label = "(0,0)", 
           hjust=-0.1, vjust=-0.3, size = 8) + 
  annotate("text", x=d_rect$x1max - rect_adj$x1, 
           y=d_rect$x2min - rect_adj$x2,
           label = "(0,1)", 
           hjust=1.1, vjust=-0.3, size = 8) + 
  annotate("text", x=d_rect$x1min - rect_adj$x1, 
           y=d_rect$x2max - rect_adj$x2,
           label = "'(0,'*y[2][max]*')'", 
           parse = TRUE,
           hjust=-0.1, vjust=1.2, size = 8) + 
  geom_segment(data=d_rect, aes(
    x = x1min  - rect_adj$x1, # 0 - 0.03, 
    y = -0.31, 
    xend = x1max - rect_adj$x1, #1 - 0.03, 
    yend = -0.31), #-0.35),
    arrow = arrow(length = unit(0.03, "npc"),
                  ends = "both"), 
    colour = "black")+
  annotate("text", x=0.5, y=-0.38, 
           label = "r[1]", parse = TRUE, colour = "black", size = 8) +
  geom_segment(data=d_rect, aes(
    x = -0.25, 
    y = x2min - rect_adj$x2, #0 - 0.05, 
    xend = -0.25, 
    yend = x2max - rect_adj$x2), #r2 - 0.05),
    arrow = arrow(length = unit(0.03, "npc"),
                  ends = "both"), 
    colour = "black")+ 
  annotate("text", x=-0.3, y=0.4, 
           label = "r[2]", parse = TRUE, colour = "black", size = 8) +
  geom_segment(data = a1, aes(
    x = x, #-0.1 + 0.2087578, 
    y = y, #-0.15, 
    xend = xend, #-0.1 + 0.2087578*2, 
    yend = yend), #-0.15),
    arrow = arrow(length = unit(0.03, "npc"),
                  ends = "both"), 
    colour = "black")+ # a1 = 0.2087578
  annotate("text", 
           x=(a1$x+a1$xend)/2, 
           y=a1$y, 
           label = "a[1]", parse = TRUE, 
           colour = "black",
           vjust = 1.2, size = 8) +
  geom_segment(data = a2, aes(
    x = x, #-0.15, 
    y = y, #-0.1*r2 + 0.1807896*2, 
    xend = xend, #-0.15, 
    yend = yend), #-0.1*r2 + 0.1807896*3),
    arrow = arrow(length = unit(0.03, "npc"),
                  ends = "both"), 
    colour = "black") + # a2 = 0.1807896
  annotate("text", x=a2$x, y=(a2$y+a2$yend)/2, 
           label = "a[2]", parse = TRUE,
           colour = "black", hjust=-0.2, size = 8) +
  annotate("text", x=-0.18, y=-0.24, 
           label = "'('*s[1]*','*s[2]*')'", parse = TRUE,
           colour = "black", size = 8) +
  geom_segment(data = l, aes(
    x = x, #-0.15, 
    y = y, #-0.1*r2 + 0.1807896*2, 
    xend = xend, #-0.15, 
    yend = yend), #-0.1*r2 + 0.1807896*3),
    arrow = arrow(length = unit(0.03, "npc"),
                  ends = "both"), 
    colour = "black") + 
  annotate("text", x=l$x + 0.03, y=(l$y+l$yend)/2, 
           label = "l", parse = TRUE,
           colour = "black", hjust=-0.2, size = 8) +
  coord_equal()
```

```{r}
#| label: fig-hex-param
#| fig-cap: "The components of the hexagon grid illustrating notation. The grey region indicates the extent of the data, which the grid must cover."
#| out-width: 30%

hex_param_vis
```

### Binning the data

Observations are grouped into bins based on their nearest centroid. This produces a reduction in size of the data from $n$ to $m$, where $m$ (number of non-empty bins) $\leq b$ (total number of bins). This can be defined using the function $u: \{1, \dots, n\} \rightarrow \{1, \dots, m\}$, where
$$u(i) = \arg\min_{j = 1, \dots, m} \sqrt{(y_{i1} - C^{(2)}_{j1})^2 + (y_{i2} - C^{(2)}_{j2})^2},$$ maps observation $i$ into $H_h = \{i| u(i) = h\}$. 
<!--By default, the bin centroid is used to represent a hexagon center (@fig-NLDR-two-curvy (c)), but any measure of center, such as a mean or weighted mean of the points within each hexagon, could be used.--> Then $\mathbfit{\widehat{Y}}_i = \mathbfit{C}^{(2)}_{u(i)} (i=1, \dots, n) \in \{\mathbfit{C}^{(2)} = (C^{(2)}_{j1}, C^{(2)}_{j2}), j=1, \dots, m\}$ are considered the fitted values, from which we can also calculate \gD{} residuals. The bin id, $h$, for each observation is needed to construct the model representation in high dimensions. 

### Indicating neighborhood

Delaunay triangulation [@lee1980;@alb2024] is used to connect centroids to indicate neighboring observations in the NLDR layout (@fig-NLDR-two-curvy (d)). These are retained to preserve the neighborhood information when the model is lifted into \pD{}. 

## Rendering the model in \pD{} {#sec-highd}

The last step is to lift the \gD{} model into \pD{} by computing \pD{} vectors that represent bin centroids. We use the \pD{} mean of the points in a given hexagon, $H_h$, denoted $C_{h}^{(p)}$, to map the centroid $C_{h}^{(2)} = (c_{h1}, c_{h2})$ to a point in \pD{}. Let the $j^{th}$ component of the \pD{} mean be

$$C_{hj}^{(p)} = \frac{1}{n_h}\sum_{i \in H_h} x_{ij}, ~~~h = 1, \dots, b;~ j=1, \dots, p; ~ n_h > 0.$$
We can consider this procedure to be a pseudo-inverse of $g$, $\mathbfit{C}^{p} = \tilde{g}(\mathbfit{C}^{(2)})$, $\tilde{g}\text{:}~ \mathbfit{C}^{(2)} \subset \mathbb{R}^2 \mapsto \mathbfit{C}^{p} \subset \mathbb{R}^p$. Then $\mathbfit{\hat{X}}_i (i=1, \dots, n) \in \{\mathbfit{C}^{(p)} = (C^{(p)}_{j1}, C^{(p)}_{j2}), j=1, \dots, m\}$ are considered the fitted values, from which we can also calculate \pD{} residuals.

```{r}
#| label: fig-best-fit-tsne
#| fig-cap: "Lifting the \\gD{} fitted model (a) into \\pD{} (b,c). Two projections of the \\pD{} fitted model overlaying the data are shown in b, c. The fit is reasonably tight with the data in cluster 1, but slightly less so in cluster 2, probably because it is $3\\text{-}D$. Notice also that, in the \\gD{} layout, the two clusters have internal gaps which creates a model with some holes. This lacy pattern happens regardless of the hyper-parameter choice for tSNE, but this doesn't severely impact the \\pD{} model representation."
#| fig-width: 8
#| fig-height: 2
#| fig-pos: "!ht"

trimesh_two_curvy_removed1_tsne_with_data +
  two_curvy_proj_tsne_first_model1 +
  two_curvy_proj_tsne_all_model1 +
  plot_layout(ncol = 3)
```


## Measuring the fit {#sec-summary}

<!-- Existing approaches -->

<!-- All NLDR methods internally optimize a quantity to produce a layout for any particular hyper-parameter set. These are not always made available in the model output, and may not be universally comparable between hyper-parameter choices and methods.  -->

<!-- Several common metrics are often used to assess the quality of any NLDR layout, based on the preservation of the global and local structure of the data. The $RNX$ curve quantifies the neighborhood agreement between \pD{} and \kD{} spaces, by computing the area under the curve ($ARNX$) across a range of neighborhood scales [@john2015]. A high value indicates better preservation of a balance of global and local structure. Random Triplet Accuracy (RTA) and Centroid Triplet Accuracy compare the order of \gD{} and \pD{} distances of random triplets of points [@yingfan2021]. High values indicate preservation of the geometry, suggesting both local and global structure preservation. The Shepard diagram [@shepard1962] and its associated Spearman correlation (SC) [@spearman1961] between \pD{} and \kD{} distances. High values indicate preservation of global structure. The Global Score (GS) measures how well an embedding retains the overall geometry of the data relative to a PCA baseline [@amid2022]. Higher values indicate better preservation of global structure. The metric RTA, SC, GS, and ARNX have been reversed (rRTA, rSC, rGS, and rARNX) so that they align with HBE - the lower the value, the better the layout. -->

<!-- Fitted values,  Error calculation-->

<!-- None of the above measures is particularly well-suited to assessing our model fit, as we will show later. Thus, we need a different approach to measuring model fit. Because the model here is similar to a confirmatory factor analysis model (see a general explanation in @brown2015), $\widehat{T} + \Epsilon$, our approach is similar to the ones used in this area. It is based on "residuals" computed as the difference between the fitted model and observed values in \pD{}. Observations are associated with their bin center, $C_{h}^{(p)}$, which are also considered to be the *fitted values*. In factor analysis language, these fitted values might also be denoted as $\widehat{X}$. The error is computed by taking the squared \pD{} Euclidean distance of points from their bin centroid, which we will call hexbin error (HBE): -->

<!-- $$ HBE = \sqrt{\frac{1}{n}\sum_{h = 1}^{m}\sum_{i = 1}^{n_h}\sum_{j = 1}^{p} (\mathbfit{x}_{hij} - C^{(p)}_{hj})^2}$${#eq-equation1}  -->

<!-- where $n$ is the number of observations, $m$ is the number of non-empty bins, $n_h$ is the number of observations in $h^{th}$ bin, $p$ is the number of variables and $\mathbfit{x}_{hij}$ is the $j^{th}$ dimensional data of $i^{th}$ observation in $h^{th}$ hexagon. We can consider $e_{hi} = \sqrt{\sum_{j = 1}^{p} (\mathbfit{x}_{hij} - C^{(p)}_{hj})^2}$ to be the residual for each observation. @fig-p-d-error-in-2d-two-curvy shows plots of $e$ as a density (a), coloring the points in the NLDR layout (b), and the points in a tour (c). It can be seen that the biggest residuals are in one cluster, which occurs due to the intentional design that one cluster is slightly \tD{} and perfectly captured by a \gD{} layout.  -->

<!--updated according to the reviewes-->

All NLDR methods internally optimize an objective function to produce a layout for a given set of hyper-parameters. These objective functions are not always made available to the user in the model output and or comparable across methods.

A range of established metrics are commonly used to assess the quality of NLDR embeddings, each emphasizing different aspects of structure preservation. For example, the $RNX$ curve summarizes neighborhood agreement between \pD{} and \kD{} spaces across a range of scales, with its area under the curve (ARNX) reflecting a balance between local and global structure preservation [@john2015]. Random Triplet Accuracy (RTA) evaluates the consistency of distance orderings among triplets of points in \gD{} and \pD{} spaces, providing a measure of geometric preservation [@yingfan2021]. The Shepard diagram [@shepard1962], together with its associated Spearman correlation (SC) [@spearman1961], assesses the monotonic relationship between pairwise distances in the two spaces and is often interpreted as a measure of global structure preservation. The Global Score (GS) compares the embedding to a PCA baseline and evaluates how well the overall geometry is retained [@amid2022].

These metrics provide useful but complementary perspectives, as they capture different and often competing objectives, such as preserving local neighborhoods versus global distances. As a result, it is common for these measures to rank embeddings differently, reflecting the multi-objective nature of NLDR rather than a deficiency of any single metric.

To facilitate comparison with our proposed measure, we use reversed forms of these metrics (rRTA, rSC, rGS, and rARNX), (XXX is this still true) so that lower values consistently indicate better performance.

While the above measures evaluate properties of the embedding itself, our goal is different: to assess how well the embedding supports a representation of the data in the original \pD{} space. This motivates a complementary approach based on model residuals.

We treat the embedding as defining a fitted representation in \pD{}, where each observation is associated with a bin centroid, $C_{h}^{(p)}$, which serves as its fitted value. The discrepancy between observed and fitted values can then be interpreted as a residual.

We define the hexbin error (HBE) as the root mean squared Euclidean distance between observations and their associated bin centroids:

$$ HBE = \sqrt{\frac{1}{n}\sum_{i = 1}^{n}\sum_{h = 1}^{m}\sum_{j = 1}^{p} (\mathbfit{x}_{ij} - C^{(p)}_{hj})^2\cdot\mathbf{1}[i \in H_h]}$${#eq-equation1}

where $n$ is the number of observations, $m$ is the number of non-empty bins, $n_h$ is the number of observations in $h^{th}$ bin, $p$ is the number of variables and $\mathbfit{x}_{ij}$ is the $j^{th}$ dimensional data of $i^{th}$ observation. We can consider $e_i = \sqrt{\sum_{j = 1}^{p} (\mathbfit{x}_{ij} - \mathbfit{\hat{x}}_{ij})^2}$ to be the residual for each observation. @fig-p-d-error-in-2d-two-curvy visualizes these residuals as a density (a), as colors in the NLDR layout (b), and in a tour (c). Larger residuals are concentrated in one cluster, reflecting the intentional design that this cluster has intrinsic higher-dimensional structure (approximately \tD{}) and therefore cannot be fully represented in a \gD{} embedding. Importantly, HBE is not intended to replace existing embedding-quality metrics. Instead, it provides a complementary diagnostic perspective, highlighting where the embedding fails to represent the original data structure in \pD{}. Differences between HBE and other metrics can therefore be interpreted as informative signals about trade-offs between local and global structure preservation.

```{r}
#| label: fig-p-d-error-in-2d-two-curvy
#| fig-cap: "Examining the distribution of residuals in a jittered dotplot (a), \\gD{} NLDR layout (b), and a tour of \\fD{} data space (c). Color indicates residual value ($e_{i}$), dark indicating high value. Most large residuals are distributed in cluster 2, which is due to the \\gD{}'s inability to capture the \\tD{} structure."
#| out-width: 100%
#| fig-width: 8
#| fig-height: 3

 error_plot_two_curvy_quasi + error_plot_two_curvy + two_curvy_proj_tsne_first_model1_coloured +
  plot_layout(ncol=3)
```


## Prediction into \gD{} {#sec-prediction}

NLDR methods focus on visualization solution for exploration without the ability to place new observations in a layout. Partially, this is also due to the methods being designed to operate only on the interpoint distance matrices instead of $\pD{}$ data. Only UMAP provides a way to predict the layout positions of new observations [@tomasz2023] when the full data is available, and this is made available in the `umap` [@leland2018] software. Other approaches not illustrated in this work, such as neural network autoencoders [@hinton2006], and parametric tSNE [@van2009], provide some mechanisms for embedding new data.  A benefit of our approach, is that with the model representation, we can predict layout positions for any method. 

The procedure for the new observation: (i) identify the closest bin centroid in \pD{}, $C^{(p)}_{h}$, and (ii) assign the corresponding centroid in \gD{}, $C^{(2)}_{h}$, as the predicted embedding. This is not entirely different from how UMAP provides predictions: 
find the nearest neighbors in \pD{}, and interpolate their positions in \gD{} to provide the predicted embedding. This can be considered to be an estimate of the mapping $g$ that maps the new observation ($\mathbfit{x}^*$) into the layout ($\mathbfit{y}^*$), $\mathbfit{y}^*  = \widehat{g}(\mathbfit{x}^*), \widehat{g}\text{:}~ \mathbb{R}^p \rightarrow \mathbb{R}^2$.

@fig-prediction-comparison shows the results for UMAP (a) and our predictions(b) made using `quollr`. The `2NC7` data is divided into training and test sets, using a 70% split. The training data is used to make the \gD{} layout. (This is similar but not quite the same as the layout on the full data seen in @fig-toy-rmse plot b.) The test set is held out for prediction. In both cases it can be seen that the test set matches the training set in the layout quite well. The main differences is that our predictions are to the bin centroid, so it looks gridded. Adding some jitter to the predicted values would make them look more similar to the UMAP predictions. Just because we can, doesn't mean we should - for this data, the method provides demonstrably reasonable results, and better than having no method at all. **Add tSNE, PHATE, TriMAP and PACMAP training and quollr predictions to this plot, in a 3x2 layout**

```{r}
#| label: fig-prediction-comparison
#| fig-cap: "Comparison of predictions made my UMAP and by our method, relative to the training data. Dark points are the predicted test set, and lighter points are the training sample used to create the UMAP layout."
#| fig-height: 2
#| fig-width: 4
#| out-width: 80%
# Code to generate the saved data is in scripts/two_nonlinear/compare_pred_approaches.R
d_2NC7_tr <- read_rds(here::here("data/two_nonlinear/2NC7_tr.rds"))
d_2NC7_ts <- read_rds(here::here("data/two_nonlinear/2NC7_ts.rds"))

UMAP_d_2NC7_tr <- read_rds("data/two_nonlinear/UMAP-2NC7-tr.rds")
load(file="data/two_nonlinear/UMAP_fit.rda")

UMAP_d_2NC7_ts_pred <- read_rds("data/two_nonlinear/two_non_linear_diff_shaped_close_clusters_umap_predict_test.rds")

num_bins_x_2NC7 <- 40

algo_obj_2NC7 <- fit_highd_model(
  highd_data = d_2NC7_tr[, 1:8],
  nldr_data = UMAP_d_2NC7_tr[, 1:3],
  b1 = num_bins_x_2NC7,
  q = 0.1,
  hd_thresh = 0)

## Predict positions using quollr
d_2NC7_ts_qpred <- predict_emb(highd_data = d_2NC7_ts[,1:8],
                               model_highd = algo_obj_2NC7$model_highd,
                               model_2d = algo_obj_2NC7$model_2d)
d_2NC7_ts_qpred <- d_2NC7_ts_qpred |>
  rename(UMAP1 = pred_emb_1,
         UMAP2 = pred_emb_2)

# Scale to match quollr scaled data
UMAP_2NC7_tr_scaled <- algo_obj_2NC7$nldr_scaled_obj$scaled_nldr
ar <- (algo_obj_2NC7$nldr_scaled_obj$lim2[2]-
         algo_obj_2NC7$nldr_scaled_obj$lim2[1])/
      (algo_obj_2NC7$nldr_scaled_obj$lim1[2]-
         algo_obj_2NC7$nldr_scaled_obj$lim1[1])
UMAP_2NC7_ts_pred_scaled <- UMAP_d_2NC7_ts_pred |>
  mutate(UMAP1 = (UMAP1 - algo_obj_2NC7$nldr_scaled_obj$lim1[1])/
           (algo_obj_2NC7$nldr_scaled_obj$lim1[2]-
              algo_obj_2NC7$nldr_scaled_obj$lim1[1]),
         UMAP2 = (UMAP2 - algo_obj_2NC7$nldr_scaled_obj$lim2[1])/
           (algo_obj_2NC7$nldr_scaled_obj$lim2[2]-
              algo_obj_2NC7$nldr_scaled_obj$lim2[1])*ar)

# Comparison 1: training vs UMAP predictions
d_umap <- bind_rows(UMAP_2NC7_tr_scaled, UMAP_2NC7_ts_pred_scaled) |>
  mutate(type = factor(c(rep("train", NROW(UMAP_2NC7_tr_scaled)),
                         rep("test", NROW(UMAP_2NC7_ts_pred_scaled)))))

pcompare1 <- ggplot(d_umap,
       aes(x = UMAP1, y = UMAP2, colour = type)) +
  geom_point(alpha = 0.5, size = 0.5) +
  scale_color_brewer("", palette = "Paired", direction = -1) +
  ggtitle("a. UMAP") +
  theme(aspect.ratio = 1,
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        legend.position = "none")

# Comparison 2: training vs quollr predictions
d_quollr <- bind_rows(UMAP_2NC7_tr_scaled, d_2NC7_ts_qpred) |>
  mutate(type = factor(c(rep("train", NROW(UMAP_2NC7_tr_scaled)),
                         rep("test", NROW(d_2NC7_ts_qpred)))))

pcompare2 <- ggplot(d_quollr,
       aes(x = UMAP1, y = UMAP2, colour = type)) +
  geom_point(alpha = 0.5, size = 0.5) +
  scale_color_brewer("", palette = "Paired", direction = -1) +
  ggtitle("b. quollr") +
  theme(aspect.ratio = 1,
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        legend.position = "none")

pcompare1 + pcompare2 + plot_layout(ncol=2)

```
<!-- NLDR methods are primarily designed for visualization and exploration rather than reconstruction, and do not explicitly provide out-of-sample prediction. Of the five methods studied here, only UMAP provides a `predict()` function for embedding new data points based on the learned manifold [@tomasz2023]. Several other approaches, not used here, PCA, neural network autoencoders [@hinton2006], and parametric tSNE [@van2009] support prediction.  -->

<!-- A benefit of our approach is that for any NLDR method, it provides a way to predict the layout position of a new observation, $x'$. The steps are (1) determine the closest bin centroid in \pD{}, $C^{(p)}_{h}$, and (2) predict the embedding to be the bin centroid in \gD{}, $C^{(2)}_{h}$.  -->



<!-- To provide an initial assessment, we compare this approach with the UMAP `predict()` function using a train/test split. Predictions are evaluated in terms of neighborhood preservation and their consistency with the fitted model representation. The results suggest that the approach provides a reasonable approximation for placing new observations relative to the fitted structure.

This method is intended as a simple extension of the model rather than as a replacement for established out-of-sample embedding techniques. A more comprehensive evaluation across a broader range of datasets and methods remains an important direction for future work.-->

```{r}
#| label: prediction-two-curvy
#| eval: false
# NEED TO REMOVE THIS CODE

predict_highdvis1 <- read_rds("data/two_nonlinear/test_two_non_linear_diff_shaped_close_clusters_highd_vis_predict_bin_13.rds")

predict_highdvis2 <- read_rds("data/two_nonlinear/test_two_non_linear_diff_shaped_close_clusters_highd_vis_predict_bin_30.rds")

predict_highdvis3 <- read_rds("data/two_nonlinear/test_two_non_linear_diff_shaped_close_clusters_highd_vis_predict_bin_54.rds")

umap_test <- read_rds("data/two_nonlinear/test_umap_predict_scaled_two_non_linear_diff_shaped_close_clusters.rds")

test_two_curvy_umap1 <- ggplot(umap_test, 
                                  aes(x = UMAP1, y = UMAP2)) +
  geom_point(alpha = 0.5, colour = "#d0d1e6", size = 0.5) +
  geom_point(data = predict_highdvis1, aes(
    x = pred_emb_1, y = pred_emb_2
  ), alpha = 0.5, colour = "#045a8d", size = 0.5) +
  interior_annotation(label = "a1", cex = 1.5)

test_two_curvy_umap2 <- ggplot(umap_test, 
                                  aes(x = UMAP1, y = UMAP2)) +
  geom_point(alpha = 0.5, colour = "#d0d1e6", size = 0.5) +
  geom_point(data = predict_highdvis2, aes(
    x = pred_emb_1, y = pred_emb_2
  ), alpha = 0.5, colour = "#045a8d", size = 0.5) +
  interior_annotation(label = "b1", cex = 1.5)

test_two_curvy_umap3 <- ggplot(umap_test, 
                                  aes(x = UMAP1, y = UMAP2)) +
  geom_point(alpha = 0.5, colour = "#d0d1e6", size = 0.5) +
  geom_point(data = predict_highdvis3, aes(
    x = pred_emb_1, y = pred_emb_2
  ), alpha = 0.5, colour = "#045a8d", size = 0.5) +
  interior_annotation(label = "c1", cex = 1.5)

```

```{r}
#| eval: false
umap_test <- read_rds("data/two_nonlinear/test_umap_scaled_two_non_linear_diff_shaped_close_clusters.rds")

test_two_curvy_umap11 <- ggplot(umap_test, 
                                  aes(x = UMAP1, y = UMAP2)) +
  geom_point(alpha = 0.5, colour = "#d0d1e6", size = 0.5) +
  geom_point(data = predict_highdvis1, aes(
    x = pred_emb_1, y = pred_emb_2
  ), alpha = 0.5, colour = "#045a8d", size = 0.5) +
  interior_annotation(label = "a1", cex = 1.5)

test_two_curvy_umap21 <- ggplot(umap_test, 
                                  aes(x = UMAP1, y = UMAP2)) +
  geom_point(alpha = 0.5, colour = "#d0d1e6", size = 0.5) +
  geom_point(data = predict_highdvis2, aes(
    x = pred_emb_1, y = pred_emb_2
  ), alpha = 0.5, colour = "#045a8d", size = 0.5) +
  interior_annotation(label = "b1", cex = 1.5)

test_two_curvy_umap31 <- ggplot(umap_test, 
                                  aes(x = UMAP1, y = UMAP2)) +
  geom_point(alpha = 0.5, colour = "#d0d1e6", size = 0.5) +
  geom_point(data = predict_highdvis3, aes(
    x = pred_emb_1, y = pred_emb_2
  ), alpha = 0.5, colour = "#045a8d", size = 0.5) +
  interior_annotation(label = "c1", cex = 1.5)
```

```{r}
#| eval: false
dist_comp_quollr_13 <- read_rds("data/two_nonlinear/quollr_predict_13_dist.rds")

dist_comp_umap_predict <- read_rds("data/two_nonlinear/umap_predict_dist.rds")

diff_quollr_13_umap_predict <- dist_comp_quollr_13 |>
  mutate(diff = quollr_predict1_dist - dist_comp_umap_predict$umap_predict_dist)

dist_cmp_plt11 <- ggplot(diff_quollr_13_umap_predict, aes(x = diff, y = after_stat(density))) +
  geom_density(color = "red", linewidth = 1)


dist_comp_quollr_30 <- read_rds("data/two_nonlinear/quollr_predict_30_dist.rds")

diff_quollr_30_umap_predict <- dist_comp_quollr_30 |>
  mutate(diff = quollr_predict2_dist - dist_comp_umap_predict$umap_predict_dist)

dist_cmp_plt12 <- ggplot(diff_quollr_30_umap_predict, aes(x = diff, y = after_stat(density))) +
  geom_density(color = "red", linewidth = 1)

dist_comp_quollr_54 <- read_rds("data/two_nonlinear/quollr_predict_54_dist.rds")

diff_quollr_54_umap_predict <- dist_comp_quollr_54 |>
  mutate(diff = quollr_predict3_dist - dist_comp_umap_predict$umap_predict_dist)

dist_cmp_plt13 <- ggplot(diff_quollr_54_umap_predict, aes(x = diff, y = after_stat(density))) +
  geom_density(color = "red", linewidth = 1)
```

```{r}
#| eval: false
dist_comp_quollr_13 <- read_rds("data/two_nonlinear/quollr_predict_13_dist.rds")

dist_comp_umap_test <- read_rds("data/two_nonlinear/umap_test_dist.rds")

diff_quollr_13_umap_test <- dist_comp_quollr_13 |>
  mutate(diff = quollr_predict1_dist - dist_comp_umap_test$umap_test_dist)

dist_cmp_plt21 <- ggplot(diff_quollr_13_umap_test, aes(x = diff, y = after_stat(density))) +
  geom_density(color = "red", linewidth = 1)


dist_comp_quollr_30 <- read_rds("data/two_nonlinear/quollr_predict_30_dist.rds")

diff_quollr_30_umap_test <- dist_comp_quollr_30 |>
  mutate(diff = quollr_predict2_dist - dist_comp_umap_test$umap_test_dist)

dist_cmp_plt22 <- ggplot(diff_quollr_30_umap_test, aes(x = diff, y = after_stat(density))) +
  geom_density(color = "red", linewidth = 1)

dist_comp_quollr_54 <- read_rds("data/two_nonlinear/quollr_predict_54_dist.rds")

diff_quollr_54_umap_test <- dist_comp_quollr_54 |>
  mutate(diff = quollr_predict3_dist - dist_comp_umap_test$umap_test_dist)

dist_cmp_plt23 <- ggplot(diff_quollr_54_umap_test, aes(x = diff, y = after_stat(density))) +
  geom_density(color = "red", linewidth = 1)

diff_umap_predict_test <- dist_comp_umap_predict |>
  mutate(diff = umap_predict_dist - dist_comp_umap_test$umap_test_dist)

dist_cmp_plt <- ggplot(diff_umap_predict_test, aes(x = diff, y = after_stat(density))) +
  geom_density(color = "red", linewidth = 1)

```

```{r}
#| label: predict
#| fig-cap: ""
#| fig-width: 9
#| fig-height: 15
#| fig-pos: "!ht"
#| eval: false

test_two_curvy_umap1 + test_two_curvy_umap2 + test_two_curvy_umap3 +
  test_two_curvy_umap11 + test_two_curvy_umap21 + test_two_curvy_umap31 +
  dist_cmp_plt11 + dist_cmp_plt12 + dist_cmp_plt13 +
  dist_cmp_plt21 + dist_cmp_plt22 + dist_cmp_plt23 +
  dist_cmp_plt +
  plot_layout(ncol = 3)
```

## Interactive graphics {#sec-interactive}

Matching points in the \gD{} layout with their positions in \pD{} is useful when assessing the fit, deciding on a layout, and basically understanding what a layout communicates about the data. The interactive \gD{} layout [@chapman2020] and the langevitour [@harisson2024] view with the fitted model overlaid can be linked using a browsable HTML widget (@joe2025, @joe2024). A rectangular "brush" is used to select points in one plot, which will highlight the corresponding points in the other plot(s). Because the langevitour is dynamic, brush events that become active will pause the animation, so that a user can interrogate the current view. This approach is illustrated on the examples in the Appendix, and video supplements, to show how it can help to understand how an NLDR has organized the observations.

## Choosing from a selection of layouts {#sec-bestfit}
<!--updated according to the reviews-->

@fig-toy-rmse shows the diagnostic plot of HBE vs binwidth (i) for a selection of plots from which to use. Color in all plots indicates the layout (a-f). The details of how they were generated is in the Appendix. There is nothing special about this selection of layouts other than, all five methods were used: (a, f) tSNE, (b) UMAP, (c) PHATE, (d) TriMAP, (e) PACMap. (XXX check these) They represent different patterns, but all agree on there being two clusters, with varying distance and shape between them. The process to make a decision is:

1. Fit the model to a range of binwidths for each layout, calculating HBE.
2. Plot each trace against binwidth. 
3. Better representations will have universally lower HBE. When the lines cross one needs to decide whether the local structure (smaller binwidth) is more important than larger structure (large binwidth).
4. Examine the model for the chosen layout in \pD{}. The exact choice of binwidth isn't important, but we recommend a reasonably small one, so that wireframe is smoother, and there is a chance to see small details of departures between the fitted model and the data. (See @fig-best-fit-tsne b, c for this data.)

For this set of layouts, either (a) or (f) would be preferable to the other layouts - we used (a) to illustrate the approach in the previous sections. 

Note that, using binwidth as the horizontal axis is a choice. An alternative is to use average bin count. In theory, this is the more instinctive choice. It summarizes the grid occupancy of the data; smaller binwidth for one layout might be equivalent to a larger binwidth in a different layout. Using binwidth sets up these possible mismatches. However, 
it tends to lose the ability to penalize layouts that exaggerate inter-cluster distance (see plots in the Appendix).

It can also be useful to calculate the suite of other metrics to measure the fit, for each layout, which is done in plot ii. The comparison with HBE requires that a single value is chosen. Here $a_1=0.05$ is used, but the actual choice wouldn't substantially matter because the ordering of the curves is quite consistent across binwidths. Note that in this parallel coordinate plot, the individual metrics are individually scaled to range between 0 and 1.  HBE shows most agreement with rARNX, both rank a and f the best. They differ in how they rate (1) layout d - HBE would consider it the worst, and rARNX to be not quite as good as a and f - and (2) layout c - rARNX considers it to be the worst, and HBE in the mddle. The other three metrics roughly agree with each other, and almost the polar opposite of HBE. This is most likely explained by the tension between local and global structure. Most will prefer bigger distance between clusters in the layout, or at least not penalise this. The problem is that this leads to incorrect perception when the clusters are not widely separated, as is the case for this data (@fig-best-fit-tsne). 

The final check is to examine the chosen model fit (or several top-rated fits) in the \pD{} data space @fig-best-fit-tsne (b, c). As with all model fitting, if the data closely hugs the model representation, in all projections shown by the tour, then the choice is further supported. This is a sutiable approach to better understand why metrics disagree on layouts. The Appendix has a comparison of the model fits for layout a and c. Layout (c) twists extremely along the 2- and \tD{} manifolds where the data lies. Twisting in the fitted model appears to be very common for all methods for many data sets, which one could imagine a solution for squashing higher dimensions into lower ones while preserving neighborhood. In this case it is quite extreme. 

You might find that some elements of the \pD{} structure is captured, but others are not. If this happens, a strategy break the data into subsets and repeat the process of deciding on a layout. Showing multiple layouts for a data set may better illustrate its nature. This might be advisable with 2NC7 also. At first glance layout (b) looks prefable, because the distribution of observations within each cluster is smooher than in either (a) or (f). This likely reflects that the layout better captures the local structure better. 

<!--illustrates the approach to compare the fits for different representations and assess the strength of any fit. What does it mean to be a best fit for this problem? Analysts use an NLDR layout to display the structure present in high-dimensional data in a convenient \gD{} display. It is a competitor to linear dimension reduction that can better represent nonlinear associations, such as clusters. However, these methods can hallucinate, suggesting patterns that don't exist, and grossly exaggerate other patterns. Having a layout that best fits the high-dimensional structure is desirable, but more important is to identify bad representations so they can be avoided. The goal is to help users decide on the most useful and appropriate low-dimensional representation of the high-dimensional data.-->

<!-- A particular pattern that we commonly see is that analysts tend to pick layouts with clusters that have big separations between them. When you examine their data in a tour, it is almost always that we see there are no big separations, and actually often the suggested clusters are not even present. While we don't expect that analysts include animated gifs of tours in their papers, we should expect that any \gD{} representation adequately indicates the clustering that is present, and honestly show lack of separation or lack of clustering when it doesn't exist. It is important for analysts to have tools to select the accurate representation, not the pretty but wrong representation.-->

```{r}
#| label: read-two-curvy-clust-nldr
# Read a variety of different NLDR representations of two_non_linear_diff_shaped_close_clusters
# and plot them on same aspect ratio
tsne_two_curvy <- read_rds("data/two_nonlinear/two_non_linear_diff_shaped_close_clusters_tsne_perplexity_47.rds")

nldr_two_curvy1 <- tsne_two_curvy |>
  ggplot(aes(x = tSNE1,
             y = tSNE2))+
  geom_point(alpha=0.1, size=1, colour='#a65628') +
  interior_annotation("a") +
  theme(aspect.ratio = 1) 

tsne_two_curvy <- read_rds("data/two_nonlinear/two_non_linear_diff_shaped_close_clusters_tsne_perplexity_62.rds")

nldr_two_curvy6 <- tsne_two_curvy |>
  ggplot(aes(x = tSNE1,
             y = tSNE2))+
  geom_point(alpha=0.1, size=1, colour='#984ea3') +
  interior_annotation("f") +
  theme(aspect.ratio = 1) 

umap_two_curvy <- read_rds("data/two_nonlinear/two_non_linear_diff_shaped_close_clusters_umap_n-neigbors_15_min-dist_0.1.rds")

nldr_two_curvy2 <- umap_two_curvy |>
  ggplot(aes(x = UMAP1,
             y = UMAP2)) +
  geom_point(alpha=0.1, size=1, colour='#636363') +
  interior_annotation("b", c(0.5, 0.88)) +
  theme(aspect.ratio = 1) 

phate_two_curvy <- read_rds("data/two_nonlinear/two_non_linear_diff_shaped_close_clusters_phate_knn_5.rds")

nldr_two_curvy3 <- phate_two_curvy |>
  ggplot(aes(x = PHATE1,
             y = PHATE2))+
  geom_point(alpha=0.1, size=1, colour='#ff7f00') +
  interior_annotation("c") +
  theme(aspect.ratio = 1) 

trimap_two_curvy <- read_rds("data/two_nonlinear/two_non_linear_diff_shaped_close_clusters_trimap_n-inliers_12_n-outliers_4_n-random_3.rds")

nldr_two_curvy4 <- trimap_two_curvy |>
  ggplot(aes(x = TriMAP1,
             y = TriMAP2))+
  geom_point(alpha=0.1, size=1, colour='#4daf4a') +
  interior_annotation("d") +
  theme(aspect.ratio = 1)

pacmap_two_curvy <- read_rds("data/two_nonlinear/two_non_linear_diff_shaped_close_clusters_pacmap_n-neighbors_10_init_random_MN-ratio_0.5_FP-ratio_2.rds")

nldr_two_curvy5 <- pacmap_two_curvy |>
  ggplot(aes(x = PaCMAP1,
             y = PaCMAP2))+
  geom_point(alpha=0.1, size=1, colour='#e41a1c') +
  interior_annotation("e") +
  theme(aspect.ratio = 1)
```

<!--two_nonlinear/04_gen_mse_for_diff_methods.R-->
```{r}
#| label: combine-data-two_curvy

error_two_curvy_umap <- read_rds("data/two_nonlinear/error_two_non_linear_diff_shaped_close_clusters_umap.rds")
error_two_curvy_tsne <- read_rds("data/two_nonlinear/error_two_non_linear_diff_shaped_close_clusters_tsne.rds")
error_two_curvy_phate <- read_rds("data/two_nonlinear/error_two_non_linear_diff_shaped_close_clusters_phate.rds")
error_two_curvy_trimap <- read_rds("data/two_nonlinear/error_two_non_linear_diff_shaped_close_clusters_trimap.rds")
error_two_curvy_pacmap <- read_rds("data/two_nonlinear/error_two_non_linear_diff_shaped_close_clusters_pacmap.rds")
error_two_curvy_tsne2 <- read_rds("data/two_nonlinear/error_two_non_linear_diff_shaped_close_clusters_tsne2.rds")

error_two_curvy <- bind_rows(error_two_curvy_umap, 
                         error_two_curvy_tsne,
                         error_two_curvy_phate,
                         error_two_curvy_trimap,
                         error_two_curvy_pacmap,
                         error_two_curvy_tsne2)

error_two_curvy_filtered <- error_two_curvy |> 
  dplyr::filter(b1 == 24) |>
  dplyr::select(method, HBE)

error_two_curvy <- error_two_curvy |>
  mutate(a1 = round(a1, 2)) |>
  filter(b1 >= 5) |>
  group_by(method, a1) |>
  filter(HBE == min(HBE)) |>
  ungroup() |>
  mutate(method = factor(method, levels = c("PaCMAP", "PHATE", "TriMAP", "tSNE", "UMAP", "tSNE2")))
```

```{r}
#| label: error-comp-two_curvy

HBE_plot_two_curvy <-
  plot_hbe(error_two_curvy) +
  interior_annotation("i", position = c(0.07, 0.9)) +
  scale_x_continuous(breaks = seq(0, 0.25, 0.05)) +
    #sort(unique(error_two_curvy$a1))[
    #  seq(1, length(
    #    unique(error_two_curvy$a1)), 
    #    by = 5)]) +
  scale_color_manual(
    values=c("PaCMAP" = '#e41a1c', "PHATE" = '#ff7f00', 
             "TriMAP" = '#4daf4a', "tSNE" = '#a65628', 
             "UMAP" = '#636363', "tSNE2" = '#984ea3')) +
  theme(aspect.ratio = 0.685) 
```

<!--two_nonlinear/07_example_evaluation_metrics.R-->
```{r}
nldr_metrics <- read_rds("data/two_nonlinear/nldr_eval_metrics.rds")

nldr_metrics <- inner_join(nldr_metrics, error_two_curvy_filtered, by = c("Method" = "method")) |>
  rename("sc" = "r") |>
  mutate(`rARNX` = 1 - `AUC(R_NX)`) |>
  mutate(`rRTA` = 1 - RTA) |>
  mutate(`rSC` = 1 - sc) |>
  mutate(`rGS` = 1 -  GS)

comp_metrics <- plot_comp_metrics(nldr_metrics) +
  interior_annotation("ii", position = c(0.95, 0.9)) +
  interior_annotation("better", position = c(0.12, 0.15)) +
  interior_annotation("worse", position = c(0.12, 0.85)) +
  scale_color_manual(
    values=c("PaCMAP" = '#e41a1c', "PHATE" = '#ff7f00', 
             "TriMAP" = '#4daf4a', "tSNE" = '#a65628', 
             "UMAP" = '#636363', "tSNE2" = '#984ea3')) +
  theme(aspect.ratio = 0.685)
```

```{r}
#| fig-cap: "Assessing which of the 6 NLDR layouts (a-f) on the 2NC7 data is the better representation using (i) HBE for varying binwidth ($a_1$), and (ii) scaled evaluation metrics (rRTA, rSC, rGS, rARNX, and HBE using $a_1=0.05$). Color represents the NLDR layout. Using HBE (i), layouts a and f are nearly indistinguishable from each other but have universally lower values than the others: either layout could adequately represent the data. HBE mostly agrees with rARNX, except for layout d. The other metrics rank the layouts in reverse oder! The problem though is that they favor more separation, which is misleading for this data. This can be validated by viewing the data in a tour."
#| label: fig-toy-rmse
#| out-width: 100%
#| fig-height: 3.5
#| fig-width: 5
#| fig-pos: "t"

left_column <- HBE_plot_two_curvy/comp_metrics +
  plot_layout(heights = c(1, 1))

# Arrange six NLDR plots into two columns (right side)
right_columns <- wrap_plots(
  nldr_two_curvy1, nldr_two_curvy2, nldr_two_curvy3,
  nldr_two_curvy4, nldr_two_curvy5, nldr_two_curvy6,
  ncol = 2
)

# Place comp_metrics below and match full width
final_plot_two_curvy <- left_column | right_columns

# Show the result
final_plot_two_curvy
```

<!-- @fig-toy-rmse compares the metrics rARNX, rRTA, rSC, rGS, along with HBE computed on $a_1=0.05$ for the six layouts shown in @fig-toy-rmse. This is a parallel coordinate plot where the y-axis shows a normalized score to ensure the metrics are on the same scale. Each line corresponds to one layout. 

There is some agreement between the metrics. All, except rARNX and HBE, agree that layout d is best. rARNX and HBE agree that layout f is best or very close to best. Layout a is best according to HBE and rARNX but considered to be much less optimal by rRTA, rSC and rGS. Layout c is considered poor by rRTA, rSC, rGS, and rARNX. This illustrates how difficult it is to use the numerical metrics alone to decide on the best layout. 

The problem with rSC is that correlation is not a good measure in the presence of clusters - the further the clusters are apart in the layout, produces a Shepard plot with two clusters of distances, which will produce a high correlation value. Similar reasoning would explain why rRTA and rGS behave similarly: they put too much emphasis on the global structure. Thus, for the 2NC7 data, further apart clusters score better, overly emphasizing that there are two clusters, even though this separation is not accurately reflecting the difference in \pD{}.-->


## Applications {#sec-applications}

To illustrate the approach, we use two examples: PBMC3k data (single cell gene expression), where an NLDR layout is used to represent cluster structure present in the \pD{} data, and MNIST hand-written digits, where NLDR is used to represent a low-dimensional nonlinear manifold in \pD{}.

### PBMC3k {#sec-pbmc}

This is a benchmark single-cell RNA-Seq data set collected on Human Peripheral Blood Mononuclear Cells (PBMC3k) as used in @pbmc2019. Single-cell data measures the gene expression of individual cells in a sample of tissue (see, for example, @haque2017). This type of data is used to obtain an understanding of cellular level behavior and heterogeneity in their activity. Clustering of single-cell data is used to identify groups of cells with similar expression profiles. NLDR is often used to summarize the cluster structure. Usually, NLDR does not use the cluster labels to compute the layout, but uses color to represent the cluster labels when it is plotted. 

In this data, there are $2622$ single cells and $1000$ gene expressions (variables). Following the same pre-processing as @chen2024, different NLDR techniques were performed on the first nine principal components. @fig-NLDR-variety shows this data using a variety of methods and different hyper-parameters. You can see that the result is wildly different depending on the choices. Layout a is a reproduction of the layout that was published in @chen2024. This layout suggests that the data has three very well-separated clusters, each with an odd shape. The question is whether this accurately represents the cluster structure in the data, or whether they should have chosen b or c or d or e or f or g or h. This is what our new method can help with -- to decide which is the more accurate \gD{} representation of the cluster structure in the \pD{} data. 

@fig-pbmc-rmse shows HBE across a range of binwidths ($a_1$) for each of the layouts in @fig-NLDR-variety. The layouts were generated using tSNE and UMAP with various hyper-parameter settings, while PHATE, PaCMAP, and TriMAP were applied using their default settings. Lines are color coded to match the color of the layouts shown on the right. Lower HBE indicates a better fit. Using a range of binwidths shows how the model changes, with possibly the best model being one that is universally low HBE across all binwidths. It can be seen that layout f is sub-optimal with universally higher HBE. Layout a, the published one, is better, but it is not as good as layouts b, d, or e. With some imagination, the layout d perhaps shows three barely distinguishable clusters. Layout e shows three, possibly four, clusters that are more separated. The choice reduces from eight to these two. Layout d has slightly better HBE when the $a_1$ is small, but layout e beats it at larger values. Thus, we could argue that layout e is the most accurate representation of the cluster structure of these eight.

To further assess the choices, we need to look at the model in the data space, by using a tour to show the wireframe model overlaid on the data in the $9\text{-}D$ space (@fig-model-pbmc-author-proj). Here we compare the published layout (a) versus what we argue is the best layout (e). The top row (a1, a2, a3) corresponds to the published layout, and the bottom row (e1, e2, e3) corresponds to the optimal choice according to our procedure. The middle and right plots show two projections. The primary difference between the two models is that the model of layout e does not fill out to the extent of the data but concentrates in the center of each point cloud. Both suggest that three clusters are a reasonable interpretation of the structure, but the layout e more accurately reflects the separation between them, which is small.

<!--Fit the best model for author suggestion and compute error-->
<!--pbmc3k/13_gen_model_with_UMAP.R-->
```{r}
#| label: hexbin-pbmc
#| 
umap_pbmc_scaled_with_cluster <- read_rds("data/pbmc3k/umap_pbmc_scaled_with_cluster.rds")

umap_tr_from_to_df_pbmc <- read_rds("data/pbmc3k/umap_tr_from_to_df_pbmc.rds")
  
trimesh_removed_pbmc <- ggplot() + 
  geom_point(
    data = umap_pbmc_scaled_with_cluster,
    aes(
      x = UMAP1,
      y = UMAP2,
      colour = cluster
    ),
    alpha = 0.2
  )  +
  geom_segment(data = umap_tr_from_to_df_pbmc, 
               aes(
                 x = x_from, 
                 y = y_from, 
                 xend = x_to, 
                 yend = y_to),
               colour = "#000000",
               linewidth = 1) +
  scale_color_manual(values = c("#8dd3c7", "#fdb462", "#bebada")) +
  interior_annotation("a1", c(0.08, 0.9), cex = 2) +
  theme(
    aspect.ratio = 1
  )
```


```{r}
#| label: pbmc-umap-model-proj

## First projection
proj_obj1 <- read_rds("data/pbmc3k/pbmc_umap_proj_obj1.rds")

pbmc_proj_umap_model1 <- plot_proj(
  proj_obj = proj_obj1, 
  point_param = c(1.5, 0.2), # size, alpha, color
  plot_limits = c(-0.65, 0.65), 
  axis_text_size = 4,
  is_category = TRUE) +
  interior_annotation(label = "a2", cex = 2) + 
  scale_color_manual(values = c("#8dd3c7", "#fdb462", "#bebada"))

## Second projection
proj_obj2 <- read_rds("data/pbmc3k/pbmc_umap_proj_obj2.rds")

pbmc_proj_umap_model2 <- plot_proj(
    proj_obj = proj_obj2, 
    point_param = c(1.5, 0.2), # size, alpha, color
    plot_limits = c(-1, 0.75), 
    axis_text_size = 4,
    is_category = TRUE) +
  interior_annotation(label = "a3", cex = 2) +
  scale_color_manual(values = c("#8dd3c7", "#fdb462", "#bebada"))

```

<!--pbmc3k/08_gen_mse_for_diff_methods.R-->
```{r}
#| label: combine-data-pbmc

error_pbmc_umap <- read_rds("data/pbmc3k/error_pbmc_umap_30_min_dist_0.3.rds")
error_pbmc_umap2 <- read_rds("data/pbmc3k/error_pbmc_umap_5_min_dist_0.01.rds")
error_pbmc_umap3 <- read_rds("data/pbmc3k/error_pbmc_umap_5_min_dist_0.8.rds")
error_pbmc_tsne <- read_rds("data/pbmc3k/error_pbmc_tsne_5.rds")
error_pbmc_tsne2 <- read_rds("data/pbmc3k/error_pbmc_tsne_30.rds")
error_pbmc_phate <- read_rds("data/pbmc3k/error_pbmc_phate_5.rds")
error_pbmc_trimap <- read_rds("data/pbmc3k/error_pbmc_trimap_12_4_3.rds")
error_pbmc_pacmap <- read_rds("data/pbmc3k/error_pbmc_pacmap_30_random_0.9_5.rds")

error_pbmc <- bind_rows(error_pbmc_umap, 
                        error_pbmc_umap2,
                        error_pbmc_umap3,
                        error_pbmc_tsne,
                        error_pbmc_tsne2,
                        error_pbmc_phate,
                        error_pbmc_trimap,
                        error_pbmc_pacmap)

error_pbmc <- error_pbmc |>
  mutate(a1 = round(a1, 2)) |>
  filter(b1 >= 5) |>
  group_by(method, a1) |>
  filter(HBE == min(HBE)) |>
  ungroup()

error_pbmc <- error_pbmc |>
  mutate(method = factor(method,
                         levels = c("UMAP_5_min_dist_0.8", "UMAP_30_min_dist_0.3", "UMAP_5_min_dist_0.01","tsne_5", "tsne_30", "phate_5", "trimap_12_4_3", "pacmap")))
```

```{r}
#| label: error-comp-pbmc

error_plot_pbmc <- plot_hbe(error_pbmc) +
  scale_x_continuous(breaks = seq(0, 0.25, 0.05)) +
                       #sort(unique(error_pbmc$a1))[c(1, 5, 9, 13, 17, 21, 26)]) +
  scale_color_manual(values=c('#999999','#a65628','#e41a1c','#377eb8','#4daf4a','#ff7f00','#984ea3','#f781bf')) +
  ylab("HBE")
```

```{r}
#| fig-cap: "Assessing which of the 8 NLDR layouts on the PBMC3k data  (shown in @fig-NLDR-variety) is the better representation using HBE for varying (i) binwidth ($a_1$), and (ii) average bin count ($\\bar{n}_h$). Color used for the lines and points in the left plot and in the scatterplots represents NLDR layout (a-h). Layout f is universally poor. Layouts a, c, g, and h that show large separations between clusters are universally suboptimal. Layout d with little separation performs well at tiny binwidth (where most points are in their own bin) and poorly as binwidth increases. The choice of the best is between layouts b and e, which have small separations between oddly shaped clusters. Layout e is chosen as the best. Plot (ii), which accounts for the density within clusters by using average bin count, shows reduced differences between layouts, indicating that part of the variation in (i) is driven by cluster density rather than true structural differences."
#| label: fig-pbmc-rmse
#| out-width: 100%

design <- "
AABB
AABB
AABB
AABB
"
error_plot_pbmc + 
  wrap_plots(nldr1c, nldr2c, nldr3c, nldr4c,
             nldr5c, nldr6c, nldr7c, nldr8c, ncol = 2) +
  plot_layout(design = design)
```

<!--best choice-->
<!--Fit the best model and compute error-->
<!--pbmc3k/14_gen_model_with_tSNE.R-->
```{r}
#| label: tsne-num-bins-pbmc
tsne_pbmc_scaled_best_with_cluster <- read_rds("data/pbmc3k/tsne_pbmc_scaled_best_with_cluster.rds")

tsne_tr_from_to_df_pbmc <- read_rds("data/pbmc3k/tsne_tr_from_to_df_pbmc.rds")

trimesh_removed_pbmc_best <- ggplot() + 
  geom_point(
    data = tsne_pbmc_scaled_best_with_cluster,
    aes(
      x = tSNE1,
      y = tSNE2,
      colour = cluster
    ),
    alpha = 0.3
  ) +
  geom_segment(data = tsne_tr_from_to_df_pbmc, 
               aes(
                 x = x_from, 
                 y = y_from, 
                 xend = x_to, 
                 yend = y_to),
               colour = "#000000",
               linewidth = 1) +
  scale_color_manual(values = c("#8dd3c7", "#fdb462", "#bebada")) +
  interior_annotation("e1", cex = 2, c(0.08, 0.9)) +
  theme(
    aspect.ratio = 1
  )
```

```{r}
#| label: pbmc-tsne-model-proj-best

## First projection
proj_obj1 <- read_rds("data/pbmc3k/pbmc_tsne_proj_obj1.rds")

pbmc_proj_tsne_model1_best <- plot_proj(
  proj_obj = proj_obj1,
  point_param = c(1.5, 0.2), # size, alpha, color
  plot_limits = c(-0.65, 0.65), 
  axis_text_size = 4,
  is_category = TRUE) +
  interior_annotation(label = "e2", cex = 2) +
  scale_color_manual(values = c("#8dd3c7", "#fdb462", "#bebada"))

## Second projection
proj_obj2 <- read_rds("data/pbmc3k/pbmc_tsne_proj_obj2.rds")

pbmc_proj_tsne_model2_best <- plot_proj(
  proj_obj = proj_obj2,
  point_param = c(1.5, 0.2), # size, alpha, color
  plot_limits = c(-1, 0.75), 
  axis_text_size = 4,
  is_category = TRUE) +
  interior_annotation(label = "e3", cex = 2) + 
  scale_color_manual(values = c("#8dd3c7", "#fdb462", "#bebada"))
```

```{r}
#| echo: false
#| fig-cap: "Compare the published \\gD{} layout (a) made with UMAP and the \\gD{} layout selected by HBE plot (e) made by tSNE. The two plots on the right show  projections from a tour, with the models overlaid. The published layout a suggested three very separated clusters, but this is not present in the data. While there may be three clusters, they are not well-separated. The difference in model fit also indicates this: the published layout a does not spread out fully into the point cloud like the model generated from layout e. This supports the choice that layout e is the better representation of the data, because it does not exaggerate separation between clusters."
#| label: fig-model-pbmc-author-proj
#| fig-width: 10
#| fig-height: 15
#| fig-pos: "!ht"
#| out-width: 90%

free(trimesh_removed_pbmc) + free(trimesh_removed_pbmc_best) +
  pbmc_proj_umap_model1 + pbmc_proj_tsne_model1_best +
  pbmc_proj_umap_model2 + pbmc_proj_tsne_model2_best +
  plot_layout(nrow=2, byrow=FALSE) &
  theme(legend.position='none')
```

### MNIST hand-written digits {#sec-mnist}

The digit "1" of the MNIST dataset [@lecun1998] consists of $7877$ grayscale images of handwritten "1"s. Each image is $28 \times 28$ pixels, which corresponds to $784$ variables. The first $10$ principal components, explaining $83\%$ of the total variation, are used. This data essentially lies on a nonlinear manifold in high dimensions, defined by the shapes that "1"s make when sketched. We expect that the best layout captures this type of structure and does not exhibit distinct clusters. 

```{r}
#| label: read-mnist-nldr
# Read a variety of different NLDR representations of MNIST
# and plot them on same aspect ratio
tsne_mnist <- read_rds("data/mnist/mnist_tsne30.rds")

nldr_mnist1 <- tsne_mnist |>
  ggplot(aes(x = emb1,
             y = emb2))+
  geom_point(alpha=0.1, size=1, colour='#984ea3') +
  interior_annotation("a", c(0.08, 0.9)) +
  theme(aspect.ratio = 1)

tsne_mnist2 <- read_rds("data/mnist/mnist_tsne89.rds")

nldr_mnist6 <- tsne_mnist2 |>
  ggplot(aes(x = emb1,
             y = emb2))+
  geom_point(alpha=0.1, size=1, colour='#999999') +
  interior_annotation("f", c(0.08, 0.9)) +
  theme(aspect.ratio = 1)

umap_mnist <- read_rds("data/mnist/mnist_umap.rds")

nldr_mnist2 <- umap_mnist |>
  ggplot(aes(x = emb1,
             y = emb2)) +
  geom_point(alpha=0.1, size=1, colour='#a65628') +
  interior_annotation("b") +
  theme(aspect.ratio = 1)

phate_mnist <- read_rds("data/mnist/mnist_phate.rds")

nldr_mnist3 <- phate_mnist |>
  ggplot(aes(x = emb1,
             y = emb2))+
  geom_point(alpha=0.1, size=1, colour='#377eb8') +
  interior_annotation("c", c(0.94, 0.92)) +
  theme(aspect.ratio = 1)

trimap_mnist <- read_rds("data/mnist/mnist_trimap.rds")

nldr_mnist4 <- trimap_mnist |>
  ggplot(aes(x = emb1,
             y = emb2))+
  geom_point(alpha=0.1, size=1, colour='#ff7f00') +
  interior_annotation("d", position = c(0.86, 0.92)) +
  theme(aspect.ratio = 1)

pacmap_mnist <- read_rds("data/mnist/mnist_pacmap.rds")

nldr_mnist5 <- pacmap_mnist |>
  ggplot(aes(x = emb1,
             y = emb2))+
  geom_point(alpha=0.1, size=1, colour='#e41a1c') +
  interior_annotation("e", c(0.08, 0.9)) +
  theme(aspect.ratio = 1)
```

<!--mnist/03_gen_mse_for_diff_methods.R-->
```{r}
#| label: combine-data-mnist

error_mnist_umap <- read_rds("data/mnist/error_mnist_umap.rds")
error_mnist_tsne <- read_rds("data/mnist/error_mnist_tsne.rds")
error_mnist_tsne2 <- read_rds("data/mnist/error_mnist_tsne2.rds")
error_mnist_phate <- read_rds("data/mnist/error_mnist_phate.rds")
error_mnist_trimap <- read_rds("data/mnist/error_mnist_trimap.rds")
error_mnist_pacmap <- read_rds("data/mnist/error_mnist_pacmap.rds")

error_mnist <- bind_rows(error_mnist_umap, 
                        error_mnist_tsne,
                        error_mnist_tsne2,
                        error_mnist_phate,
                        error_mnist_trimap,
                        error_mnist_pacmap)

error_mnist <- error_mnist |>
  mutate(a1 = round(a1, 2)) |>
  filter(b1 >= 5) |>
  group_by(method, a1) |>
  filter(HBE == min(HBE)) |>
  ungroup() |>
  mutate(method = factor(method,
                         levels = c("UMAP", "tSNE", "tSNE2", "PHATE", "TriMAP", "PaCMAP")))
```

```{r}
#| label: error-comp-mnist

error_plot_mnist <- plot_hbe(error_mnist) +
  scale_x_continuous(breaks = seq(0, 0.25, 0.05)) +
                       #sort(
    #unique(error_mnist$a1))[append(seq(1, length(unique(error_mnist$a1)), 
 #                                      by = 5), 24)]) +
  scale_color_manual(
    values=c('#a65628', '#984ea3', '#999999', 
             '#377eb8', '#ff7f00', '#e41a1c')) +
  theme(aspect.ratio=1.5) +
  ylab("HBE")
```


```{r}
#| label: fig-mnist-rmse
#| fig-cap: "Assessing which of the 6 NLDR layouts of the MNIST digit 1 data is the better representation using HBE for varying (i) binwidth ($a_1$), and (ii) average bin count ($\\bar{n}_h$). Color is used for the lines and points in the left plot to match the scatterplots of the NLDR layouts (a-f). Layout c is universally poor. Layouts a and f that show a big cluster and a small circular cluster are universally optimal. Layout a performs well at tiny binwidth (where most points are in their own bin) and not as well as f with larger binwidth, thus layout f is the best choice. Plot (ii), which accounts for the density within clusters by using average bin count, shows reduced differences between layouts, indicating that part of the variation in (i) is driven by cluster density rather than true structural differences."
#| fig-pos: "!ht"
#| out-width: 92%

error_plot_mnist + wrap_plots(nldr_mnist1, nldr_mnist2, nldr_mnist3, 
                                    nldr_mnist4, nldr_mnist5, nldr_mnist6, ncol = 2)
```

@fig-mnist-rmse compares the fit of six layouts computed using UMAP (b), PHATE (c), TriMAP (d), PaCMAP (e) with default hyper-parameter setting and two tSNE runs, one with default hyper-parameter setting (a) and the other changing perplexity to $89$ (f). The layouts are reasonably similar in that they all have the observations in a single blob. Some (b, c) have a more curved shape than others. Layout e is the most different, having a linear shape and a single very large outlier. Both a and f have a small clump of points, perhaps slightly disconnected from the other points, in the lower to middle right. 

The layout plots are colored to match the lines in the HBE vs binwidth ($a_1$) plot. Layouts a, b, and f fit the data better than c, d, e, and layout f appears to be the best fit. @fig-clust-mnist shows this model in the data space in two projections from a tour. The data is curved in the $10\text{-}D$ space, and the fitted model captures this curve. The small clump of points in the \gD{} layout is highlighted in both displays. These are almost all inside the curve of the bulk of points and are sparsely located. The fact that they are packed together in the \gD{} layout is likely due to the handling of density differences by the NLDR. 

An interesting aside is that the rather strange layout e, which has what looks like a single point far from the remaining observations, is actually similar to this one. That point is actually a clump of points corresponding to some of the diffuse points interior to the curve of the bulk of points. This is easy to see using the linked brushing tool.

The next step is to investigate the \gD{} layout to understand what information is learned from this representation. @fig-model-error-mnist summarizes this investigation. Plot a shows the layout with points colored by their residual value - darker color indicates larger residual and poor fit. The plots b, c, d, e show samples of hand-written digits taken from inside the colored boxes. Going from top to bottom around the curve shape, we can see that the "1"s are drawn from right slant to a left slant. The "1"s in d (black box) tend to have the extra up stroke, but are quite varied in appearance. The "1"s shown in the plots labelled e correspond to points with big residuals. They can be seen to be more strangely drawn than the others. Overall, this \gD{} layout shows a useful way to summarize the variation in ways "1"s are drawn.

<!--PaCMAP param: n_components=2, n_neighbors=10, init=random, MN_ratio=0.9, FP_ratio=2.0-->

<!-- Fit the model and compute error-->
<!--mnist/04_gen_model_with_tSNE.R-->
```{r}
#| label: hexbin-mnist

sc_ltr_pos_mnist <- c(0.96, 0.96)

tsne_minst_scaled <- read_rds("data/mnist/mnist_tsne_minst_scaled.rds")

tsne_plot_mnist_clust <- ggplot() + 
  geom_point(data = tsne_minst_scaled,
             aes(
               x = emb1,
               y = emb2,
               colour = cluster
             ),
             alpha=0.2)+
  scale_color_manual(values=c('#d9d9d9', '#ff7f00')) +
  interior_annotation("a", c(0.08, 0.93), cex = 2.5) +
  theme(legend.position = "none") 

## Compute error
error_df <- read_rds("data/mnist/mnist_error_df.rds")
  
highd_error_2d_mnist <- error_df |>
  ggplot(aes(x = emb1,
             y = emb2,
             colour = sqrt_row_wise_tot_error)) +
  geom_point(alpha=0.2) +
  scale_colour_continuous_sequential(palette = "YlOrRd", n_interp = 20) +
  geom_rect(aes(xmin = 0.82, xmax = 0.92, ymin = 0.38, ymax = 0.58),
               fill = "transparent", colour = "#000000", linewidth = 1) +
  geom_rect(aes(xmin = 0.9, xmax = 1, ymin = 1.25, ymax = 1.45),
               fill = "transparent", colour = "#375cb1", linewidth = 1)  +
  geom_rect(aes(xmin = 0.35, xmax = 0.45, ymin = 0.55, ymax = 0.7),
               fill = "transparent", colour = "#008000", linewidth = 1)  +
  geom_rect(aes(xmin = 0.7, xmax = 0.8, ymin = 0.25, ymax = 0.4),
               fill = "transparent", colour = "#4c389a", linewidth = 1) +
  interior_annotation(label = "a", c(0.08, 0.93), cex = 2) +
  theme(legend.position = "none",
        aspect.ratio = 1,
        plot.margin = margin(0, 0, 0, 0),
        panel.spacing = unit(0, "pt"),
        axis.ticks.length = unit(0, "pt"),
        strip.text = element_blank(),   # Optional, if you're using facets
        axis.title = element_blank()   # Optional
        )
```

```{r}
#| label: mnist-model-proj

proj_obj1 <- read_rds("data/mnist/mnist_proj_obj1.rds")

mnist_proj_tsne_model1_link <- plot_proj(
    proj_obj = proj_obj1,
    point_param = c(1.2, 0.5), # size, alpha, color
    line_param = c(0.5, 0.5, "#000000"), # linewidth, alpha
    plot_limits = c(-1.1, 0.9), 
    axis_text_size = 5, 
    is_category = TRUE) +
  interior_annotation(label = "b", cex = 2.5) +
  scale_color_manual(values = c("#d9d9d9",'#ff7f00')) +
  theme(
      legend.position = "none"
  )

## Second projection
proj_obj2 <- read_rds("data/mnist/mnist_proj_obj2.rds")

mnist_proj_tsne_model2_link <- plot_proj(
    proj_obj = proj_obj2, 
    point_param = c(1.2, 0.5), # size, alpha, color
    line_param = c(0.5, 0.5, "#000000"), # linewidth, alpha
    plot_limits = c(-1, 1), 
    axis_text_size = 5, 
    is_category = TRUE) +
  interior_annotation(label = "c", cex = 2.5) +
  scale_color_manual(values = c("#d9d9d9",'#ff7f00')) +
  theme(
      legend.position = "none"
  )

```

```{r}
#| label: fig-clust-mnist
#| fig-cap: "Summary from exploring tSNE layout of the MNIST digit 1 data (a in Figure 12) using linked brushing. There is a big nonlinear cluster (grey) and a small cluster (orange) located very close to one corner of the big cluster in \\gD{} (a). The MNIST digit 1 data has a nonlinear structure in $10\\text{-}D$. Two \\gD{} projections from a tour on $10\\text{-}D$ reveal that the small orange cluster is actually a diffuse set of points wrapped within the grey cluster, which is C-shaped in the high dimensions."
#| fig-height: 5
#| fig-width: 15
#| fig-pos: "!ht"
#| out-width: 100%

tsne_plot_mnist_clust + mnist_proj_tsne_model1_link + mnist_proj_tsne_model2_link +
  plot_layout(ncol = 3)
```

```{r}
#| label: img-diff-pos

## Data with pixel values

img_right_top <- read_rds("data/mnist/mnist_img_right_top.rds")

right_top_img <- plot_digit_img(
  img_right_top, color_on = "#375cb1", title_text = "b")

img_middle <- read_rds("data/mnist/mnist_img_middle.rds")

middle_img <- plot_digit_img(
  img_middle, color_on = "#008000", title_text = "c")

img_right_bottom <- read_rds("data/mnist/mnist_img_right_bottom.rds")

right_bottom_img <- plot_digit_img(
  img_right_bottom, color_on = "#4c389a", title_text = "d")

img_error_outside <- read_rds("data/mnist/mnist_img_error_outside.rds")

imge_error_sample_outside <- plot_digit_img(
  img_error_outside, color_on = "#000000", title_text = "e")

##### inside
img_error_inside <- read_rds("data/mnist/mnist_img_error_inside.rds")

imge_error_sample_inside <- plot_digit_img(
  img_error_inside, color_on = "#d95f0e", title_text = "f")

```

```{r}
#| label: fig-model-error-mnist
#| fig-cap: "Summary of the layout structure, and large errors, relative to the MNIST digit 1 shape: (a) layout colored by residual value, and at right (b-e) are images of samples of observations taken at locations around the layout, showing similarity in how the 1's were drawn. Set (f) are images corresponding to large residuals in the big cluster (darker orange in plot a). Along the big cluster, the angle of digit 1 changes (b-d). The small cluster has larger residuals, and the images show that these tend to be European style with a flag at the top and a base at the bottom. The set in (f) shows various poorly written digits."
#| fig-width: 10
#| fig-height: 10
#| fig-pos: "!ht"
#| out-width: 100%

# Define your current layout
top_layout <- right_top_img + middle_img + right_bottom_img + imge_error_sample_outside + imge_error_sample_inside + plot_layout(nrow = 5)

final_layout <- wrap_plots(highd_error_2d_mnist, top_layout)

final_layout
```

## Discussion {#sec-discussion}

We have developed an approach to help assess and compare NLDR layouts, generated by different methods and hyper-parameter choice(s). It depends on conceptualizing the \gD{} layout as a model, allowing for the creation of a wireframe representation of the model that can be lifted into \pD{}. The fit is assessed by viewing the model in the data space, computing residuals, and HBE. Different layouts can be compared using the HBE, providing a quantitative metric to decide on the most suitable NLDR layout to represent the \pD{} data. Global and local preservation of structure is assessed by examining the HBE across a range of binwidths. It also provides a way to predict the values of new \pD{} observations in the \gD{}, which could be useful for implementing uncertainty checks, such as using training and testing samples.

The new methodology is accompanied by an R package called `quollr`, so that it is readily usable and broadly accessible. The package has methods to fit the model, compute diagnostics, and also visualize the results, with interactivity. We have primarily used the `langevitour` software [@harisson2024] to view the model in the data space, but other tour software such as `tourr` [@wickham2011] and `detourr` [@hart2022] could also be used.

Two examples illustrating usage are provided: the PBMC3k data, where the NLDR is summarizing clustering in \pD{}, and hand-written digits illustrating how NLDR represents an intrinsically low-dimensional nonlinear manifold. We examined a typical published usage of UMAP with the PBMC3k dataset [@chen2024]. As is typical of UMAP layout with default settings, the separation between clusters is grossly exaggerated. The layout even suggests separation where there is none.  Our approach provides a way to choose a reasonable layout and avoids the use of misleading layouts in the future. In the hand-written digits [@lecun1998], we illustrate how our model fit statistics show that a flat disc layout is superior to the curved-shaped layouts, and how to identify oddly written "1"s using the residuals of the fitted model.

This work can be applied with existing metrics for evaluating NLDR layout, such as ARNX, RTA, SC, and RGS. It provides an additional evaluation metric, and importantly allows any layout to be viewed in the \pD{} data space. This latter aspect can help to disentangle conflicting suggestions by the different metrics. 

Additional exploration of distance measures to summarize the fit could be a valuable direction for future work. We have used Euclidean distance, but other measures, such as geodesic distances [@joshua2000], may better capture curved or nonlinear relationships in the data and are worth exploring.

This work has also revealed some interesting behaviors of NLDR methods, including twisting, flattened "pancake" clusters in \pD{}, and severe effects of density differences. These are described in more detail in the supplementary materials.

Researchers usually use \gD{} layouts, but if a \kD{} ($k>2$) layout is provided, the approach developed here could be extended. Potential approaches include $3\text{-}D$ binning, $k$-means clustering, or even special implementations of convex hulls. 

## Supplementary Materials {#sec-supplementary}

All the materials to reproduce the paper can be found at [https://github.com/JayaniLakshika/paper-nldr-vis-algorithm](https://github.com/JayaniLakshika/paper-nldr-vis-algorithm). 

The supplementary materials provide additional details on the methods and hyper-parameters used to generate layouts, video links of animated \pD{} tours, notation summaries, and the R and Python scripts used in the study. They also describe the generation of the 2NC7 data, computation of hexagon grid configurations, and data binning procedures. Further sections highlight interesting NLDR behaviors observed in the data space and compare HBE with existing evaluation metrics for the PBMC3k and MNIST datasets.

The R package `quollr`, available on CRAN and at [https://jayanilakshika.github.io/quollr/](https://jayanilakshika.github.io/quollr/), provides software accompanying this paper to fit the wireframe model representation, compute diagnostics, visualize the model in the data with `langevitour`, and link multiple plots interactively. 


## Acknowledgments

These `R` packages were used for the work: `tidyverse` [@hadley2019], `Rtsne` [@jesse2015], `umap` [@tomasz2023], `patchwork` [@thomas2024], `colorspace` [@achim2020], `langevitour` [@harisson2024], `conflicted` [@hadley2023], `reticulate` [@kevin2024], `kableExtra` [@hao2024]. These `python` packages were used for the work: `trimap` [@amid2022] and `pacmap` [@yingfan2021]. The article was created with `R` packages `quarto` [@jjallaire2024]. 


## References {.unnumbered}
  
::: {#refs}
:::
      
