---
title: "Supplementary Material 6: Hexagon Grid: Configuration, Binning, and Area"
format: pdf
tbl-cap-location: bottom
header-includes: | 
  \usepackage{amsmath}
  \usepackage{float}
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \usepackage{bm}
  \def\tightlist{}
  \usepackage{setspace}
  \newcommand\pD{$p\text{-}D$}
  \newcommand\kD{$k\text{-}D$}
  \newcommand\dD{$d\text{-}D$}
  \newcommand\gD{$2\text{-}D$}
  \newcommand\tD{$3\text{-}D$}
  \newcommand\fD{$4\text{-}D$}
  \newcommand\sD{$7\text{-}D$}
  \newcommand{\To}{\textbf{to}~}
  \usepackage{abstract}
  \renewcommand{\abstractname}{}    % clear the title
  \renewcommand{\absnamepos}{empty} % originally center
---

```{r include=FALSE}
# Set up chunk for for knitr
knitr::opts_chunk$set(
  fig.width = 5,
  fig.height = 5,
  fig.align = "center",
  out.width = "100%",
  code.line.numbers = FALSE,
  fig.retina = 4,
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  cache = FALSE,
  dev.args = list(pointsize = 11)
)
```

```{r include=FALSE}
#| label: install-libraries
#| warning: false
#| echo: false

options(repos = c(CRAN = "https://cran.rstudio.com")) # Setup mirror

packages_to_check <- c("remotes", "tidyverse", "patchwork", "colorspace", "kableExtra", "conflicted", "ggforce")

for (pkg in packages_to_check) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    message(paste("Installing package:", pkg))
    install.packages(pkg)
  } else {
    installed_version <- packageVersion(pkg)
    available_version <- tryCatch({
      utils::packageDescription(pkg)$Version
    }, error = function(e) NA) # Handle cases where package info isn't readily available

    if (!is.na(available_version) && installed_version < package_version(available_version)) {
      message(paste("A newer version of package", pkg, "is available. Updating..."))
      install.packages(pkg)
    } else {
      message(paste("Package", pkg, "is up-to-date (version", installed_version, ")."))
    }
  }
}

```

```{r}
#| label: load-libraries
#| warning: false
#| echo: false
library(tibble)
library(dplyr)
library(ggplot2)
library(readr)
library(tidyr)
library(patchwork)
library(colorspace)
library(kableExtra)
library(conflicted)
library(ggforce)
library(GGally)
library(scales)

conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::select)
```

```{r}
#| label: plot-theme
theme_set(theme_linedraw() +
   theme(
     #aspect.ratio = 1,
     plot.background = element_rect(fill = 'transparent', colour = NA),
     plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
     panel.background = element_rect(fill = 'transparent', 
                                     colour = NA),
     panel.grid.major = element_blank(), 
     panel.grid.minor = element_blank(), 
     axis.title.x = element_blank(), axis.title.y = element_blank(),
     axis.text.x = element_blank(), axis.ticks.x = element_blank(),
     axis.text.y = element_blank(), axis.ticks.y = element_blank(),
     legend.background = element_rect(fill = 'transparent', 
                                      colour = NA),
     legend.key = element_rect(fill = 'transparent', 
                               colour = NA),
     legend.position = "bottom", 
     legend.title = element_blank(), 
     legend.text = element_text(size=4),
     legend.key.height = unit(0.25, 'cm'),
     legend.key.width = unit(0.25, 'cm'),
     plot.margin = margin(0, 0, 0, 0)
   )
)
```

```{r}
#| label: import-scripts
source(here::here("scripts/additional_functions.R"))
```

```{r}
#| label: code-setup
set.seed(20240110)
```

```{r}
clr_choice <- "#0077A3"
```

# Computing hexagon grid configurations

Given range of embedding component, $r_2$, number of bins along the x-axis, $b_1$, and buffer proportion, $q$, hexagonal starting point coordinates, $s_1 = \text{-}q$, and $s_2 = \text{-}qr_2$. The purpose is to find width of the hexagon, $a_1$ and number of bins along the y-axis, $b_2$.

Geometric arguments give rise to the following constraints.

$\text{min }a_1 \text{ s.t.}$

$$
s_1 - \frac{a_1}{2} < 0,
$$ {#eq-equation1}

$$
s_1 + (b_1 - 1) \times a_1 \geq 1,
$$ {#eq-equation2}

$$
s_2 - \frac{a_2}{2} < 0,
$$ {#eq-equation4}

$$
s_2 + (b_2 - 1) \times a_2 \geq r_2.
$$ {#eq-equation5}




```{r}
#| label: code-illustration
# Code to draw illustration for notation
## hexagon binning to have regular hexagons

hex_grid_temp <- read_rds(here::here("data/two_nonlinear/hex_grid_temp.rds"))
all_centroids_df_temp <- read_rds(here::here("data/two_nonlinear/all_centroids_df_tempp.rds"))
start_pt <- read_rds(here::here("data/two_nonlinear/start_pt.rds"))
d_rect <- read_rds(here::here("data/two_nonlinear/d_rect.rds"))
a1 <- read_rds(here::here("data/two_nonlinear/a2_data.rds"))
a2 <- read_rds(here::here("data/two_nonlinear/a1_data.rds"))
l <- read_rds(here::here("data/two_nonlinear/l_data.rds"))
rect_adj <- read_rds(here::here("data/two_nonlinear/rect_adj.rds"))

hex_param_vis <- ggplot() + 
  geom_polygon(data = hex_grid_temp, 
               aes(x = x, 
                   y = y, 
                   group = h),
               fill = "white", 
               colour = "#bdbdbd") +
  geom_point(data = all_centroids_df_temp, aes(
    x = c_x, 
    y = c_y), 
    colour = "#31a354", size = 0.9) +
  geom_point(data = start_pt, aes(x = c_x, 
                                  y = c_y), 
             colour = "black") + 
  geom_rect(data=d_rect, 
            aes(xmin = x1min - rect_adj$x1,# - rect_adj$s1, 
                xmax = x1max - rect_adj$x1,# - rect_adj$s1, 
                ymin = x2min - rect_adj$x2,# - rect_adj$s2, 
                ymax = x2max - rect_adj$x2),# - rect_adj$s2), 
            fill = "white", 
            colour = "black", 
            alpha = 0, 
            linewidth = 0.7) +
  geom_point(data=d_rect, aes(x=x1min - rect_adj$x1, 
                              y=x2min - rect_adj$x2)) + 
  geom_point(data=d_rect, aes(x=x1max - rect_adj$x1, 
                              y=x2min - rect_adj$x2)) + 
  geom_point(data=d_rect, aes(x=x1min - rect_adj$x1, 
                              y=x2max - rect_adj$x2)) + 
  annotate("text", x=d_rect$x1min - rect_adj$x1, 
           y=d_rect$x2min - rect_adj$x2,
           label = "(0,0)", 
           hjust=-0.1, vjust=-0.3, size = 8) + 
  annotate("text", x=d_rect$x1max - rect_adj$x1, 
           y=d_rect$x2min - rect_adj$x2,
           label = "(0,1)", 
           hjust=1.1, vjust=-0.3, size = 8) + 
  annotate("text", x=d_rect$x1min - rect_adj$x1, 
           y=d_rect$x2max - rect_adj$x2,
           label = "'(0,'*y[2][max]*')'", 
           parse = TRUE,
           hjust=-0.1, vjust=1.2, size = 8) + 
  geom_segment(data=d_rect, aes(
    x = x1min  - rect_adj$x1, # 0 - 0.03, 
    y = -0.31, 
    xend = x1max - rect_adj$x1, #1 - 0.03, 
    yend = -0.31), #-0.35),
    arrow = arrow(length = unit(0.03, "npc"),
                  ends = "both"), 
    colour = "black")+
  annotate("text", x=0.5, y=-0.38, 
           label = "r[1]", parse = TRUE, colour = "black", size = 8) +
  geom_segment(data=d_rect, aes(
    x = -0.25, 
    y = x2min - rect_adj$x2, #0 - 0.05, 
    xend = -0.25, 
    yend = x2max - rect_adj$x2), #r2 - 0.05),
    arrow = arrow(length = unit(0.03, "npc"),
                  ends = "both"), 
    colour = "black")+ 
  annotate("text", x=-0.3, y=0.4, 
           label = "r[2]", parse = TRUE, colour = "black", size = 8) +
  geom_segment(data = a1, aes(
    x = x, #-0.1 + 0.2087578, 
    y = y, #-0.15, 
    xend = xend, #-0.1 + 0.2087578*2, 
    yend = yend), #-0.15),
    arrow = arrow(length = unit(0.03, "npc"),
                  ends = "both"), 
    colour = "black")+ # a1 = 0.2087578
  annotate("text", 
           x=(a1$x+a1$xend)/2, 
           y=a1$y, 
           label = "a[1]", parse = TRUE, 
           colour = "black",
           vjust = 1.2, size = 8) +
  geom_segment(data = a2, aes(
    x = x, #-0.15, 
    y = y, #-0.1*r2 + 0.1807896*2, 
    xend = xend, #-0.15, 
    yend = yend), #-0.1*r2 + 0.1807896*3),
    arrow = arrow(length = unit(0.03, "npc"),
                  ends = "both"), 
    colour = "black") + # a2 = 0.1807896
  annotate("text", x=a2$x, y=(a2$y+a2$yend)/2, 
           label = "a[2]", parse = TRUE,
           colour = "black", hjust=-0.2, size = 8) +
  annotate("text", x=-0.18, y=-0.24, 
           label = "'('*s[1]*','*s[2]*')'", parse = TRUE,
           colour = "black", size = 8) +
  geom_segment(data = l, aes(
    x = x, #-0.15, 
    y = y, #-0.1*r2 + 0.1807896*2, 
    xend = xend, #-0.15, 
    yend = yend), #-0.1*r2 + 0.1807896*3),
    arrow = arrow(length = unit(0.03, "npc"),
                  ends = "both"), 
    colour = "black") + 
  annotate("text", x=l$x + 0.03, y=(l$y+l$yend)/2, 
           label = "l", parse = TRUE,
           colour = "black", hjust=-0.2, size = 8) +
  coord_equal()
```

```{r}
#| label: fig-hex-param
#| fig-cap: "The components of the hexagon grid illustrating notation."
#| out-width: 30%
#| fig-pos: "!ht"
#| eval: false

hex_param_vis
```

Since $a_1$ and $a_2$ are distances,

$$
a_1, a_2 > 0.
$$ Also, $(s_1, s_2) \in (\text{-}0.1, \text{-}0.05)$ as these are multiplicative offsets in the negative direction.

@eq-equation1 can be rearranged as,

$$
a_1 > 2s_1
$$

which given $s_1 < 0$ and $a_1 > 0$ will *always* be true. The same logic follows for @eq-equation4 and substituting $a_2 = \sqrt{3}a_1/{2}$, and $s_2 = \text{-}qr_2$ to @eq-equation4 can be written as,

$$
a_1 > -\frac{4}{\sqrt{3}}qr_2
$$

Also, substituting $a_2 = \sqrt{3}a_1/{2}$, $s_2 = \text{-}qr_2$ and rearranging @eq-equation5 gives:

$$
a_1 \geq \frac{2(r_2 + qr_2)}{\sqrt{3}(b_2 - 1)}.
$$ {#eq-equation6}

Similarly, substituting $s_1 = \text{-}q$ @eq-equation2 becomes,

$$
a_1 \geq \frac{(1 + q)}{(b_1 - 1)}.
$$ {#eq-equation7}

This is a linear optimization problem. Therefore, the optimal solution must occur on a vertex. So, by setting @eq-equation6 equals to @eq-equation7 gives,

$$
\frac{2(r_2 + qr_2)}{\sqrt{3}(b_2 - 1)} = \frac{(1 + q)}{(b_1 - 1)}.
$$ 

After rearranging this,

$$
b_2 = 1 + \frac{2r_2(b_1 - 1)}{\sqrt{3}}
$$

and since $b_2$ should be an integer,

$$
b_2 = \Big\lceil1 +\frac{2r_2(b_1 - 1)}{\sqrt{3}}\Big\rceil.
$$ {#eq-equation8}

Furthermore, with known $b_1$ and $b_2$, by considering @eq-equation2 or @eq-equation5 as the *binding* or *active constraint*, can compute $a_1$.

If @eq-equation2 is active, then,

$$
\frac{(1 + q)}{(b_1 - 1)} < \frac{2(r_2 + qr_2)}{\sqrt{3}(b_2 - 1)}.
$$

Rearranging this gives,

$$
r_2 > \frac{\sqrt{3}(b_2 - 1)}{2(b_1 - 1)}.
$$

Therefore, if this equality is true, then 
$$
a_1 = \frac{(1+q)}{(b_1 - 1)},
$$
otherwise, 
$$
a_1 = \frac{2r_2(1+q)}{\sqrt{3}(b_2 - 1)}.
$$

# Binning the data

Points are assigned to the bin they fall into based on the nearest centroid. If a point is equidistant from multiple centroids, it is assigned to the centroid with the smallest bin ID.

```{r}

## Data set with all centroids
all_centroids_df <- read_rds(here::here("data/two_nonlinear/two_nonlinear_all_centroids_two_curvy4.rds"))

## Generate all coordinates of hexagons
hex_grid <- read_rds(here::here("data/two_nonlinear/two_nonlinear_hex_grid_two_curvy4.rds"))

## Data to generate circle
circles <- data.frame(
  x = c(0.608, 0.372),
  y = c(0.306, 0.715),
  r = rep(0.27, 2)
)

emb_df1 <- tibble(emb1 = 0.45, emb2 = 0.47) |>
  add_row(emb1 = 0.54, emb2 = 0.56)

assign_data_vis1 <- ggplot(data = hex_grid, aes(x = x, y = y)) + 
  geom_polygon(fill = "white", color = "#bdbdbd", aes(group = h)) +
  geom_point(data = emb_df1, aes(x = emb1, y = emb2), color = "black", size = 0.7) +
  geom_text(data = all_centroids_df, aes(x = c_x, y = c_y, label = h), size = 5, colour = "#bdbdbd") +
  geom_circle(data = circles, aes(x0 = x, y0 = y, r = r), fill = NA) +
  geom_segment(x = 0.45, y = 0.47, xend = 0.58, yend = 0.3, 
               arrow = arrow(length = unit(0.02, "npc")), color = "#969696") +
  geom_segment(x = 0.54, y = 0.56, xend = 0.4, yend = 0.715, 
               arrow = arrow(length = unit(0.02, "npc")), color = "#969696")
```

```{r}
hex_data <- hex_grid |> 
  dplyr::filter(h == 8)

# Create all pairs of rows
pairs <- expand_grid(row1 = 1:nrow(hex_data), row2 = 1:nrow(hex_data)) |>
  filter(row1 < row2)

# Compute middle points for each pair of rows
middle_points <- pairs |>
  rowwise() |>
  mutate(
    h = hex_data$h[row1],
    x = (hex_data$x[row1] + hex_data$x[row2]) / 2,
    y = (hex_data$y[row1] + hex_data$y[row2]) / 2
  ) |>
  ungroup() |>
  select(h, x, y) |>
  filter(row_number() %in% c(1, 5, 6, 10, 13, 15))

points_df <- bind_rows(hex_data, middle_points)

arrow_df <- tibble(
  x = points_df$x, 
  y = points_df$y,
  xend = c(0.37, points_df$x[2] - 0.1, 0.136, points_df$x[4] - 0.1, 0.61, points_df$x[6] - 0.1, 0.33, points_df$x[8] - 0.1, 0,  points_df$x[10] - 0.1, 0.58, 0.45),
  yend = c(0.82, points_df$y[2] - 0.1, 0.4, points_df$y[4] - 0.1, 0.4, points_df$y[6] - 0.1, 0.82, points_df$y[8] - 0.1, 0.715, points_df$y[10] - 0.1, 0.4, 0.71)
)

assign_data_vis2 <- ggplot(data = hex_grid, aes(x = x, y = y)) + 
  geom_polygon(fill = "white", color = "#bdbdbd", aes(group = h)) +
  geom_point(data = points_df, aes(x = x, y = y), color = "black", size = 0.7) +
  geom_text(data = all_centroids_df, aes(x = c_x, y = c_y, label = h), size = 5, colour = "#bdbdbd") +
  geom_segment(data = arrow_df, aes(x = x, y = y, xend = xend, yend = yend), 
               arrow = arrow(length = unit(0.02, "npc")), color = "#bdbdbd") 
```

```{r}
#| label: fig-assign-data
#| fig-cap: "Binning the data. Points are assigned to the nearest centroid. If a point is equidistant from multiple centroids, assigned to the centroid with the smallest bin ID."
#| fig-width: 12
#| fig-height: 6
#| fig-pos: "!ht"

assign_data_vis1 + assign_data_vis2 +
  plot_layout(ncol = 2)
```

# Area of a hexagon

The area of a hexagon is defined as $A = 3\sqrt{3}l^2/2$, where $l$ is the side length of the hexagon. $l$ can be computed using $a_1$ and $a_2$.


```{r}
# triangle_temp <- tibble(
#   x = c(0, 4, 0),
#   y = c(0, 0, 3)
# )

triangle_line_temp <- tibble(
  x = c(4, 0, 4),
  xend = c(0, 0, 0),
  y = c(0, 0, 0),
  yend = c(0, 3, 3)
)

# triangle_arrow_temp <- tibble(
#   x = c(4 + 0.04, 0 - 0.06, 4 + 0.06),
#   xend = c(0 + 0.04, 0 - 0.06, 0 + 0.06),
#   y = c(0 - 0.04, 0 - 0.06, 0 + 0.06),
#   yend = c(0 - 0.04, 3 - 0.06, 3 + 0.06)
# )

triangle_plot <- ggplot() +
  # geom_point(
  #   data = triangle_temp,
  #   aes(
  #     x = x,
  #     y = y
  #   )
  # ) +
  geom_segment(
    data = triangle_line_temp,
    aes(
      x = x,
      xend = xend,
      y = y,
      yend = yend
    )
  ) + 
  annotate("text", x = -1.5, y = 1.5,
         label = "frac(a[2]-a[1], 2)",
         parse = TRUE,
         color = "black", hjust = -0.2, size = 8) +
annotate("text", x = 1.5, y = -0.5,
         label = "a[1]/2",
         parse = TRUE,
         color = "black", hjust = -0.2, size = 8) +
annotate("text", x = 2.5, y = 1.5,
         label = "l",
         parse = TRUE,
         color = "black", hjust = -0.2, size = 8)
```

```{r}
#| label: fig-tri-param
#| fig-cap: "The components of the right triangle illustrating notation."
#| out-width: 30%
#| fig-pos: "!ht"

triangle_plot
```

By applying the Pythagorean theorem, we obtain,

$$
l^2 = \left(\frac{a_1}{2}\right)^2 + \left(\frac{a_2 - l}{2}\right)^2.
$$
Next, rearranging the terms, we get,

$$
l^2 - \left(\frac{a_2 - l}{2}\right)^2 = \left(\frac{a_1}{2}\right)^2,
$$

$$
\left[l - \left(\frac{a_2 - l}{2}\right)\right]\left[l + \left(\frac{a_2 - l}{2}\right)\right] = \left(\frac{a_1}{2}\right)^2,
$$

$$
3l^2 + 2a_2l - (a_1^2 + a_2^2) = 0.
$$

Finally, by solving the quadratic equation, we compute,

$$
l = \frac{-2a_2 \pm \sqrt{4a_2^2 - 24[-(a_1^2 + a_2^2)]}}{6},
$$

$$
l = \frac{-a_2 \pm \sqrt{a_2^2 - 6[-(a_1^2 + a_2^2)]}}{3},
$$

where $l > 0$.

\clearpage
