---
title: "Supplementary Material 7: Tuning to decide on best binwidth for viewing in $p\\text{-}D$"
format: pdf
tbl-cap-location: bottom
header-includes: | 
  \usepackage{amsmath}
  \usepackage{float}
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \usepackage{bm}
  \def\tightlist{}
  \usepackage{setspace}
  \newcommand\pD{$p\text{-}D$}
  \newcommand\kD{$k\text{-}D$}
  \newcommand\dD{$d\text{-}D$}
  \newcommand\gD{$2\text{-}D$}
  \newcommand\tD{$3\text{-}D$}
  \newcommand\fD{$4\text{-}D$}
  \newcommand\sD{$7\text{-}D$}
  \newcommand{\To}{\textbf{to}~}
  \usepackage{abstract}
  \renewcommand{\abstractname}{}    % clear the title
  \renewcommand{\absnamepos}{empty} % originally center
---

```{r include=FALSE}
# Set up chunk for for knitr
knitr::opts_chunk$set(
  fig.width = 5,
  fig.height = 5,
  fig.align = "center",
  out.width = "100%",
  code.line.numbers = FALSE,
  fig.retina = 4,
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  cache = FALSE,
  dev.args = list(pointsize = 11)
)
```

```{r include=FALSE}
#| label: install-libraries
#| warning: false
#| echo: false

options(repos = c(CRAN = "https://cran.rstudio.com")) # Setup mirror

packages_to_check <- c("remotes", "tidyverse", "patchwork", "colorspace", "kableExtra", "conflicted", "ggforce")

for (pkg in packages_to_check) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    message(paste("Installing package:", pkg))
    install.packages(pkg)
  } else {
    installed_version <- packageVersion(pkg)
    available_version <- tryCatch({
      utils::packageDescription(pkg)$Version
    }, error = function(e) NA) # Handle cases where package info isn't readily available

    if (!is.na(available_version) && installed_version < package_version(available_version)) {
      message(paste("A newer version of package", pkg, "is available. Updating..."))
      install.packages(pkg)
    } else {
      message(paste("Package", pkg, "is up-to-date (version", installed_version, ")."))
    }
  }
}

```

```{r}
#| label: load-libraries
#| warning: false
#| echo: false
library(tibble)
library(dplyr)
library(ggplot2)
library(readr)
library(tidyr)
library(patchwork)
library(colorspace)
library(kableExtra)
library(conflicted)
library(ggforce)
library(GGally)
library(scales)

conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::select)
```

```{r}
#| label: plot-theme
theme_set(theme_linedraw() +
   theme(
     #aspect.ratio = 1,
     plot.background = element_rect(fill = 'transparent', colour = NA),
     plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
     panel.background = element_rect(fill = 'transparent', 
                                     colour = NA),
     panel.grid.major = element_blank(), 
     panel.grid.minor = element_blank(), 
     axis.title.x = element_blank(), axis.title.y = element_blank(),
     axis.text.x = element_blank(), axis.ticks.x = element_blank(),
     axis.text.y = element_blank(), axis.ticks.y = element_blank(),
     legend.background = element_rect(fill = 'transparent', 
                                      colour = NA),
     legend.key = element_rect(fill = 'transparent', 
                               colour = NA),
     legend.position = "bottom", 
     legend.title = element_blank(), 
     legend.text = element_text(size=4),
     legend.key.height = unit(0.25, 'cm'),
     legend.key.width = unit(0.25, 'cm'),
     plot.margin = margin(0, 0, 0, 0)
   )
)
```

```{r}
#| label: import-scripts
source(here::here("scripts/additional_functions.R"))
```

```{r}
#| label: code-setup
set.seed(20240110)
```

```{r}
clr_choice <- "#0077A3"
```

<!-- The model fitting is based on several parameters, including the hexagon bin parameters and the low-count bin removal process. The hexagon bin parameters define the bottom-left bin position $(s_1, \ s_2)$, the number of bins in the horizontal direction ($b_1$), which also determines the number of bins in the vertical direction ($b_2$), the total number of bins ($b$), and the total number of non-empty bins ($m$). Low count bins are removed using standardized bin counts, defined as $w_h = n_h/n, ~~h=1, \dots m$. -->

<!-- Default values are provided for each of these, but deciding on the best model fit is assisted by examining a range of values. The default number of bins $b=b_1\times b_2$ is computed based on the sample size, by setting $b_1=n^{1/3}$, consistent with the Diaconis-Freedman rule [@freedman1981]. The value of $b_2$ is determined analytically by $b_1, q, r_2$ (@eq-bin2). Values of $b_1$ between $2$ and $b_1 = \sqrt{n/r_2}$ are recommended, where the dependence on $r_2$ reflects the preservation of aspect ratio in the NLDR layout. -->

<!-- @fig-bins-two-curvy shows the hexbin grids for three choices of $b_1$. While the number of bins is the common parameter to modify, bin start positions $(s_1, \ s_2)$ can also be worth experimenting with, because they can also change bin counts.  -->


The model fitting depends on several parameters, including the hexagon bin parameters and the low-count bin removal process. The hexagon bin parameters define the bottom-left bin position $(s_1, s_2)$, the number of bins in the horizontal direction ($b_1$), which determines the number of bins in the vertical direction ($b_2$), the total number of bins ($b$), and the number of non-empty bins ($m$). Low-count bins are removed using standardized bin counts, defined as $w_h = n_h/n$, for $h = 1, \dots, m$.

Default values are provided for these parameters. In particular, the default number of bins is determined from the sample size by setting $b_1 = n^{1/3}$, consistent with the Diaconis–Freedman rule [@freedman1981], with $b_2$ computed to preserve the aspect ratio of the layout.

Rather than selecting a single set of parameters a priori, we recommend a structured procedure to examine the stability of the fitted model across a range of values. A practical workflow is as follows:

1. **Select a grid of binwidths:** Choose a sequence of values for $b_1$ ranging from coarse to fine resolution (e.g., from small values up to $b_1 = \sqrt{n/r_2}$). This controls the level of detail in the fitted wireframe.

2. **Compute HBE across the grid:** For each choice of $b_1$, compute the HBE. Stable regions where HBE changes slowly indicate parameter values where the fitted model is not overly sensitive to binning resolution.

3. **Assess bin occupancy:** Examine summary quantities such as average bin count ($\bar{n}$), average standardized bin count ($\bar{w}$), and the proportion of non-empty bins ($m/b$).
   Reasonable parameter choices typically correspond to:

   - sufficiently large $\bar{n}$ to avoid sparse bins, and
   
   - moderate values of $m/b$, indicating that bins cover the data without excessive fragmentation.

4. **Check robustness to bin removal:** Evaluate the effect of removing low-count bins using a range of thresholds. Parameter values where HBE remains stable under small amounts of bin removal indicate robustness of the fitted structure.

5. **Compare across layouts (if applicable):** When comparing multiple NLDR layouts, examine HBE as a function of average bin count or standardized bin count. This helps account for differences in point density between layouts.

This process does not yield a single "optimal" parameter choice, but instead identifies a range of stable and interpretable settings. The goal is to avoid parameter values that produce unstable fits, overly sparse binning, or misleading structure in the fitted wireframe.

```{r}
sc_ltr_pos <- c(0.08, 0.9)
sc_xlims <- c(-0.3, 1.2)
sc_ylims <- c(-0.17, 1.25)
```

```{r}
#| label: hexbin-regular-two-curvy1
hex_grid_with_counts <- read_rds(here::here("data/two_nonlinear/hex_grid_with_counts.rds"))

tsne_two_curvy_scaled <- read_rds(here::here("data/two_nonlinear/tsne_two_curvy_scaled.rds"))

hex_grid_coloured_two_curvy1 <- ggplot() + 
  geom_polygon(
    data = hex_grid_with_counts, 
    aes(x = x, y = y,
      group = h, 
      fill = w_h), colour = "grey70", linewidth=0.2) +
  geom_point(data = tsne_two_curvy_scaled,
             aes(x = emb1, y = emb2),
             colour = clr_choice,
             alpha = 0.3,
             size = 0.5) +
  scale_fill_viridis_c(direction = -1, 
    na.value = "#ffffff", option = "E") +
  interior_annotation("b", sc_ltr_pos, cex = 2) + 
  xlim(sc_xlims) + ylim(sc_ylims)
```

```{r}
#| label: hexbin-regular-two-curvy2

hex_grid_with_counts_two_curvy2 <- read_rds(here::here("data/two_nonlinear/two_nonlinear_hex_grid_with_counts_two_curvy2.rds"))
tr_from_to_df_two_curvy2 <- read_rds(here::here("data/two_nonlinear/two_nonlinear_tr_from_to_df_two_curvy2.rds"))

hex_grid_coloured_two_curvy2 <- ggplot() + 
  geom_polygon(
    data = hex_grid_with_counts_two_curvy2, 
    aes(x = x, y = y, 
        group = h, 
        fill = w_h), 
        colour = "grey70", 
        linewidth=0.2) +
  geom_point(data = tsne_two_curvy_scaled,
           aes(x = emb1, y = emb2),
           alpha = 0.3,
           size = 0.5) +
  scale_fill_viridis_c(direction = -1, 
    na.value = "#ffffff", option = "C") +
  xlim(sc_xlims) + ylim(sc_ylims) +
  interior_annotation("a", sc_ltr_pos, cex = 2)

trimesh_two_curvy2 <- ggplot() +
  geom_segment(data = tr_from_to_df_two_curvy2,
               aes(
                 x = x_from,
                 y = y_from,
                 xend = x_to,
                 yend = y_to),
               colour = "#000000") +
  geom_point(data = tsne_two_curvy_scaled,
             aes(
               x = tSNE1,
               y = tSNE2
             ),
             colour = "#636363",
            alpha = 0.5,
            size = 0.5
            ) +
  coord_equal() +
  interior_annotation("a1", sc_ltr_pos) 

```

```{r}
#| label: hexbin-regular-two-curvy3

hex_grid_with_counts_two_curvy3 <- read_rds(here::here("data/two_nonlinear/two_nonlinear_hex_grid_with_counts_two_curvy3.rds"))
tr_from_to_df_two_curvy3 <- read_rds(here::here("data/two_nonlinear/two_nonlinear_tr_from_to_df_two_curvy3.rds"))

hex_grid_coloured_two_curvy3 <-  ggplot() + 
  geom_polygon(
    data = hex_grid_with_counts_two_curvy3, 
    aes(x = x, y = y, 
        group = h, 
        fill = w_h), 
        colour = "grey70", linewidth=0.2) +
  geom_point(data = tsne_two_curvy_scaled,
           aes(x = emb1, y = emb2),
           alpha = 0.3,
           size = 0.5) +
  scale_fill_viridis_c(direction = -1, na.value = "#ffffff", option = "D") +
  xlim(sc_xlims) + ylim(sc_ylims) +
  interior_annotation("c", sc_ltr_pos, cex = 2)
```

```{r}
#| echo: false
#| label: fig-bins-two-curvy
#| fig-cap: "Hexbin density plots of tSNE layout of the 2NC7 data, using three different $b_1$ specifications yielding different $b_2, b, m$: (a) **15**, 18, 270, 98, (b) **24**, 29, 696, 209, and (c) **35**, 42, 1470, 386. Color indicates standardized counts, dark indicating high count, and light indicating low count. At the smallest binwidth, the data structure is discontinuous, suggesting that there are too many bins."
#| fig-width: 9
#| fig-height: 3

hex_grid_coloured_two_curvy2 +
  hex_grid_coloured_two_curvy1 +
  hex_grid_coloured_two_curvy3 +
  plot_layout(guides='collect', ncol = 3) &
  theme(legend.position='none', plot.tag = element_text(size = 12))
``` 

<!-- #### Handling of low-density bins-->

<!-- It is worthwhile to consider what desirable aspects of a hexbin result, that maps to summarizing the \pD{} fit well. The binning should capture the underlying data distribution closely, with a minimum number of necessary bins. An ideal binning might be indicated by a more uniform distribution of bin counts or having few relatively empty bins. To help with this assessment average bin count ($\bar{n} = \sum{n_h}/m$), average standardized bin count ($\bar{w} = \sum{w_h}/m$) and proportion of non-empty bins ($m/b$), are also computed. @fig-param-two-curvy shows some choices of plots of these quantities for a single NLDR layout, with three choices of $a_1$ indicated. Some expectations and reasoning for these plots are: -->

<!-- - HBE will increase as $a_1$ increases, so good choices will be just before a big increase. In plot a, HBE changes fairly steadily, so there is no easy choice to make. -->
<!-- - HBE can also be examined against the average standardized bin count or average bin count (plot b). This is similar to the comparison with $a_1$, but to use when comparing different NLDR layouts. Different layouts might produce different densities of points, which will not be captured well by a comparison of HBE vs $a_1$. -->
<!-- - The proportion of non-empty bins is interesting to examine across different binwidths (plot c). A good binning should have just the right number of bins to neatly cover the shape of the data, and no more or fewer. As binwidth gets smaller, $m/b$ should roughly get bigger. -->
<!-- - Bins with a small number of observations might be removed to sharpen the wireframe model. This can have adverse effects, though - failing to extend the wireframe into sparse areas, or resulting in holes in the wireframe. Plot d shows the relationship between HBE (computed for all observations despite some bin removal) and the standardized bin count cutoff used to remove bins. For all three chosen bin widths, a small number of bins can be removed without affecting HBE. -->

<!--two_nonlinear/05_gen_rm_lwd_mse.R-->
```{r}
#| label: rm-lwd-bin-error

error_rm_two_curvy <- read_rds(here::here("data/two_nonlinear/error_rm_lwd_diff_bin.rds")) |>
  mutate(b1  = as.factor(b1))

rmse_two_curvy_lwd <- ggplot(error_rm_two_curvy, 
                     aes(x = benchmark_rm_lwd, 
                         y = HBE,
                         colour = a1,
                         linetype = a1)) + 
  geom_point(
    size = 1
    ) +
  geom_line(
    linewidth = 0.5
    ) + 
  scale_x_continuous(breaks = sort(unique(error_rm_two_curvy$benchmark_rm_lwd))[c(1, 4, 7, 11, 15)]) +
  scale_color_manual(values=c("#d95f02", "#1f78b4", "#1b9e77"),
                     labels = parse_format()) +
  scale_linetype_manual(values = c("longdash", "solid", "dotted"),
                        labels = parse_format()) +
  xlab(expression(paste("low count bin cutoff based on ", w[h]))) +
  ylab("HBE") +
  #xlim(c(0, 0.14)) +
  ylim(c(0, 0.75)) +
  interior_annotation("d", position = c(0.07, 0.9), cex = 2) +
  theme_minimal() +
  theme(aspect.ratio = 0.75,
        panel.border = element_rect(fill = 'transparent'),
        plot.title = element_text(size = 20, 
                                  hjust = 0.5, 
                                  vjust = -0.5),
        axis.ticks.x = element_line(),
        axis.ticks.y = element_line(),
        legend.position = "none",
        legend.title = element_blank(),
        legend.text = element_text(size = 18),
        axis.title = element_text(size = 18),
        axis.text = element_text(size = 15))
```

```{r}
#| label: error-two-curvy

error_two_curvy_tsne <- read_rds(here::here("data/two_nonlinear/error_two_curvy_tsne.rds"))

rmse_two_curvy_b <- ggplot(error_two_curvy_tsne, 
                     aes(x = a1, 
                         y = HBE)) +
  geom_vline(xintercept = 0.03,
           colour = "#d95f02", linetype="dashed", 
           linewidth=1) +
  geom_vline(xintercept = 0.05,
             colour = "#1f78b4", 
             linetype="solid",
             linewidth=1) +
  geom_vline(xintercept = 0.08, linetype="dotted",
             colour = "#1b9e77", linewidth=1) +
  geom_line(linewidth = 0.5) + 
  geom_point(size = 1) +
  scale_x_continuous(breaks = seq(0.03, 0.28, 0.03)) + #sort(unique(round(error_two_curvy_tsne$a1, 2)))[c(1, 4, 7, 11, 13, 14, 15, 16)]) +
  labs(x = expression(paste("binwidth (", a[1], ")")), y = "HBE") +
  ylim(c(0, 0.75)) +
  interior_annotation("a", position = c(0.9, 0.1), cex = 2) +
  theme_minimal() +
  theme(aspect.ratio = 0.75,
        panel.border = element_rect(fill = 'transparent'),
        plot.title = element_text(size = 20, hjust = 0.5, vjust = -0.5),
        axis.ticks.x = element_line(),
        axis.ticks.y = element_line(),
        axis.title = element_text(size = 18),
        axis.text = element_text(size = 15))

error_two_curvy_tsne_filtered <- error_two_curvy_tsne |>
  filter(b1 %in% c(15, 24, 35))

rmse_two_curvy_relcount <- ggplot(error_two_curvy_tsne, 
                     aes(x = n_bar, 
                         y = HBE)) +
  geom_vline(xintercept = error_two_curvy_tsne_filtered$n_bar[2],
           colour = "#d95f02", linetype="dashed", 
           linewidth=1) +
  geom_vline(xintercept = error_two_curvy_tsne_filtered$n_bar[3],
             colour = "#1f78b4", 
             linetype="solid",
             linewidth=1) +
  geom_vline(xintercept = error_two_curvy_tsne_filtered$n_bar[1], linetype="dotted",
             colour = "#1b9e77", linewidth=1) +
  geom_line(linewidth = 0.5) + 
  geom_point(size = 1) +
#  scale_x_continuous(breaks = sort(unique(round(error_two_curvy_tsne$a1, 2)))[c(1, 4, 7, 11, 13, 14, 15, 16)]) +
  scale_x_log10() +
  ylim(c(0, 0.75)) +
  ylab("HBE") +
  xlab(expression(paste("average points per bin (", bar(n)[h], ")(log scale))"))) +
  #ylim(c(0, 0.75)) +
  annotate("text", x = 109, y = 0.05, label = "b", size = 7, colour = "grey70") +
  theme_minimal() +
  theme(aspect.ratio = 0.75,
        panel.border = element_rect(fill = 'transparent'),
        plot.title = element_text(size = 20, hjust = 0.5, vjust = -0.5),
        axis.ticks.x = element_line(),
        axis.ticks.y = element_line(),
        axis.title = element_text(size = 18),
        axis.text = element_text(size = 15)) 

a1_m_two_curvy <- ggplot(error_two_curvy_tsne,
                     aes(x = a1,
                         y = prop_bins)) +
  geom_vline(xintercept = 0.03,
           colour = "#d95f02", linetype="dashed", 
           linewidth=1) +
  geom_vline(xintercept = 0.05,
             colour = "#1f78b4", 
             linetype="solid",
             linewidth=1) +
  geom_vline(xintercept = 0.08, linetype="dotted",
             colour = "#1b9e77", linewidth=1) +
  geom_line(linewidth = 0.5) +
  geom_point(size = 1) +
  # scale_y_continuous(breaks = sort(unique(error_two_curvy$b_non_empty))[-3]) +
  scale_x_continuous(breaks = seq(0.03, 0.28, 0.03)) +
  labs(x = expression(paste("binwidth (", a[1], ")")), y = expression("prop. non-empty bins " * bgroup("(", m / b, ")"))) +
  interior_annotation("c", position = c(0.9, 0.1), cex = 2) +
  theme_minimal() +
  theme(aspect.ratio = 0.75,
        panel.border = element_rect(fill = 'transparent'),
        plot.title = element_text(size = 20, hjust = 0.5, vjust = -0.5),
        axis.ticks.x = element_line(),
        axis.ticks.y = element_line(),
        axis.title = element_text(size = 18),
axis.text = element_text(size = 15))
```


```{r}
#| echo: false
#| fig-cap: "Various plots to help tune the model fit: (a) HBE vs $a_1$, (b) HBE vs $\\bar{n}_h$, (c) proportion of non-empty bins ($m/b$) vs $a_1$, (d) HBE vs $w_h$ cutoff used for removing low count bins. Color indicates the three binwidths show in Figure 6: $0.03$ (orange dashed), $0.05$ (blue solid), and $0.08$ (green dotted). The better model fit will have low HBE, but reasonably sized bins that capture the data sufficiently. Proportion of non-empty bins tends to increase with $a_1$ (c). Removing a few low-count bins doesn't substantially change HBE, for all three binwidths (d)."
#| label: fig-param-two-curvy
#| out-width: 100%
#| fig-width: 12
#| fig-height: 10

rmse_two_curvy_b + rmse_two_curvy_relcount + a1_m_two_curvy +
  rmse_two_curvy_lwd + 
  plot_layout(ncol=2)
```

\clearpage
